### 2017-11-15 milestone1.tex
Work on the second milestone

\documentclass{article}

\usepackage[margin=1in]{geometry}
\usepackage{syntax}
\usepackage{float}
\usepackage{amsmath}
\usepackage{tikz}
\usetikzlibrary{arrows,automata}
\usepackage{minted}

\title{Deep Strictness: Milestone 1}
\author{Kenny, Hengchu}

\begin{document}
\maketitle

We aim to extend QuickCheck with the ability to probabilistically
verify the laziness/strictness behavior of Haskell functions. That is,
we will allow users to write a specification of the strictness
behavior of a particular function, and by fuzzing inputs to the
function, determine if the function is \emph{exactly} as strict as the
user has specified.

Before describing in detail how our proposed tool will work, we give
several definitions regarding laziness in Haskell:

A lazy function takes an \emph{input} to an \emph{output}. For now, we
consider only inputs and outputs restricted to single algebraic
datatypes which do not contain functions (that is, all functions we
can speak about are unary first-order functions). We plan to extend
this to multi-argument and higher-order functions in later stages of
this project.

When a lazy function is applied to an input, it is done so in a
\emph{context} which makes use of some part of its output---that is,
the context may not use the entire output data structure to continue
its own computation. We call the portion of the input which is used by
the context the \emph{demand} on the output. The semantics of lazy
evaluation dictate that any portion of the input which is not required
to compute the demanded portion of the output are not evaluated in the
first place---that is, evaluation only occurs as much as is necessary
to satisfy the demand of the calling context.

For example, let us consider a function \verb|f :: Int -> Int -> Int| where
\verb|f x n = fst (x, fib n)|. Although computing the second component of
the pair in \verb|f 1 100000000| would be very expensive, this computation
never occurs because the demand placed upon the pair (extracting the
first component) only requires the first component to be evaluated. We
say that \verb|f| is \emph{strict} in its first parameter (\verb|x|) and
\emph{lazy} in its second parameter (\verb|n|).

We can capture this notion formally as a function from a given demand
of the output to a demand of the input. We call this a \emph{demand
  specification}. Here, we can concretely represent a demand as a
\emph{subshape} of the input/output---that is, a data structure
corresponding to one of these data types, but which can be truncated
at any arbitrary position in the structure of the data type. We call
this derived data type the \emph{demand type} corresponding to a data
type.

For example, the corresponding demand type for the type \verb|(Int, Int)| would be:

\verb|Demand (Demand Int, Demand Int)|

where

\begin{minted}{haskell}
data Demand a = ~   -- ^ Not demanding
              | ! a -- ^ Demanding the constructor at this level
              | *   -- ^ Demanding a primitive or nullary constructor
\end{minted}

All the values of these types are:
\verb|~, (~, ~), ! (*, ~), ! (~, *)|, and \verb|! (*, *)|. These
represent all of the possible demands which a context could place on
the type \verb|(Int, Int)|.

Suppose we have a function \verb|g :: (Int, Int) -> Int|. The corresponding
demand specification for g would be a function
\verb|gSpec :: Demand Int -> Demand (Demand Int, Demand Int)|.

Concretely, consider the function \verb|fst| (for integers):

\begin{minted}{haskell}
fst :: (Int, Int) -> Int
fst (a, b) = a
\end{minted}

An accurate demand specification for \verb|fst| is:

\begin{minted}{haskell}
fstSpec :: Demand Int -> Demand (Demand Int, Demand Int)
fstSpec ~ = ~
fstSpec * = ! (*, ~)
\end{minted}

This states that if we demand nothing from the result of \verb|fst|, we do
not need to evaluate its argument at all. (This is true for
\emph{every} function in a lazy language!) However, if we demand the
resultant integer from \verb|fst|, we must evaluate the first integer in
the input pair, but we do not need to evaluate the second element of
the pair.

For functions like \verb|fst|, the demand on input is a \emph{static
  demand}: that is, for each demand on the output, any input to the
function will be evaluated to the same degree.

This is not true in general, and not merely in pathological
examples. Many (perhaps most!) useful functions have dynamic demand
behavior---that is, their demand specification depends on the values
of the input. We say that such functions have \emph{dependent demand}:
that is, the demand specification of these functions must consider the
input values of the function as well.

\begin{minted}{haskell}
take :: (Int, [Int]) -> [Int]
take (n, xs) =
  case (n, xs) of
    (0, _)     -> []
    (n',x:xs') -> x:(take ((n'-1), xs'))
    (n', [])   -> []
\end{minted}

Just given some demand on the output list, we can't guess the demand
on the second element of the input pair, although we do know that the
first element of the pair will always be evaluated.

Suppose we have demand the first 2 elements of some result of take,
there is no one input demand that corresponds to this output
demand. However, if we know what the first argument to \verb|take| is, we
can determine the degree to which the input list must be
evaluated. Note that the demand on the output still is another factor
in determining the demand on the input, since if we demand nothing on
the output, then nothing on the input will be demanded.

The demand specification of \verb|take| could be

\begin{minted}{haskell}
takeSpec :: (Int, [Int]) -> Demand [Demand Int] -> Demand (Demand Int, Demand [Demand Int])
takeSpec _       ~    =  ~
takeSpec (0, _)  ![]  =  !(*, ~)
takeSpec (n, xs) !ds  =
  case (ds, xs) of
    ([],     _)    -> !(*, ![])
    (d:ds', x:xs') ->
      let !(_, !ds'') = takeSpec (n-1, xs') !ds' in
      !(*, !(d:ds''))
\end{minted}

Determining statically whether a function meets its demand spec is
undecidable due to the Halting Problem. So, instead of statically
approximating these behavior, we choose to apply runtime
instrumentation with randomized testing, this gives a high level of
confidence.

We choose to implement this as an extension to QuickCheck
(CITATION). We will use generic programming to derive the demand types
of functions under consideration. We will use existing QuickCheck
generator infrastructure to fuzz inputs to the function, and demands
on the output of the function. Because only certain shapes of demand
are realizable on a given output, we specialize the demand generators
to produce only sensible demands on each invocation of the
function. We then run the function on the random input to obtain
un-evaluated output, and place a random demand on this output while
instrumenting the function to determine the resultant demand on the
input. We can then use the demand specification to check whether the
demand specification exactly matches what the actual demand behavior
of the function was in this trial.

Because we integrate with the existing functionality of QuickCheck, we
can employ its \emph{shrinking} abilities to produce minimal
counterexamples of input/demand pairs which fail to satisfy a given
demand specification, if this specification is faulty.

Note that there are two ways in which a specification may be faulty:
it may fail to produce an input demand which exactly matches the real
observed input demand, or it may fail to cover all possible realizable
combinations of input and demand.

To avoid unnecessary metaprogramming, we choose not to calculate an
instrumented version of the function under consideration. We do not
need to resort to such strategies because we can instead instrument
the input data structure to report to us how it is evaluated by the
function. As such, we may treat our functions as a black box.

In order to instrument a lazy data structure, we must use so-called
``unsafe'' features of Haskell---in particular, the function
\verb|unsafePerformIO| which allows us to attach a callback to a particular
lazy value so that the callback is run when that value is
evaluated. We may traverse the generated input data structure to add
such a callback to every node in that structure. By tracking which of
these callbacks are invoked when we demand a particular part of the
output of some function applied to the instrumented structure, we
produce an exact representation of the input demand of this function
given a particular output demand and input value.

Once we have obtained a pair of generated output demand and observed
input demand, it's trivial to run the demand specification on the
output demand and check whether it exactly matches the observed
demand. If it does not, we may repeat this process by shrinking inputs
to the function to compute a minimal example exhibiting the detected
violation of the demand specification.

We need not only to consider how to \emph{check} such specifications,
but also how to specify them in the first place! To that end, we will
use Haskell's generic programming features to provide a syntax for
users to succinctly write these demand specifications without worrying
about the details of our implementation. Ideally, this syntax would
look very similar to the syntax we use above, but we may need to make
some concessions due to the limitations of the implementation strategy
we choose for this interface.

Moving forward, we want to allow users to check demand specifications
of multi-argument functions, higher-order functions, and functions on
data structures which themselves contain functions.

\section{Progress as of Second Milestone}

As of now, we have successfully implemented a proof-of-concept for the
instrumentation of functions over lists. By making use of a mutable
pointer-based data structure, we managed to implement this instrumentation such
that the execution of the instrumented function is a constant multiplicative
factor slower than the execution of the uninstrumented function. We have also
(mostly) implemented the module which will enable us to handle instrumenting
functions of an arbitrary number of arguments. Tomorrow, we plan to meet with
Jos\'e Manuel Calder\'on Trilla tomorrow, who's done quite a bit on strictness
analysis. We are also anticipating collaborations with Leo Lampropoulos (expert
in property based random testing) and Antal Spector-Zabusky (well versed in
generic programming in Haskell).

Here are some examples of functions that operate on the \verb|List| data type.
Each of these functions demonstrate different strictness on the input list. For example, the lazy \verb|Context| doesn't inspect the input list at all, and thus is completely lazy. By contrast, \verb|spineStrict| evaluates the structure of the list without forcing its contents to be evaluated. A more complex example of a demand context is \verb|evenStrict| which, like |spineStrict|, evaluates the entire structure of the list, but which also evaluates every other element contained in the list.

\begin{minted}{haskell}
type Context a = a -> ()

lazy :: Context [a]
lazy = const ()

whnf :: Context [a]
whnf = flip seq ()

nthStrict :: Int -> Context [a]
nthStrict n = flip seq () . (!! n)

spineStrict :: Context [a]
spineStrict = flip seq () . foldl' (flip (:)) []

allStrict :: NFData a => Context [a]
allStrict = rnf

oddStrict :: NFData a => Context [a]
oddStrict []           = ()
oddStrict [x]          = rnf x
oddStrict (x : _ : xs) = rnf x `seq` oddStrict xs

evenStrict :: NFData a => Context [a]
evenStrict []           = ()
evenStrict [_]          = ()
evenStrict (_ : x : xs) = rnf x `seq` evenStrict xs
\end{minted}

To observe the behavior of these functions, we used the \verb|demandList|
function on each of the contexts defined above. We evaluate the same function
(\verb|take 6|) on the same data (\verb|[1..5]|), but with differing demands on
the result.

\begin{minted}{haskell}
mapM_ printDemand_primList $
   map (\context -> demandList context (take 6) [1..5]) $
      [lazy,
       whnf,
       spineStrict,
       nthStrict 2,
       allStrict,
       oddStrict,
       evenStrictt]
\end{minted}

We then print a representation of the demand on the \emph{input} which is
incurred by this evaluation. An ``O'' indicates that the data at
that list cell was forced in the corresponding context, and an underscore
indicates means that the data at its list cell was not evaluated. An ellipsis
indicates that the rest of the list was not evaluated, while a closing square
bracket indicates that the list was evaluated all the way down to its end.

\begin{verbatim}
lazy:                  ...
whnf:                  [_, ...
spineStrict:           [_, _, _, _, _, _]
flip seq () . (!! 2):  [_, _, O, ...
allStrict:             [O, O, O, O, O, O]
oddStrict:             [O, _, O, _, O, _]
evenStrict:            [_, O, _, O, _, O]
\end{verbatim}

The process of \emph{currying} takes a function of one argument which returns another function of one argument and converts it into a function taking two arguments and returning a result. Conversely, \emph{uncurrying} takes such a function of two arguments and converts it into a function of one argument which returns another function of one argument.

With clever type-level programming, we can generalize currying to functions with arbitrary numbers of arguments. This is necessary to enable our library to handle testing such ``curried'' functions, which are ubiquitous in Haskell. As such, we have implemented the following (non-trivial) functions, which will be a necessary part of our internal infrastructure.

\begin{minted}{haskell}
curryAll   :: function -> (Tuple (Args function) -> Result function)
uncurryAll :: (Tuple (Args function) -> Result function) -> function
\end{minted}

\end{document}

### 2017-11-15 build.sh

@@ -1,1 +1,2 @@
+#!/bin/bash
 pdflatex -shell-escape milestone1.tex

### 2017-11-15 build.sh
milestone 2 backup: includes higher-order stuff!

@@ -1,2 +1,2 @@
 #!/bin/bash
-pdflatex -shell-escape milestone1.tex
+pdflatex -shell-escape report.tex

### 2017-11-17 slides.tex
backing up slides

\documentclass{beamer}
\usetheme{metropolis}

\usepackage{pgfpages}
\usepackage{xcolor}
\usepackage{pifont}
\usepackage{minted}

%\setbeameroption{hide notes} % Only slides
%\setbeameroption{show only notes} % Only notes
\setbeameroption{show notes on second screen=right}

\title{Keep Your Laziness In Check}
\author{\textbf{Kenny Foner} and \textbf{Hengchu Zhang}}
\institute{University of Pennsylvania}
\date{November 20, 2017}

\definecolor{forestgreen}{rgb}{0.0, 0.75, 0.13}
\newcommand{\greentick}[0]{\color{forestgreen}\ding{51}}
\newcommand{\redcross}[0]{\color{red}\ding{55}}

\begin{document}
\frame{\titlepage}

\begin{frame}
\frametitle{Being lazy is fun and useful, but\dots}
\Large\dots sometimes it leads to unintended consequences.
\end{frame}

\iffalse
\begin{frame}[fragile]
\frametitle{Amortized queues have strange laziness}
\begin{minted}{haskell}
enQ a (front,     back) =
      (front, a : back)

deQ ([],         [])   = Nothing
deQ (a : front', back) = Just (a, (front', back))
deQ ([],         back) = Just (a, (front', []))
  where
    (a : front') = reverse back
\end{minted}
\end{frame}

\note[itemize]{
  \item The usual implementation of functional queues uses two lists: when an
        item is enqueued, it is cons'ed onto the back list, and when we dequeue,
        an item is taken from the head of the front list. When the front list
        becomes empty, the back queue is reversed and takes the role of the new
        front list.
  \item This has amortized O(1) performance cost.
  \item This queue is lazy as long as you don't empty the front, then it is
        fully lazy in the structure of the back list. It only forces the spine
        of the back list when the front list is emptied.
}

\begin{frame}
\frametitle{But it doesn't have to be this way!}
\large
Chris Okasaki:\\
``Simple and Efficient Purely Functional Queues and Deques''
\vspace{.5\baselineskip}\\
(\emph{Journal of Functional Programming}, October 1995)
\end{frame}
\fi

\note[itemize]{
\item Enqueue O(1), Dequeue O(1) NOT AMORTIZED because of laziness!
\item Guarantees valid persistently; thunks are only evaluated once
\item How do we know we have implemented the Okasaki Queues correctly?
\item Correctness != functional correctness or micro-benchmark performance:
\item == the right amount of lazy
}

\begin{frame}
\frametitle{Wouldn't it be nice to QuickCheck laziness?}
Traditional property-based testing (such as QuickCheck):
\begin{itemize}
\item[\greentick] Great for testing functional correctness
\item[\greentick] Write a specification, fuzz inputs to functions to automatically test against that specification
\item[\redcross] Can't observe (or even specify) properties beyond functional correctness
\end{itemize}
If we were able to \textbf{specify} and \textbf{observe} laziness, we could
treat it \emph{just like} functional correctness.
\end{frame}

\note[itemize]{
\item QuickCheck gives functional correctness
\item But laziness can't be observed by QuickCheck
\item What if we make it observable? Then it's just part of functional correctness!
}

\begin{frame}
%\frametitle{Introducing: StrickCheck}
\Huge\centering\textbf{StrictCheck}\vspace{.15\baselineskip}\\
\Large\centering``We actually can do that thing.''
\end{frame}

\begin{frame}[fragile]
\frametitle{Observing strictness (part I)}
\begin{minted}{haskell}
instrumentListWithRef :: IORef Int -> [a] -> [a]
instrumentListWithRef _     []       = []
instrumentListWithRef count (a : as) =
  unsafePerformIO $ do
    modifyIORef' count (\x -> x + 1)
    return (a : instrumentListWithRef count as)
\end{minted}
\end{frame}

\note[itemize]{
\item \texttt{instrumentListWithRef} injected a snippet of code at every cons cell
\item and it returns a new ``loaded'' list
\item the injected code is executed as the ``loaded'' list gets pattern-matched on
}

\begin{frame}[fragile]
\frametitle{Strictness doesn't exist in a vacuum}

\begin{itemize}
\setlength\itemsep{-0.5em}
\item<1->[]
\begin{minted}{haskell}
type Context a = a -> ()
\end{minted}
\item<2->[]
\begin{minted}{haskell}
-----------------------------------------------
lazy, whnf, spineStrict :: Context [a]
lazy        = \xs -> const () xs
\end{minted}
\item<3->[]
\begin{minted}{haskell}
whnf        = \xs -> case xs of
                       []    -> ()
                       (_:_) -> ()
\end{minted}
\item<4->[]
\begin{minted}{haskell}
spineStrict =
  \xs -> whnf (foldl' (\ys y -> y : ys)) [] xs)
\end{minted}
\end{itemize}
\end{frame}

\note[itemize]{
  \item The context type gives a function that returns a trivial value of the
        unit type
  \item It might look like there's only one such function \texttt{const ()}
  \item There can actually be all kinds of varying strictness behavior
  \item Forcing the unit value triggers the strictness behavior on the input \texttt{a}
  \item This gives us an easy way to observe the strictness behavior of a context
}

\begin{frame}[fragile]
\frametitle{Demanding an answer, lazily}
\begin{itemize}
\item<1->[]
\begin{minted}{haskell}
evaluate :: () -> IO ()
evaluate () = return ()
\end{minted}
\item<2->[]
\begin{minted}{haskell}
demandCount :: Context b -> ([a] -> b) -> [a] -> Int
demandCount c f as =
  unsafePerformIO $ do
    count <- newIORef 0
    let observableList =
          instrumentListWithRef count as
    evaluate ((c . f) observableList)
    readIORef count
\end{minted}
\end{itemize}
\end{frame}

\note[itemize]{
  \item The seemingly trivial evaluate function forces the unit value returned
        by a context, thus triggering evaluation of the demanded part of
        observableList
  \item \texttt{demandCount} takes a context, and a function that operates over
        lists and the input list
  \item it applies instrumentListWithRef on the input list, producing an
        observableList, and applies the function and context over the
        observableList, triggering the injected instrumentation code
}

\begin{frame}[fragile]
\frametitle{Examples of \texttt{demandCount}}
\begin{itemize}
\item<1->[]
\begin{verbatim}
ghci> let f = take 6
\end{verbatim}
\item<2->[]
\begin{verbatim}
ghci> demandCount lazy f [1..]
0
\end{verbatim}
\item<3->[]
\begin{verbatim}
ghci> demandCount whnf f [1..]
1
\end{verbatim}
\item<4->[]
\begin{verbatim}
ghci> demandCount spineStrict f [1..]
6
\end{verbatim}
\end{itemize}
\end{frame}

\note[itemize]{
\item Just some examples showing demandCount
\item Note that we simply treat \texttt{f} as a black box
\item We don't need to know anything more about \texttt{f} other than the fact
      it typechecks with demandCount!
}

\begin{frame}[fragile]
\begin{columns}
\begin{column}{0.7\textwidth}
\frametitle{Beyond lists and numbers}
\begin{figure}
\centering
\includegraphics[width=0.7\textwidth]{binary-tree}
\end{figure}
\end{column}
\begin{column}{0.2\textwidth}
{\fontsize{100}{50}\textrm{\textbf{?}}}
\end{column}
\end{columns}
\end{frame}

\note[itemize]{
\item This gives us a bare example that counts the number of cons cells forced,
      but what if we want more fine grained information of arbitrary algebraic
      data types?
\item Numbers can't directly capture the structure of a tree, so it can't
      capture the nodes of a tree that get forced in a computation
}

\begin{frame}[fragile]
\frametitle{\texttt{ListDemand}}
\begin{minted}{haskell}
data List a =
    Cons a (List a)
  | Nil

data Thunk a = T | E a

data ListDemand d =
    ConsD (Thunk d)
          (Thunk (ListDemand d))
  | NilD

data IntDemand = IntD
\end{minted}
\end{frame}

\note[itemize]{
  \item Now, we can exactly characterize the demand on a list of Ints by
        composing the type ListDemand and IntDemand
}

\begin{frame}[fragile]
\frametitle{Examples}
\begin{minted}{haskell}
data ListDemand d =
    ConsD (Thunk d)
          (Thunk (ListDemand d))
  | NilD

data IntDemand = IntD
---------------------------------------------------
ConsD T (ConsD (E IntD) (ConsD T T))

ConsD (E IntD) (ConstD T (ConsD (E IntD) (E NilD)))
\end{minted}
\end{frame}

\note[itemize]{
  \item The 2nd Int is forced, and we force 3 cons cells, we don't force
        anything in the rest of the list
  \item The 1st and 3rd Int is forced, and the list's spine is forced
  \item Note that these represent concrete observations instrumented on given
        inputs, they do not represent the general behavior of contexts
}

\begin{frame}[fragile]
\frametitle{\texttt{TreeDemand}}
\begin{minted}{haskell}
data Tree a =
    Node (Tree a) a (Tree a)
  | Leaf

data TreeDemand d =
    NodeD (Thunk (TreeDemand d))
          (Thunk d)
          (Thunk (TreeDemand d))
  | LeafD
\end{minted}
\end{frame}

\note[itemize]{
  \item The demand type for a binary Tree
  \item We can either demand the value at an internal node, or one of the
        subtrees
  \item The demand type represents all of the possible prefixes/subshapes of its
        corresponding data type
}

\begin{frame}[fragile]
\frametitle{\texttt{ListDemand} vs \texttt{TreeDemand}}
\begin{minted}{haskell}
data ListDemand d =
    ConsD (Thunk d)
          (Thunk (ListDemand d))
  | NilD

data TreeDemand d =
    NodeD (Thunk (TreeDemand d))
          (Thunk d)
          (Thunk (TreeDemand d))
  | LeafD
\end{minted}
\end{frame}

\note[itemize]{
\item Notice the similarity between ListDemand and TreeDemand with their
      corresponding data types
\item In general, the demand type of a data type simply interleaves a
      \texttt{Thunk} at every constructor field
}

\begin{frame}[fragile]
\frametitle{Computing demand, generically}
\begin{itemize}
\item<1->[]
\begin{minted}{haskell}
demandList :: Context b -> ([Int] -> b)
           -> [Int]
           -> (b, Thunk (ListDemand IntDemand))

demandTree :: Context b -> (Tree Int -> b)
           -> Tree Int
           -> (b, Thunk (TreeDemand IntDemand))
\end{minted}
\item<2->[]
\begin{minted}{haskell}
-----------------------------------------------
demand     :: Context b -> (a -> b) -> a
           -> (b, Thunk (Demand a))
\end{minted}
\end{itemize}
\end{frame}

\note[itemize]{
\item FIX THE MARGIN ISSUES
\item There is a generic transformation between a data type and its
      corresponding data type representing a demand upon it
\item \texttt{Demand} is a type level function that calculates the demand type
\item We return a \texttt{Thunk} of demand because the input might not be
      demanded at all
\item As the input value is traversed by the function, which is driven by the
      context, the evaluation of each thunk builds a pointer based data
      structure which is isomorphic to the demand data type, which is then
      frozen into a demand representation
}

\begin{frame}[fragile]
\frametitle{Generic \texttt{Demand} calculation}
\begin{minted}{haskell}
type family Demand (x :: *) :: * where
 Demand (a -> b)  = FuncDemand
 Demand (a :+: b) = Demand a :+: Demand b
 Demand (a :*: b) = Thunk (Demand a) :*: Thunk (Demand b)
\end{minted}
\end{frame}

\note[itemize]{
\item For simplicity, we speak of types in terms of sums and products
\item \texttt{FuncDemand} is isomorphic to IntDemand in that it only captures
      whether the function was used at all
}

\begin{frame}[fragile]
\frametitle{First-order specifications}
\begin{minted}{haskell}
-- uncurried take
take     :: (Int, [a]) -> [a]

takeSpec :: Int -> [a]
         -> Demand [a]
         -> (Thunk (Demand Int), Thunk (Demand [a]))
\end{minted}
\end{frame}

\note[itemize]{
  \item Having demonstrated how to reify demand behaviors into value, we now
        need to figure out how to write down a specification to check if whether
        a particular run of the program has the specified laziness according to
        the specification
  \item Specifications goings backwards from demands on the output to demands on
        the inputs of the function. This is because how much input is demanded
        depends on how much output is demanded, hence the arrows go the other
        way.
  \item Note that \texttt{takeSpec} takes the same inputs as \texttt{take}
        would. This is necessary in general because the demand behavior of
        programs can depend on their inputs, as it is the case for
        \texttt{take}!
}

\begin{frame}[fragile]
\frametitle{Connecting specification with observation}
\Huge PLACEHOLDER FOR PICTURE
\end{frame}

\note[itemize]{
  \item To QuickCheck \texttt{takeSpec}, we use mechanisms from QuickCheck to
        fuzz an Int and a [a] as inputs for \texttt{take}
  \item We use the \texttt{CoArbitrary} mechanism to generate random contexts
        that exerts non-trivial demand on the output values of \texttt{take}
  \item We use the generic \texttt{demand} function to kick off the entire
        computation
  \item We have now observed the demand on the input and also the demand on the
        output of take, so we can just straightforwardly compare that to what
        the specification expects.
  \item If we find a counterexample, we shrink the inputs first, and then shrink
        the reified demand, and re-exert that new demand on the output.
}

\begin{frame}[fragile]
\frametitle{Higher-order specifications}
\Huge PLACEHOLDER FOR PICTURE
\end{frame}

\note[itemize]{
  \item There is in fact also a mechanical transformation between the type of a
        function to the type of its demand specification
  \item Moreover, this transformation works even with high-order functions
  \item Higher-order specifications will take specification of their input
        functions because their demand behavior is parameterized over the demand
        behavior of their input functions
  \item At testing time, these specifications can be fuzzed through the same
        \texttt{CoArbitrary} mechanism
}

\begin{frame}[fragile]
\frametitle{Contributions}
With \textbf{StrictCheck}, you will be able to:
\begin{itemize}
\item \textbf{Observe} laziness from within Haskell
\item \textbf{Specify} laziness properties as Haskell functions
\item \textbf{Test} implementations against those specifications
\item \textbf{For all types},\footnotemark\, including higher-order functions and
      data types containing functions
\end{itemize}
The implementation is a work in progress.
\footnotetext[1]{Simple (i.e. non-indexed, non-existential) types}
\end{frame}

\note{
StrictCheck is a lightweight tool that adds a very small overhead for
instrumentation and observation of functional programs in Haskell. It also
provides infrastructure for writing specifications and checking
}

\end{document}

### 2017-11-17 build.sh
build.sh now takes .tex file as argument, added binary-tree.png

@@ -1,2 +1,2 @@
 #!/bin/bash
-pdflatex -shell-escape report.tex
+pdflatex -shell-escape $1

### 2017-11-19 slides.tex
backup slides
backup

@@
 \title{Keep Your Laziness In Check}
-\author{\textbf{Kenny Foner} and \textbf{Hengchu Zhang}}
+\author{\textbf{Kenny Foner}, \textbf{Hengchu Zhang}, and \textbf{Leo Lampropoulos}}
 \institute{University of Pennsylvania}
@@
 \begin{frame}
-\frametitle{Being lazy is fun and useful, but\dots}
-\Large\dots sometimes it leads to unintended consequences.
+\frametitle{Being lazy is sometimes vital, but\dots}
+\Large\dots it can lead to unintended consequences.
 \end{frame}
 
-\iffalse
+\note[itemize]{
+\item Ubiquitous in Haskell programming, used in OCaml programming
+\item (FRP) In the broader world, it is intrinsic in web-based functional
+      reactive programming
+\item (Spark) and in the field of big-data processing, for example, Spark
+}
+
 \begin{frame}[fragile]
-\frametitle{Amortized queues have strange laziness}
-\begin{minted}{haskell}
-enQ a (front,     back) =
-      (front, a : back)
-
-deQ ([],         [])   = Nothing
-deQ (a : front', back) = Just (a, (front', back))
-deQ ([],         back) = Just (a, (front', []))
-  where
-    (a : front') = reverse back
+\frametitle{Lazy lists}
+\begin{minted}{ocaml}
+type 'a lazyList =
+  | Cons of ('a Lazy.t *
+             'a lazyList Lazy.t)
+  | Nil
 \end{minted}
@@
 \note[itemize]{
-  \item The usual implementation of functional queues uses two lists: when an
-        item is enqueued, it is cons'ed onto the back list, and when we dequeue,
-        an item is taken from the head of the front list. When the front list
-        becomes empty, the back queue is reversed and takes the role of the new
-        front list.
+\item In this talk, we'll analyze functions that operate over lazy data structures
+\item for instance, a lazy list like this
+\item A lazy list is just a list with all of its parameters wrapped in a lazy
+      computation
+}
+
+\begin{frame}[fragile]
+\frametitle{Lazy queues}
+\begin{minted}{ocaml}
+let enQ a (front,                     back) =
+          (front, lazy (Cons (lazy a, back)))
+
+let deQ (front, back) =
+  match Lazy.force front with
+    | Cons (a, front') -> (a, (front', back))
+    | Nil ->
+        let Cons (a, front') = reverseLazyList back
+        in (a, (front', lazy Nil))
+\end{minted}
+\end{frame}
+
+\note[itemize]{
+  \item As a micro example to motivate why thinking about laziness can be
+        non-trivial, let's look at this queue
+  \item In this slide, we use call-by-name ML for simplicity of presentation,
+        but it can be implemented in OCaml using the Lazy module
   \item This has amortized O(1) performance cost.
@@
 
+\iffalse
 \begin{frame}
@@
 \end{frame}
-\fi
 
 \note[itemize]{
+\item Kenny:
 \item Enqueue O(1), Dequeue O(1) NOT AMORTIZED because of laziness!
@@
 }
+\fi
 
@@
 \item[\greentick] Write a specification, fuzz inputs to functions to automatically test against that specification
-\item[\redcross] Can't observe (or even specify) properties beyond functional correctness
+\item[\redcross] Can only specify and check functional correctness properties
 \end{itemize}
@@
 \item But laziness can't be observed by QuickCheck
+\item At the 3rd bullet point, say that laziness is not generally considered
+      part of functional correctness much like how asymptotic runtime
+      is not part of functional correctness because they are not directly
+      observable
 \item What if we make it observable? Then it's just part of functional correctness!
@@
 \begin{frame}[fragile]
-\frametitle{Observing strictness (part I)}
-\begin{minted}{haskell}
-instrumentListWithRef :: IORef Int -> [a] -> [a]
-instrumentListWithRef _     []       = []
-instrumentListWithRef count (a : as) =
-  unsafePerformIO $ do
-    modifyIORef' count (\x -> x + 1)
-    return (a : instrumentListWithRef count as)
+\frametitle{Observing strictness}
+TODO: Just picture?
+\begin{minted}{ocaml}
+instrumentListWithRef : int ref
+                      -> 'a lazyList
+                      -> 'a lazyList
 \end{minted}
@@
 \note[itemize]{
-\item \texttt{instrumentListWithRef} injected a snippet of code at every cons cell
-\item and it returns a new ``loaded'' list
-\item the injected code is executed as the ``loaded'' list gets pattern-matched on
+\item Hengchu:
+\item \texttt{instrumentListWithRef} clones the structure of the original list,
+      and attaches a lazy effectful computation to each constructor in the list.
+      So as the list is evaluated, each step of evaluation triggers one update
 }
@@
 \frametitle{Strictness doesn't exist in a vacuum}
-
-\begin{itemize}
-\setlength\itemsep{-0.5em}
-\item<1->[]
-\begin{minted}{haskell}
-type Context a = a -> ()
-\end{minted}
-\item<2->[]
-\begin{minted}{haskell}
------------------------------------------------
-lazy, whnf, spineStrict :: Context [a]
-lazy        = \xs -> const () xs
-\end{minted}
-\item<3->[]
-\begin{minted}{haskell}
-whnf        = \xs -> case xs of
-                       []    -> ()
-                       (_:_) -> ()
-\end{minted}
-\item<4->[]
-\begin{minted}{haskell}
-spineStrict =
-  \xs -> whnf (foldl' (\ys y -> y : ys)) [] xs)
-\end{minted}
-\end{itemize}
+TODO: Pictures with color to show how much is demanded
 \end{frame}
@@
 \note[itemize]{
-  \item The context type gives a function that returns a trivial value of the
-        unit type
-  \item It might look like there's only one such function \texttt{const ()}
-  \item There can actually be all kinds of varying strictness behavior
+  \item These pictures are illustrations of a demand context on a lazy list
+  \item These functions have the type from list to unit
+  \item It might look like there's only one such function (namely the lazy
+        implementation)
+  \item There can actually be many different demand behaviors
   \item Forcing the unit value triggers the strictness behavior on the input \texttt{a}
@@
 \frametitle{Demanding an answer, lazily}
-\begin{itemize}
-\item<1->[]
-\begin{minted}{haskell}
-evaluate :: () -> IO ()
-evaluate () = return ()
+\begin{minted}{ocaml}
+demandCount context f xs =
+    let count          = ref 0
+        observableList =
+          instrumentListWithRef count xs
+    in context (f observableList); !count
 \end{minted}
-\item<2->[]
-\begin{minted}{haskell}
-demandCount :: Context b -> ([a] -> b) -> [a] -> Int
-demandCount c f as =
-  unsafePerformIO $ do
-    count <- newIORef 0
-    let observableList =
-          instrumentListWithRef count as
-    evaluate ((c . f) observableList)
-    readIORef count
-\end{minted}
-\end{itemize}
 \end{frame}
@@
 \note[itemize]{
-  \item The seemingly trivial evaluate function forces the unit value returned
-        by a context, thus triggering evaluation of the demanded part of
-        observableList
   \item \texttt{demandCount} takes a context, and a function that operates over
@@
         observableList, triggering the injected instrumentation code
+  \item context is exerting some demand on the output of f
 }
@@
 \begin{verbatim}
-ghci> let f = take 6
+> let f xs = takeLazy (lazy 6, xs)
+> let list = toLazyList [1; 2; 3; 4; 5]
 \end{verbatim}
@@
 \begin{verbatim}
-ghci> demandCount lazy f [1..]
-0
+> demandCount noDemand f list
 \end{verbatim}
@@
 \begin{verbatim}
-ghci> demandCount whnf f [1..]
-1
+0
 \end{verbatim}
@@
 \begin{verbatim}
-ghci> demandCount spineStrict f [1..]
-6
+> demandCount firstCons f list
+\end{verbatim}
+\item<5->[]
+\begin{verbatim}
+1
+\end{verbatim}
+\item<6->[]
+\begin{verbatim}
+> demandCount spineStrict f list
+\end{verbatim}
+\item<7->[]
+\begin{verbatim}
+5
 \end{verbatim}
@@
 \item Just some examples showing demandCount
+\item takeLazy is a function that takes the specified number of elements from a
+      lazy list and returns a lazyList
 \item Note that we simply treat \texttt{f} as a black box
@@
 \begin{frame}[fragile]
+TODO: erase the numbers in tree
 \begin{columns}
@@
 \note[itemize]{
+\item Kenny:
+\item Arbitrary lazy algebraic datatypes
 \item This gives us a bare example that counts the number of cons cells forced,
@@
 \frametitle{\texttt{ListDemand}}
-\begin{minted}{haskell}
-data List a =
-    Cons a (List a)
+\begin{minted}{ocaml}
+type 'a lazyList =
+  | Cons of ('a Lazy.t *
+             'a lazyList Lazy.t)
   | Nil
+------------------------------------
+type 'a thunk = T | E of 'a
 
-data Thunk a = T | E a
-
-data ListDemand d =
-    ConsD (Thunk d)
-          (Thunk (ListDemand d))
+type 'd listDemand =
+  | ConsD of ('d thunk *
+              'd listDemand thunk)
   | NilD
 
-data IntDemand = IntD
+type intDemand = IntD
 \end{minted}
@@
 \frametitle{Examples}
-\begin{minted}{haskell}
-data ListDemand d =
-    ConsD (Thunk d)
-          (Thunk (ListDemand d))
+\begin{minted}{ocaml}
+type 'd listDemand =
+  | ConsD of ('d thunk *
+              'd listDemand thunk)
   | NilD
 
-data IntDemand = IntD
+type intDemand = IntD
 ---------------------------------------------------
-ConsD T (ConsD (E IntD) (ConsD T T))
+ConsD (T,
+  ConsD (E IntD,
+    ConsD (T, T)))
 
-ConsD (E IntD) (ConstD T (ConsD (E IntD) (E NilD)))
+ConsD (E IntD,
+  ConsD (T,
+    ConsD (E IntD, E NilD)))
 \end{minted}
@@
 \frametitle{\texttt{TreeDemand}}
-\begin{minted}{haskell}
-data Tree a =
-    Node (Tree a) a (Tree a)
+\begin{minted}{ocaml}
+type 'a lazyTree =
+  | Node of ('a lazyTree Lazy.t *
+             'a Lazy.t *
+             'a lazyTree Lazy.t)
   | Leaf
 
-data TreeDemand d =
-    NodeD (Thunk (TreeDemand d))
-          (Thunk d)
-          (Thunk (TreeDemand d))
+type 'd treeDemand =
+  | NodeD of ('d treeDemand thunk *
+              'd thunk *
+              'd treeDemand thunk)
   | LeafD
@@
 \frametitle{\texttt{ListDemand} vs \texttt{TreeDemand}}
-\begin{minted}{haskell}
-data ListDemand d =
-    ConsD (Thunk d)
-          (Thunk (ListDemand d))
+\begin{minted}{ocaml}
+type 'd listDemand =
+  | ConsD of ('d thunk *
+              'd listDemand thunk)
   | NilD
 
-data TreeDemand d =
-    NodeD (Thunk (TreeDemand d))
-          (Thunk d)
-          (Thunk (TreeDemand d))
+type 'd treeDemand =
+  | NodeD of ('d treeDemand thunk *
+              'd thunk *
+              'd treeDemand thunk)
   | LeafD
@@
 \item<1->[]
-\begin{minted}{haskell}
-demandList :: Context b -> ([Int] -> b)
-           -> [Int]
-           -> (b, Thunk (ListDemand IntDemand))
+\begin{minted}{ocaml}
+demandList : 'b context -> (int lazyList -> 'b)
+           -> int lazyList
+           -> ('b * intDemand listDemand thunk)
 
-demandTree :: Context b -> (Tree Int -> b)
-           -> Tree Int
-           -> (b, Thunk (TreeDemand IntDemand))
+demandTree : 'b context -> (int lazyTree -> 'b)
+           -> int lazyTree
+           -> ('b * intDemand treeDemand thunk)
 \end{minted}
 \item<2->[]
-\begin{minted}{haskell}
------------------------------------------------
-demand     :: Context b -> (a -> b) -> a
-           -> (b, Thunk (Demand a))
+\begin{minted}{ocaml}
+------------------------------------------------
+demand     : 'b context -> ('a -> 'b) -> 'a
+           -> ('b * ('a DEMAND) thunk)
 \end{minted}
@@
 \note[itemize]{
-\item FIX THE MARGIN ISSUES
-\item There is a generic transformation between a data type and its
-      corresponding data type representing a demand upon it
-\item \texttt{Demand} is a type level function that calculates the demand type
+\item There is a determinate relation between a lazy data structure and its
+      demand representation
 \item We return a \texttt{Thunk} of demand because the input might not be
@@
       frozen into a demand representation
+\item We expect that the 'a parameter is some kind of lazy structure, and 'b is
+      also some kind of lazy structure
+\item We use generic programming to implement the DEMAND type for all algebraic
+      data types
 }
 
+\iffalse
 \begin{frame}[fragile]
@@
 \frametitle{First-order specifications}
-\begin{minted}{haskell}
--- uncurried take
-take     :: (Int, [a]) -> [a]
-
-takeSpec :: Int -> [a]
-         -> Demand [a]
-         -> (Thunk (Demand Int), Thunk (Demand [a]))
+\begin{minted}{ocaml}
+deQSpec :  (int lazyList Lazy.t * int lazyList Lazy.t)
+        -> (intDemand thunk *
+             (intDemand listDemand thunk *
+              intDemand listDemand thunk))
+        -> (intDemand listDemand thunk *
+            intDemand listDemand thunk)
+deQSpec (front, back) (dInt, (dFront, dBack)) =
+  match Lazy.force front with
+    | Cons (_, _) -> (E (ConsD (dInt, dFront)), dBack)
+    | Nil -> (dBack,
+              (spineStrictD |back|)
+               U ((spineStrictD
+                   (|back| - |dFront|))
+                  @ (reverse (ConsD dInt dFront)))
 \end{minted}
@@
   \item To QuickCheck \texttt{takeSpec}, we use mechanisms from QuickCheck to
-        fuzz an Int and a [a] as inputs for \texttt{take}
-  \item We use the \texttt{CoArbitrary} mechanism to generate random contexts
-        that exerts non-trivial demand on the output values of \texttt{take}
+        fuzz an Int and a [a] as inputs
+  \item We can generate random contexts
+        that exerts non-trivial demand on the output values
   \item We use the generic \texttt{demand} function to kick off the entire
@@
         the reified demand, and re-exert that new demand on the output.
+  \item Higher-order also works! but we don't want to go into the details
 }
 
+\iffalse
 \begin{frame}[fragile]
@@
 
+\begin{frame}[fragile]
+\frametitle{Have questions?}
+Come ask us about \dots
+\begin{itemize}
+\item Higher-order specifications
+\item Generic programming
+\item Random generation of demand contexts
+\item Details of instrumentation
+\item What language are you actually implementing this in? \\ (Hint: it's not ML)
+\item[] \dots
+\end{itemize}
+\end{frame}
+
 \end{document}

### 2017-11-20 slides.tex
added images for illustrations
Remove todo about erasing numbers
Add new items to potential questions

@@
 \usepackage{pifont}
+\usepackage{float}
 \usepackage{minted}
@@
 \note[itemize]{
+\item PBRT is a technique for randomized testing which allows users to write
+  down an executable specification for a function and test whether it holds for
+  randomly generated inputs.
 \item QuickCheck gives functional correctness
 \item But laziness can't be observed by QuickCheck
-\item At the 3rd bullet point, say that laziness is not generally considered
+\item Say that laziness is not generally considered
       part of functional correctness much like how asymptotic runtime
@@
 \frametitle{Observing strictness}
-TODO: Just picture?
+\begin{figure}
+\centering
+\includegraphics[width=0.6\textwidth]{instrumentListWithRef}
+\caption{Instrumented List}
+\end{figure}
 \begin{minted}{ocaml}
@@
 \frametitle{Strictness doesn't exist in a vacuum}
-TODO: Pictures with color to show how much is demanded
+\begin{figure}
+\centering
+\includegraphics[width=0.7\textwidth]{noDemand}
+\caption{No Demand}
+\end{figure}
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Strictness doesn't exist in a vacuum}
+\begin{figure}
+\centering
+\includegraphics[width=0.7\textwidth]{whnf}
+\caption{Weak Head Normal Form}
+\end{figure}
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Strictness doesn't exist in a vacuum}
+\begin{figure}
+\centering
+\includegraphics[width=0.7\textwidth]{spinestrict}
+\caption{Spine Strict}
+\end{figure}
 \end{frame}
@@
   \item These pictures are illustrations of a demand context on a lazy list
-  \item These functions have the type from list to unit
-  \item It might look like there's only one such function (namely the lazy
-        implementation)
-  \item There can actually be many different demand behaviors
-  \item Forcing the unit value triggers the strictness behavior on the input \texttt{a}
-  \item This gives us an easy way to observe the strictness behavior of a context
-}
+  \item This gives us an easy way to observe the strictness behavior of a
+    context
+  \item While these pictures might look like a data structure, they are meant to
+    describe the behavior of a context of evaluation on a lazy data structure of
+    this shape }
 
@@
 \begin{frame}[fragile]
-TODO: erase the numbers in tree
 \begin{columns}
@@
 \note[itemize]{
-  \item Now, we can exactly characterize the demand on a list of Ints by
-        composing the type ListDemand and IntDemand
-}
+\item Now, we can exactly characterize a demand on a list of Ints by composing
+  the type ListDemand and IntDemand
+\item The type ListDemand is a reification of the demand placed upon a lazy list
+  during the course of evaluation in some context }
 
@@
 \begin{frame}[fragile]
-\frametitle{Computing demand, generically}
+\frametitle{Observing demand, generically}
 \begin{itemize}
@@
       demanded at all
-\item As the input value is traversed by the function, which is driven by the
-      context, the evaluation of each thunk builds a pointer based data
-      structure which is isomorphic to the demand data type, which is then
-      frozen into a demand representation
 \item We expect that the 'a parameter is some kind of lazy structure, and 'b is
@@
 \frametitle{Connecting specification with observation}
-\Huge PLACEHOLDER FOR PICTURE
+\begin{figure}
+\centering
+\includegraphics[width=1.0\textwidth]{specflow}
+\end{figure}
 \end{frame}
@@
 \note[itemize]{
-  \item To QuickCheck \texttt{takeSpec}, we use mechanisms from QuickCheck to
-        fuzz an Int and a [a] as inputs
+\item We randomly fuzz the inputs to the function, and
   \item We can generate random contexts
-        that exerts non-trivial demand on the output values
+        that exert non-trivial demand on the output values
   \item We use the generic \texttt{demand} function to kick off the entire
@@
   \item We have now observed the demand on the input and also the demand on the
-        output of take, so we can just straightforwardly compare that to what
+        output, so we can just straightforwardly compare that to what
         the specification expects.
@@
 \frametitle{Higher-order specifications}
-\Huge PLACEHOLDER FOR PICTURE
 \end{frame}
@@
 \item Higher-order specifications
-\item Generic programming
-\item Random generation of demand contexts
+\item Generic programming for any arity and datatype
+\item Random generation of partial demand contexts
+\item Efficiency of our implementation
 \item Details of instrumentation

### 2017-11-20 slides.tex
More edits to slide content

@@
         non-trivial, let's look at this queue
-  \item In this slide, we use call-by-name ML for simplicity of presentation,
-        but it can be implemented in OCaml using the Lazy module
   \item This has amortized O(1) performance cost.
@@
 \item QuickCheck gives functional correctness
-\item But laziness can't be observed by QuickCheck
-\item Say that laziness is not generally considered
-      part of functional correctness much like how asymptotic runtime
-      is not part of functional correctness because they are not directly
-      observable
+\item But laziness can't currently be observed by tools like QuickCheck
+\item Laziness is not generally considered part of functional correctness much
+  like how asymptotic runtime is not part of functional correctness because they
+  are not directly observable
 \item What if we make it observable? Then it's just part of functional correctness!
@@
 \begin{frame}[fragile]
-\frametitle{\texttt{ListDemand}}
+  \frametitle{Reifying demand on lazy lists}
 \begin{minted}{ocaml}
@@
 \begin{frame}[fragile]
-\frametitle{\texttt{TreeDemand}}
+  \frametitle{Reifying demand on lazy trees}
 \begin{minted}{ocaml}
@@
 \begin{frame}[fragile]
-\frametitle{\texttt{ListDemand} vs \texttt{TreeDemand}}
+  \frametitle{Comparing demands on lazy lists and trees}
 \begin{minted}{ocaml}
@@
 \item \texttt{FuncDemand} is isomorphic to IntDemand in that it only captures
-      whether the function was used at all
-}
-\fi
+  whether the function was used at all } \fi
+
+\begin{frame}[fragile]
+  \frametitle{Recalling \texttt{deQ}}
+  \begin{minted}{ocaml}
+type 'a lazyList =
+  | Cons of ('a Lazy.t *
+             'a lazyList Lazy.t)
+  | Nil
+
+let deQ (front, back) =
+  match Lazy.force front with
+    | Cons (a, front') -> (a, (front', back))
+    | Nil ->
+        let Cons (a, front') = reverseLazyList back
+        in (a, (front', lazy Nil))
+  \end{minted}
+\end{frame}
 
@@
 \item What language are you actually implementing this in? \\ (Hint: it's not ML)
-\item[] \dots
+\item[] \ \ $\vdots$
 \end{itemize}

### 2017-11-20 slides.tex
almost ready

@@
 \item In this talk, we'll analyze functions that operate over lazy data structures
+\item In OCaml, Lazy is a standard library module that provides operators for
+      suspending computations into lazy values
 \item for instance, a lazy list like this
@@
 
-let deQ (front, back) =
+let deQ (front, back)
+  : ('a Lazy.t * ('a lazyList Lazy.t *
+                  'a lazyList Lazy.t)) =
   match Lazy.force front with
@@
 \begin{minted}{ocaml}
-deQSpec : (int lazyList Lazy.t * int lazyList Lazy.t)
-        -> (intDemand thunk *
-             (intDemand listDemand thunk *
-              intDemand listDemand thunk))
-        -> (intDemand listDemand thunk *
-            intDemand listDemand thunk)
-deQSpec (front, back) (dInt, (dFront, dBack)) =
+let spec_deQ (front, back) (dInt, (dFront, dBack)) =
   match Lazy.force front with
-    | Cons (_, _) -> (E (ConsD (dInt, dFront)), dBack)
-    | Nil -> (dBack,
-              (spineStrictD |back|)
-               U ((spineStrictD
-                   (|back| - |dFront|))
-                  @ (reverse (ConsD dInt dFront)))
+    | Cons (_, _) ->
+        (E (ConsD (dInt, dFront)), dBack)
+    | Nil -> (E NilD,
+              spineStrictAs back
+                (pad (length back - lengthD dFront)
+                     (reverseD (ConsD dInt dFront))))
 \end{minted}
@@
         the specification
-  \item Specifications goings backwards from demands on the output to demands on
+  \item Specifications go backwards from demands on the output to demands on
         the inputs of the function. This is because how much input is demanded
@@
         way.
-  \item Note that \texttt{takeSpec} takes the same inputs as \texttt{take}
-        would. This is necessary in general because the demand behavior of
-        programs can depend on their inputs, as it is the case for
-        \texttt{take}!
+  \item Specifications are passed in the actual values of the inputs because
+        their demand behaviors can be dependent on those inputs
+  \item spineStrictAs takes a lazy data structure, and unions a spine strict
+        demand corresponding to the whole structure with another demand on some
+        sub-part of that structure
 }
@@
         that exert non-trivial demand on the output values
-  \item We use the generic \texttt{demand} function to kick off the entire
-        computation
-  \item We have now observed the demand on the input and also the demand on the
+  \item We observe the demand on the input and also the demand on the
         output, so we can just straightforwardly compare that to what
         the specification expects.
-  \item If we find a counterexample, we shrink the inputs first, and then shrink
-        the reified demand, and re-exert that new demand on the output.
-  \item Higher-order also works! but we don't want to go into the details
 }
@@
 
-\note{
-StrictCheck is a lightweight tool that adds a very small overhead for
-instrumentation and observation of functional programs in Haskell. It also
-provides infrastructure for writing specifications and checking
-}
-
 \begin{frame}[fragile]
 \frametitle{Have questions?}
-Come ask us about \dots
+Ask us about \dots
 \begin{itemize}
@@
 \item Random generation of partial demand contexts
+\item Shrinking of contexts and inputs
 \item Efficiency of our implementation

### 2017-12-03 presentations/report.tex
Reorganize to stash presentation and paper materials in subdirs

--- 2017-11-15 milestone1.tex
@@
 
-\title{Deep Strictness: Milestone 1}
+\title{Deep Strictness: Milestone 2}
 \author{Kenny, Hengchu}
@@
 
-We aim to extend QuickCheck with the ability to probabilistically
-verify the laziness/strictness behavior of Haskell functions. That is,
-we will allow users to write a specification of the strictness
-behavior of a particular function, and by fuzzing inputs to the
-function, determine if the function is \emph{exactly} as strict as the
-user has specified.
+We aim to extend QuickCheck with the ability to probabilistically verify the
+laziness/strictness behavior of Haskell functions. That is, we will allow users
+to write a specification of the strictness behavior of a particular function,
+and by fuzzing inputs to the function, determine if the function is
+\emph{exactly} as strict as the user has specified.
 
-Before describing in detail how our proposed tool will work, we give
-several definitions regarding laziness in Haskell:
+Before describing in detail how our proposed tool will work, we give several
+definitions regarding laziness in Haskell:
 
-A lazy function takes an \emph{input} to an \emph{output}. For now, we
-consider only inputs and outputs restricted to single algebraic
-datatypes which do not contain functions (that is, all functions we
-can speak about are unary first-order functions). We plan to extend
-this to multi-argument and higher-order functions in later stages of
-this project.
+A lazy function takes an \emph{input} to an \emph{output}. For now, we consider
+only inputs and outputs restricted to single algebraic datatypes which do not
+contain functions (that is, all functions we can speak about are unary
+first-order functions). We plan to extend this to multi-argument and
+higher-order functions in later stages of this project.
 
-When a lazy function is applied to an input, it is done so in a
-\emph{context} which makes use of some part of its output---that is,
-the context may not use the entire output data structure to continue
-its own computation. We call the portion of the input which is used by
-the context the \emph{demand} on the output. The semantics of lazy
-evaluation dictate that any portion of the input which is not required
-to compute the demanded portion of the output are not evaluated in the
-first place---that is, evaluation only occurs as much as is necessary
-to satisfy the demand of the calling context.
+When a lazy function is applied to an input, it is done so in a \emph{context}
+which makes use of some part of its output---that is, the context may not use
+the entire output data structure to continue its own computation. We call the
+portion of the input which is used by the context the \emph{demand} on the
+output. The semantics of lazy evaluation dictate that any portion of the input
+which is not required to compute the demanded portion of the output are not
+evaluated in the first place---that is, evaluation only occurs as much as is
+necessary to satisfy the demand of the calling context.
 
 For example, let us consider a function \verb|f :: Int -> Int -> Int| where
-\verb|f x n = fst (x, fib n)|. Although computing the second component of
-the pair in \verb|f 1 100000000| would be very expensive, this computation
-never occurs because the demand placed upon the pair (extracting the
-first component) only requires the first component to be evaluated. We
-say that \verb|f| is \emph{strict} in its first parameter (\verb|x|) and
-\emph{lazy} in its second parameter (\verb|n|).
+\verb|f x n = fst (x, fib n)|. Although computing the second component of the
+pair in \verb|f 1 100000000| would be very expensive, this computation never
+occurs because the demand placed upon the pair (extracting the first component)
+only requires the first component to be evaluated. We say that \verb|f| is
+\emph{strict} in its first parameter (\verb|x|) and \emph{lazy} in its second
+parameter (\verb|n|).
 
-We can capture this notion formally as a function from a given demand
-of the output to a demand of the input. We call this a \emph{demand
+We can capture this notion formally as a function from a given demand of the
+output to a demand of the input. We call this a \emph{demand
   specification}. Here, we can concretely represent a demand as a
-\emph{subshape} of the input/output---that is, a data structure
-corresponding to one of these data types, but which can be truncated
-at any arbitrary position in the structure of the data type. We call
-this derived data type the \emph{demand type} corresponding to a data
-type.
+\emph{subshape} of the input/output---that is, a data structure corresponding to
+one of these data types, but which can be truncated at any arbitrary position in
+the structure of the data type. We call this derived data type the \emph{demand
+  type} corresponding to a data type.
 
@@
 
-This states that if we demand nothing from the result of \verb|fst|, we do
-not need to evaluate its argument at all. (This is true for
-\emph{every} function in a lazy language!) However, if we demand the
-resultant integer from \verb|fst|, we must evaluate the first integer in
-the input pair, but we do not need to evaluate the second element of
-the pair.
+This states that if we demand nothing from the result of \verb|fst|, we do not
+need to evaluate its argument at all. (This is true for \emph{every} function in
+a lazy language!) However, if we demand the resultant integer from \verb|fst|,
+we must evaluate the first integer in the input pair, but we do not need to
+evaluate the second element of the pair.
 
-For functions like \verb|fst|, the demand on input is a \emph{static
-  demand}: that is, for each demand on the output, any input to the
-function will be evaluated to the same degree.
+For functions like \verb|fst|, the demand on input is a \emph{static demand}:
+that is, for each demand on the output, any input to the function will be
+evaluated to the same degree.
 
-This is not true in general, and not merely in pathological
-examples. Many (perhaps most!) useful functions have dynamic demand
-behavior---that is, their demand specification depends on the values
-of the input. We say that such functions have \emph{dependent demand}:
-that is, the demand specification of these functions must consider the
-input values of the function as well.
+This is not true in general, and not merely in pathological examples. Many
+(perhaps most!) useful functions have dynamic demand behavior---that is, their
+demand specification depends on the values of the input. We say that such
+functions have \emph{dependent demand}: that is, the demand specification of
+these functions must consider the input values of the function as well.
 
@@
 
-Just given some demand on the output list, we can't guess the demand
-on the second element of the input pair, although we do know that the
-first element of the pair will always be evaluated.
+Just given some demand on the output list, we can't guess the demand on the
+second element of the input pair, although we do know that the first element of
+the pair will always be evaluated.
 
-Suppose we have demand the first 2 elements of some result of take,
-there is no one input demand that corresponds to this output
-demand. However, if we know what the first argument to \verb|take| is, we
-can determine the degree to which the input list must be
-evaluated. Note that the demand on the output still is another factor
-in determining the demand on the input, since if we demand nothing on
-the output, then nothing on the input will be demanded.
+Suppose we have demand the first 2 elements of some result of take, there is no
+one input demand that corresponds to this output demand. However, if we know
+what the first argument to \verb|take| is, we can determine the degree to which
+the input list must be evaluated. Note that the demand on the output still is
+another factor in determining the demand on the input, since if we demand
+nothing on the output, then nothing on the input will be demanded.
 
@@
 
-Determining statically whether a function meets its demand spec is
-undecidable due to the Halting Problem. So, instead of statically
-approximating these behavior, we choose to apply runtime
-instrumentation with randomized testing, this gives a high level of
-confidence.
+Determining statically whether a function meets its demand spec is undecidable
+due to the Halting Problem. So, instead of statically approximating these
+behavior, we choose to apply runtime instrumentation with randomized testing,
+this gives a high level of confidence.
 
-We choose to implement this as an extension to QuickCheck
-(CITATION). We will use generic programming to derive the demand types
-of functions under consideration. We will use existing QuickCheck
-generator infrastructure to fuzz inputs to the function, and demands
-on the output of the function. Because only certain shapes of demand
-are realizable on a given output, we specialize the demand generators
-to produce only sensible demands on each invocation of the
-function. We then run the function on the random input to obtain
-un-evaluated output, and place a random demand on this output while
-instrumenting the function to determine the resultant demand on the
-input. We can then use the demand specification to check whether the
-demand specification exactly matches what the actual demand behavior
-of the function was in this trial.
+We choose to implement this as an extension to QuickCheck (CITATION). We will
+use generic programming to derive the demand types of functions under
+consideration. We will use existing QuickCheck generator infrastructure to fuzz
+inputs to the function, and demands on the output of the function. Because only
+certain shapes of demand are realizable on a given output, we specialize the
+demand generators to produce only sensible demands on each invocation of the
+function. We then run the function on the random input to obtain un-evaluated
+output, and place a random demand on this output while instrumenting the
+function to determine the resultant demand on the input. We can then use the
+demand specification to check whether the demand specification exactly matches
+what the actual demand behavior of the function was in this trial.
 
-Because we integrate with the existing functionality of QuickCheck, we
-can employ its \emph{shrinking} abilities to produce minimal
-counterexamples of input/demand pairs which fail to satisfy a given
-demand specification, if this specification is faulty.
+Because we integrate with the existing functionality of QuickCheck, we can
+employ its \emph{shrinking} abilities to produce minimal counterexamples of
+input/demand pairs which fail to satisfy a given demand specification, if this
+specification is faulty.
 
-Note that there are two ways in which a specification may be faulty:
-it may fail to produce an input demand which exactly matches the real
-observed input demand, or it may fail to cover all possible realizable
-combinations of input and demand.
+Note that there are two ways in which a specification may be faulty: it may fail
+to produce an input demand which exactly matches the real observed input demand,
+or it may fail to cover all possible realizable combinations of input and
+demand.
 
-To avoid unnecessary metaprogramming, we choose not to calculate an
-instrumented version of the function under consideration. We do not
-need to resort to such strategies because we can instead instrument
-the input data structure to report to us how it is evaluated by the
-function. As such, we may treat our functions as a black box.
+To avoid unnecessary metaprogramming, we choose not to calculate an instrumented
+version of the function under consideration. We do not need to resort to such
+strategies because we can instead instrument the input data structure to report
+to us how it is evaluated by the function. As such, we may treat our functions
+as a black box.
 
-In order to instrument a lazy data structure, we must use so-called
-``unsafe'' features of Haskell---in particular, the function
-\verb|unsafePerformIO| which allows us to attach a callback to a particular
-lazy value so that the callback is run when that value is
-evaluated. We may traverse the generated input data structure to add
-such a callback to every node in that structure. By tracking which of
-these callbacks are invoked when we demand a particular part of the
-output of some function applied to the instrumented structure, we
-produce an exact representation of the input demand of this function
-given a particular output demand and input value.
+In order to instrument a lazy data structure, we must use so-called ``unsafe''
+features of Haskell---in particular, the function \verb|unsafePerformIO| which
+allows us to attach a callback to a particular lazy value so that the callback
+is run when that value is evaluated. We may traverse the generated input data
+structure to add such a callback to every node in that structure. By tracking
+which of these callbacks are invoked when we demand a particular part of the
+output of some function applied to the instrumented structure, we produce an
+exact representation of the input demand of this function given a particular
+output demand and input value.
 
-Once we have obtained a pair of generated output demand and observed
-input demand, it's trivial to run the demand specification on the
-output demand and check whether it exactly matches the observed
-demand. If it does not, we may repeat this process by shrinking inputs
-to the function to compute a minimal example exhibiting the detected
-violation of the demand specification.
+Once we have obtained a pair of generated output demand and observed input
+demand, it's trivial to run the demand specification on the output demand and
+check whether it exactly matches the observed demand. If it does not, we may
+repeat this process by shrinking inputs to the function to compute a minimal
+example exhibiting the detected violation of the demand specification.
 
-We need not only to consider how to \emph{check} such specifications,
-but also how to specify them in the first place! To that end, we will
-use Haskell's generic programming features to provide a syntax for
-users to succinctly write these demand specifications without worrying
-about the details of our implementation. Ideally, this syntax would
-look very similar to the syntax we use above, but we may need to make
-some concessions due to the limitations of the implementation strategy
-we choose for this interface.
+We need not only to consider how to \emph{check} such specifications, but also
+how to specify them in the first place! To that end, we will use Haskell's
+generic programming features to provide a syntax for users to succinctly write
+these demand specifications without worrying about the details of our
+implementation. Ideally, this syntax would look very similar to the syntax we
+use above, but we may need to make some concessions due to the limitations of
+the implementation strategy we choose for this interface.
 
-Moving forward, we want to allow users to check demand specifications
-of multi-argument functions, higher-order functions, and functions on
-data structures which themselves contain functions.
+Moving forward, we want to allow users to check demand specifications of
+multi-argument functions, higher-order functions, and functions on data
+structures which themselves contain functions.
 
@@
 Here are some examples of functions that operate on the \verb|List| data type.
-Each of these functions demonstrate different strictness on the input list.
-For example, the lazy \verb|Context| doesn't inspect the input list at all, and thus is completely lazy.
-By contrast, \verb|spineStrict| evaluates the structure of the list without forcing its contents to be evaluated.
-A more complex example of a demand context is \verb|evenStrict| which, like |spineStrict|, evaluates the entire structure of the list, but which also evaluates every other element contained in the list.
+Each of these functions demonstrate different strictness on the input list. For
+example, the lazy \verb|Context| doesn't inspect the input list at all, and thus
+is completely lazy. By contrast, \verb|spineStrict| evaluates the structure of
+the list without forcing its contents to be evaluated. A more complex example of
+a demand context is \verb|evenStrict| which, like \verb|spineStrict|, evaluates
+the entire structure of the list, but which also evaluates every other element
+contained in the list.
 
@@
 We then print a representation of the demand on the \emph{input} which is
-incurred by this evaluation. An ``O'' indicates that the data at
-that list cell was forced in the corresponding context, and an underscore
-indicates means that the data at its list cell was not evaluated. An ellipsis
-indicates that the rest of the list was not evaluated, while a closing square
-bracket indicates that the list was evaluated all the way down to its end.
+incurred by this evaluation. An ``O'' indicates that the data at that list cell
+was forced in the corresponding context, and an underscore indicates means that
+the data at its list cell was not evaluated. An ellipsis indicates that the rest
+of the list was not evaluated, while a closing square bracket indicates that the
+list was evaluated all the way down to its end.
 
@@
 spineStrict:  [_, _, _, _, _, _]
-flip seq () . (!! 2):  [_, _, O, ...
+nthStrict 2:  [_, _, O, ...
 allStrict:    [O, O, O, O, O, O]
@@
 
-The process of \emph{currying} takes a function of one argument which returns another function of one argument and converts it into a function taking two arguments and returning a result.
-Conversely, \emph{uncurrying} takes such a function of two arguments and converts it into a function of one argument which returns another function of one argument.
+The process of \emph{currying} takes a function of one argument which returns
+another function of one argument and converts it into a function taking two
+arguments and returning a result. Conversely, \emph{uncurrying} takes such a
+function of two arguments and converts it into a function of one argument which
+returns another function of one argument.
 
-With clever type-level programming, we can generalize currying to functions with arbitrary numbers of arguments.
-This is necessary to enable our library to handle testing such ``curried'' functions, which are ubiquitous in Haskell.
-As such, we have implemented the following (non-trivial) functions, which will be a necessary part of our internal infrastructure.
+With clever type-level programming, we can generalize currying to functions with
+arbitrary numbers of arguments. This is necessary to enable our library to
+handle testing such ``curried'' functions, which are ubiquitous in Haskell. As
+such, we have implemented the following (non-trivial) functions, which will be a
+necessary part of our internal infrastructure.
 
@@
 
+There is a mechanical transformation between the type of a function and the
+corresponding type of its demand specification. Similarly, there is a mechanical
+transformation from the type of a value to the type of the representation of the
+demand upon it.
+
+In order to perform these transformations, we use a feature in Haskell called
+Type Families. Type Families allow us to write recursive functions over the
+structure of types to compute other types. The Haskell compiler GHC has another
+feature called Generics, which transforms any type to a generic representation
+that uses sums, products and arrows. We have named our two Type Families
+\verb|Spec| and \verb|Demand| respectively, and here is their definition.
+
+\begin{minted}{haskell}
+data Thunk a = T | E a
+
+Demand (a :+: b) = Demand a :+: Demand b
+Demand (a :*: b) = Thunk (Demand a) :*: Thunk (Demand b)
+Demand (a -> b)  = FuncDemand
+
+Spec (a :+: b)   = Spec a :+: Spec b
+Spec (a :*: b)   = Spec a :*: Spec b
+Spec (a -> b)    = (a, Spec a) -> Demand b -> Thunk (Demand a)
+\end{minted}
+
+We use the data type \verb|Thunk| to introduce an additional layer of
+indirection. A value of type \verb|Demand a| can only ever be a prefix of its
+corresponding value at the type \verb|a|, and thunks represent the locations
+where the remaining structure can be chopped off by a lack of further evaluation
+of the data structure.
+
+With higher-order functions, uncurrying all of the arguments of the higher order
+function before applying \verb|Spec| is critical to generate the most
+fine-grained type of its demand specification. Consider the example \verb|map|
+which maps a function over a list.
+\begin{minted}{haskell}
+map :: (a -> b) -> [a] -> [b]
+\end{minted}
+
+If we don't uncurry \verb|map|, then its specification will have type
+\begin{figure}[H]
+  \centering
+\begin{minted}{haskell}
+(a -> b, (a, Spec a) -> Demand b
+                     -> Thunk (Demand a))
+-> Demand ([a] -> [b])
+-> Thunk (Demand (a -> b))
+\end{minted}
+\caption{curried \texttt{map} specification type}
+\end{figure}
+
+This type just simplifies to
+\begin{minted}{haskell}
+(a -> b, (a, Spec a) -> Demand b
+                     -> Thunk (Demand a))
+-> FuncDemand
+-> Thunk FuncDemand
+\end{minted}
+
+\verb|FuncDemand| only captures whether the function was called or not. However,
+in a typical implementation of \verb|map|, the usage of the input function
+depends on whether the input list is empty or not. It's impossible to specify
+the demand behavior of Haskell standard library's implementation of \verb|map|
+using this type signature.
+
+However, consider the uncurried \verb|map|, which has type
+\begin{minted}{haskell}
+(a -> b, [a]) -> [b]
+\end{minted}
+
+\clearpage
+Applying \verb|Spec| to this type produces
+\begin{figure}[H]
+  \centering
+\begin{minted}{haskell}
+((a -> b, [a]), ((a, Spec a) -> Demand b
+                             -> Thunk (Demand a), Spec [a]))
+-> Demand [b]
+-> Thunk (Demand [a])
+\end{minted}
+\caption{uncurried \texttt{map} specification type}
+\end{figure}
+
+Here, we see that this type in fact takes the input list \verb|[a]| as an input
+to the specification as well! This allows a precise and correct specification of
+\verb|map|. This observation applies to general higher-order functions.
+
+We will integrate the \verb|Spec| and \verb|Demand| Type Families with GHC's
+generics mechanism to automatically derive specification types of algebraic data
+types. We will then write generic generators for representations of
+specifications and demands, and then we will develop a presentable API for
+programmers to write property based test cases to check the demand behaviors of
+Haskell programs.
 \end{document}

### 2017-12-03 presentations/slides.tex
modified slides for 700 presentation

--- 2017-11-20 slides.tex
@@
+\newif\ifFinalProject
+\FinalProjecttrue
+
+\newif\ifNotFinalProject
+\ifFinalProject\NotFinalProjectfalse\else\NotFinalProjecttrue\fi
+
 \documentclass{beamer}
@@
 
+% Slides for CIS700 Final Presentation
+\ifFinalProject
+\begin{frame}[fragile]
+\frametitle{Lazy Evaluation?}
+\begin{itemize}
+\item<1->[]
+\begin{minted}{haskell}
+fibonacci =
+  0 : 1 : (zipWith (+) fibonacci (tail fibonacci))
+\end{minted}
+\item<2->[]
+\begin{minted}{haskell}
+ghci> fibonacci !! 10
+ghci> 55  
+\end{minted}
+\item<3->[]
+\begin{minted}{haskell}
+ghci> fibonacci !! 11
+ghci> 89
+\end{minted}
+\end{itemize}
+\end{frame}
+
+\note[itemize]{
+\item Let's start with an example
+\item \texttt{fibonacci} builds a Haskell list, where \text{:} is list cons constructor
+\item \texttt{fibonacci} builds a circular structure that contains a reference to itself
+\item This is an infinite list
+\item but its definition doesn't cause the interpreter to go to an infinite loop
+\item Core idea is: code doesn't execute unless output is used
+}
+
+\begin{frame}[fragile]
+  \frametitle{How to \textit{demand} a lazy value?}
+  \centering
+  \begin{figure}
+    \includegraphics[width=0.75\textwidth]{binary-tree}
+  \end{figure}
+\end{frame}
+
+\begin{frame}[fragile]
+  \centering
+  \begin{figure}
+    \includegraphics[width=0.75\textwidth]{binary-tree-filled-1}
+  \end{figure}
+\end{frame}
+
+\begin{frame}[fragile]
+  \centering
+  \begin{figure}
+    \includegraphics[width=0.75\textwidth]{binary-tree-filled-2}
+  \end{figure}
+\end{frame}
+
+\begin{frame}[fragile]
+  \centering
+  \begin{figure}
+    \includegraphics[width=0.75\textwidth]{binary-tree-filled-3}
+  \end{figure}
+\end{frame}
+
+\note[itemize]{
+\item A lazy value can be thought of as a pointer graph with regions that have
+  not matierialized/evaluated yet
+\item In this graph, we use white regions to represent parts of the value that
+  have not been evaluated yet
+\item We evaluate more of the value by following the pointers
+}
+
+\begin{frame}[fragile]
+\frametitle{Pattern Matching}
+\begin{minted}{haskell}
+tail :: [Int] -> [Int]
+tail xs = case xs of
+            [] -> []
+            (_ : xs) -> xs
+\end{minted}
+\end{frame}
+
+\note[itemize]{
+\item In a functional language like Haskell, we traverse the pointer graphs of
+  lazy values by performing pattern matching on them
+\item In the \texttt{tail} function for lists, we evaluate the input list to its
+  first constructor
+\item Useful: think about Randoop, it iteratively generates new API calls, and
+  alternates between generating and testing.
+\item If Randoop were able to generate these call sequences lazily, then from
+  Randoop developer's perspective there can just be a module that focuses on
+  generation and one that focuses on testing
+\item Since unused generated API calls are not actually evaluated, it doesn't
+  cost additional CPU resource
+}
+
+% End of slides for CIS700 Final Presentation
+\fi
+
 \begin{frame}
@@
 
+\ifNotFinalProject
 \begin{frame}[fragile]

### 2017-12-21 paper/main.tex
Initial paper files with rough itemized skeleton

% !!! PLEASE DON'T CHANGE THESE !!! INSTEAD DEFINE YOUR OWN texdirectives.tex !!!
% These flags are for the final camera ready version for POPL
% Draft, full version flags
\newif\ifdraft\drafttrue
\newif\iffull\fullfalse
\newif\ifanon\anonfalse
\newif\iflast\lasttrue

% !!! PLEASE DON'T CHANGE THESE !!! INSTEAD DEFINE YOUR OWN texdirectives.tex !!!
\makeatletter \@input{texdirectives} \makeatother

\documentclass[acmsmall,review]{acmart}\settopmatter{}

%% Note: Authors migrating a paper from PACMPL format to traditional
%% SIGPLAN proceedings format should change 'acmlarge' to
%% 'sigplan,10pt'.

%% Some recommended packages.
\usepackage{booktabs}   %% For formal tables:
                        %% http://ctan.org/pkg/booktabs
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
                        %% http://ctan.org/pkg/subcaption

\usepackage{xspace}
\usepackage{preamble}
\usepackage{relsize}
%\usepackage{mathpartir}

\makeatletter\if@ACM@journal\makeatother
%% Journal information (used by PACMPL format)
%% Supplied to authors by publisher for camera-ready submission
% \setcopyright{rightsretained}
% \acmJournal{PACMPL}
% \acmYear{2018} \copyrightyear{2018} \acmVolume{2} \acmNumber{POPL} \acmArticle{45} \acmMonth{1} \acmPrice{}\acmDOI{10.1145/3158133}

%% Copyright information
%% Supplied to authors (based on authors' rights management selection;
%% see authors.acm.org) by publisher for camera-ready submission
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\copyrightyear{2017}           %% If different from \acmYear

%% Bibliography style
\bibliographystyle{ACM-Reference-Format}
%% Citation style
%% Note: author/year citations are required for papers published as an
%% issue of PACMPL.
\citestyle{acmauthoryear}   %% For author/year citations

\def\l{l}

\begin{document}

%% Title information
\title{Keep your Laziness in Check}
                                        %% when present, will be used in
                                        %% header instead of Full Title.
%\titlenote{with title note}             %% \titlenote is optional;
%                                        %% can be repeated if necessary;
%                                        %% contents suppressed with 'anonymous'
%\subtitle{}                     %% \subtitle is optional
%\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

%% Author information
%% Contents and number of authors suppressed with 'anonymous'.
%% Each author should be introduced by \author, followed by
%% \authornote (optional), \orcid (optional), \affiliation, and
%% \email.
%% An author may have multiple affiliations and/or emails; repeat the
%% appropriate command.
%% Many elements are not rendered, but should be provided for metadata
%% extraction tools.

%% Author with single affiliation.
\author{Kenneth Fonner}
%\authornote{with author1 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
%\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
%  \position{Position1}
%  \department{Department1}              %% \department is recommended
  \institution{University of Pennsylvania}            %% \institution is required
%  \streetaddress{Street1 Address1}
%  \city{City1}
%  \state{State1}
%  \postcode{Post-Code1}
  \country{USA}
}
\email{kwf@very.science}          %% \email is recommended

%% Author with single affiliation.
\author{Hengchu Zhang}
%\authornote{with author1 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
%\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
%  \position{Position1}
%  \department{Department1}              %% \department is recommended
  \institution{University of Pennsylvania}            %% \institution is required
%  \streetaddress{Street1 Address1}
%  \city{City1}
%  \state{State1}
%  \postcode{Post-Code1}
  \country{USA}
}
\email{hengchu@seas.upenn.edu}          %% \email is recommended

%% Author with single affiliation.
\author{Leonidas Lampropoulos}
%\authornote{with author1 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
%\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
%  \position{Position1}
%  \department{Department1}              %% \department is recommended
  \institution{University of Pennsylvania}            %% \institution is required
%  \streetaddress{Street1 Address1}
%  \city{City1}
%  \state{State1}
%  \postcode{Post-Code1}
  \country{USA}
}
\email{llamp@seas.upenn.edu}          %% \email is recommended

%% Paper note
%% The \thanks command may be used to create a "paper note" ---
%% similar to a title note or an author note, but not explicitly
%% associated with a particular element.  It will appear immediately
%% above the permission/copyright statement.
%\thanks{with paper note}                %% \thanks is optional
                                        %% can be repeated if necesary
                                        %% contents suppressed with 'anonymous'

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
StrictCheck.

Because LazyCheck was taken.
\end{abstract}

%% 2012 ACM Computing Classification System (CSS) concepts
%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10011007.10011006.10011008</concept_id>
<concept_desc>Software and its engineering~General programming languages</concept_desc>
<concept_significance>500</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Software and its engineering~General programming languages}
%% End of generated code

%% Keywords
%% comma separated list
\keywords{Random Testing, Laziness, Haskell}  %% \keywords is optional

%% \maketitle
%% Note: \maketitle command must come after title commands, author
%% commands, abstract environment, Computing Classification System
%% environment and commands, and keywords command.
\maketitle

% \lk{
% \begin{description}
%   \item[slant] \the\fontdimen1\font
%   \item[interword space] \the\fontdimen2\font
%   \item[interword stretch] \the\fontdimen3\font
%   \item[interword shrink] \the\fontdimen4\font
%   \item[extra space] \the\fontdimen7\font
%   \item[xspaceskip] \the\xspaceskip
%   \item[hyphenchar] \the\hyphenchar\font
% \end{description}
% }

\section{Introduction}
\label{sec:intro}

\begin{itemize}
\item Laziness is a fundamental feature in Haskell \leo{WhyFPmatters?}
\item However, reasoning about programs with laziness (performance, behavior) is hard.
\item It is very easy to use a too-lazy or too-strict implementation that yields different results.
\item FRP quote here?
\item
\item Example: folds.
\item \lk{fold} to calculate sum breaks for large lists because of thunks
\item \lk{fold'} is good for some, but not for average because of whnf.
\item We need an implementation that is strict in the pair as well.
\item
\item Which leads to the natural question, how can we test the laziness behavior of functions?
\item Even worse, how can we even {\em specify} the laziness behavior?
\end{itemize}

Our technical contributions are as follows:
\begin{itemize}
\item We introduce a novel approach to specifying the laziness behavior of functions that is
{\em precise} \leo{define}. \leo{Also add drawbacks/partiality here? If not here, where in the intro?}
\item We propose a random testing approach to test such specifications.
\item We implement StrictCheck. \leo{In order for these last two points to be different, we need
something about the cool/awesome hacks in the implementation to point out here}
\item As a side contribution? We fix the function generaton of QuickCheck to produce lazy/partial functions instead
of fully strict ones all the time.
\end{itemize}
%
Section~\ref{sec:related} discusses related work. We conclude and draw
directions for future work in Section~\ref{sec:concl}.

\section{Strictness Specifications, by Example}
\label{sec:quickchick}

\leo{Punned names for sections? Or is that just an urns thing? :P }

\begin{itemize}
\item In this section we will showcase our proposed specification language \leo{It's not a language tho}
\item while introducing the running example of our paper: maps.
\item Maps are particularly well-suited for a discussion of laziness because of the interesting different
possible behaviors: strict-maps, value-lazy maps \leo{Point out Data.Map.Strict vs Data.Map}, and fully lazy maps
\item For simplicity, we will use binary search trees as maps, assuming the keys are integers and forego balancing
as the added complexity is irrelevant to the laziness discussion. \leo{Can we claim it wouldn't matter at all? I think so}
\end{itemize}

\paragraph*{Binary Search Trees}

\leo{data structures - code - short explanation}

\paragraph*{Characterizing Laziness}

\begin{itemize}
\item Specification of laziness - TreeD data structure \leo{Mention how it will be generically derived}
\item Strict / Spine-strict maps
\item Fully lazy map as a between spec
\end{itemize}

\section{Evaluation}
\label{sec:eval}

\section{Related Work}
\label{sec:related}

\leo{If we want to break up this into smaller chunks, use paragraph*}

\section{Conclusion and future work}
\label{sec:concl}

Machine Learning for automatically synthesizing specs? Let's go crazy!

%% Acknowledgments
\begin{acks}                            %% acks environment is optional
                                        %% contents suppressed with 'anonymous'
  %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
  %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
  %% acknowledge financial support and will be used by metadata
  %% extraction tools.
We are grateful to
%
Matt Parsons,
Matthew Weaver,
Jennifer Paykin,
Antal Spector-Zabusky,
and the Penn PLClub 
for their useful comments. 
This material is based upon work supported by the
\grantsponsor{GS100000001}{National Science
  Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
No.~\grantnum{GS100000001}{1421243} ({\em Random Testing for Language
Design}), Grant No.~\grantnum{GS100000001}{1521523} ({\em Expeditions
in Computing: The Science of Deep Specification}), and \leo{Kenny,
Hengchu, ask for grant numbers names} Any opinions, findings, and
conclusions or recommendations expressed in this material are those of
the author and do not necessarily reflect the views of the National
Science Foundation.

\end{acks}

%% Bibliography
%\bibliography{bibfile}
\bibliography{local}

%% Appendix
%\appendix
%\section{Appendix}
% 
%Text of appendix \ldots

\end{document}

### 2017-12-29 paper/main.tex
[paper] Some more work on the intro example

@@
 \item FRP quote here?
-\item
-\item Example: folds.
-\item \lk{fold} to calculate sum breaks for large lists because of thunks
-\item \lk{fold'} is good for some, but not for average because of whnf.
-\item We need an implementation that is strict in the pair as well.
-\item
+\end{itemize}
+
+Consider the following implementation of the \lk{sum} function:
+\leo{I created an environment for this so that we can tweak them all as needed}
+\begin{inlinecode}
+sum = foldl (+) 0
+\end{inlinecode}
+%
+\begin{itemize}
+\item As many readers/Haskellers have probably encountered, such an implementation
+   can be inefficient due to space leaks because of laziness.
+\item When summing over a large list, instead of calculating the running sum as we traverse
+   the list, we build an increasingly larger thunk that will be forced in the end.
+\item An easy solution is to use a stricter fold, like \lk{foldl'}, which forces the
+   fold accumulator in weak head normal form.
+\item However, even that is not enough for all cases.
+\item Consider a single-traversal computation of an average where we keep both a running sum and a running length in an accumulator.
+\end{itemize}
+%
+\begin{inlinecode}
+avg = uncurry (/) . foldl (\(s,n) x -> (s + x, n + 1)) (0,0)
+\end{inlinecode}
+%
+\begin{itemize}
+\item In this case, both \lk{foldl} and \lk{foldl'}, construct a thunk, since forcing a pair into weak head normal form does not force either component.
+\item We need an implementation that is strict in those components as well.
 \item Which leads to the natural question, how can we test the laziness behavior of functions?
@@
 \end{itemize}
+
+\leo{We probably need a paragraph that explains what we want to do here, not directly a jump into the contributions}
 

### 2018-01-04 paper/main.tex
working on the intro

--- 2017-12-21 paper/main.tex
@@
 
+Laziness is ubiquitous in modern software engineering. The big-data framework
+Spark uses laziness to drive computations from demand (\hzh{find references to
+spark}), Apple's new programming language Swift includes lazy evaluation for
+initializing expensive resources only when they are needed, and many functional
+reactive programming frameworks for the modern web use laziness.
+
+However, reasoning about the performance of programs with laziness is very
+hard. It is very easy to accidentally make a program too lazy or too strict, and
+this often makes the program behave differently from what the programmer had
+intended.
+
+Since laziness is a fundamental feature of the programming language Haskell, we
+will explore how to observe and specify laziness using just the features
+provided by Haskell, without custom compiler support.
+
+\leo{
 \begin{itemize}
-\item Laziness is a fundamental feature in Haskell \leo{WhyFPmatters?}
+\item Laziness is a fundamental feature in Haskell (quote WhyFPmatters?)
 \item However, reasoning about programs with laziness (performance, behavior) is hard.
@@
 \item FRP quote here?
-\item
-\item Example: folds.
-\item \lk{fold} to calculate sum breaks for large lists because of thunks
-\item \lk{fold'} is good for some, but not for average because of whnf.
-\item We need an implementation that is strict in the pair as well.
-\item
-\item Which leads to the natural question, how can we test the laziness behavior of functions?
-\item Even worse, how can we even {\em specify} the laziness behavior?
 \end{itemize}
+}
+
+Consider the following implementation of the \lk{sum} function:
+\leo{I created an environment for this so that we can tweak them all as needed}
+\begin{inlinecode}
+sum = foldl (+) 0
+\end{inlinecode}
+
+Although this seemingly innocent implementation of sum will produce correct
+results on small lists of numbers, it will quickly run out of memory on larger
+lists. The culprit here is the strictness behavior of \lk{foldl}---it does not
+force the accumulated result during the fold. Calling this \lk{sum} on a list
+will produce a thunk that is linear in the size of of the input list.
+
+We can fix this space leak for \lk{sum} using the strict version \lk{foldl'},
+which forces the accumulator. However, even this is not enough in
+general. Consider a single-traversal computation of an average where we keep
+both a running sum and a running length in an accumulator.
+
+\begin{inlinecode}
+avg = uncurry (/) . foldl (\(s,n) x -> (s + x, n + 1)) (0,0)
+\end{inlinecode}
+
+In this case, both \lk{foldl} and \lk{foldl'} construct thunks linear in the
+size of the input list, since forcing a pair into weak head normal form does not
+force either component.
+
+How can we test and catch these laziness bugs before they manifest themselves in
+production? Is it possible to specify the laziness behavior using a precise
+language just as we would for functional correctness?
+
+\leo{We probably need a paragraph that explains what we want to do here,
+not directly a jump into the contributions}
+
+In this paper, we present solutions to these questions. We will demonstrate how
+to observe laziness from within Haskell itself, and how to write down a
+specification for laziness also using Haskell.
 
@@
 \item We propose a random testing approach to test such specifications.
-\item We implement StrictCheck. \leo{In order for these last two points to be different, we need
+\item We demonstrate non-trivial generic programming techniques to apply StrictCheck to almost all data types. \leo{In order for these last two points to be different, we need
 something about the cool/awesome hacks in the implementation to point out here}

### 2018-01-09 paper/main.tex
More work on the intro.

--- 2017-12-21 paper/main.tex
@@
 
-\begin{itemize}
-\item Laziness is a fundamental feature in Haskell \leo{WhyFPmatters?}
-\item However, reasoning about programs with laziness (performance, behavior) is hard.
-\item It is very easy to use a too-lazy or too-strict implementation that yields different results.
-\item FRP quote here?
-\item
-\item Example: folds.
-\item \lk{fold} to calculate sum breaks for large lists because of thunks
-\item \lk{fold'} is good for some, but not for average because of whnf.
-\item We need an implementation that is strict in the pair as well.
-\item
-\item Which leads to the natural question, how can we test the laziness behavior of functions?
-\item Even worse, how can we even {\em specify} the laziness behavior?
-\end{itemize}
+\leo{I want to keep a counter on how many times we'll rewrite this intro :)}
+\leo{We *have* to mention Haskell in this list. Maybe with John's Why FP matters?}
+Laziness is ubiquitous in modern software engineering: the big-data
+framework Spark uses laziness to drive computations on demand
+(\hzh{find references to spark}), Apple's new programming language
+Swift includes lazy evaluation for initializing expensive resources as
+needed\cn, and many functional reactive programming web frameworks rely
+on it\cn. 
+%
+Unfortunately, reasoning about the performance of lazy programs with
+laziness is hard. It is very easy\leo{I don't like 'easy' here. It has
+a slightly bad antithesis with the previous 'hard' as it doesn't refer
+to the same thing} to accidentally make a program too lazy or too
+strict, which in turn makes the program behave differently than 
+the programmer had intended.
+
+\leo{Didn't we have an FRP quote by someone that Laziness was the third source of
+bugs or something? Work that in?}
+
+\hzh{Since laziness is a fundamental feature of the programming language Haskell, we
+will explore how to observe and specify laziness using just the features
+provided by Haskell, without custom compiler support.}
+\leo{It feels more natural to go directly into the example that explains this difficulty.
+Mentioning haskell needs to happen earlier, maybe in the original
+list. Or in a ``starting with Haskell'' way}
+
+For example, consider the following simple Haskell implementation of the \lk{sum} function:
+\leo{I created an environment for this so that we can tweak them all as needed}
+\begin{inlinecode}
+sum = foldl (+) 0
+\end{inlinecode}
+%
+Although this seemingly innocent implementation of sum will produce correct
+results on small lists of numbers, it will quickly run out of memory on larger
+lists. The culprit here is the strictness behavior of \lk{foldl}---it does not
+force the accumulated result during the fold. Instead, calling this \lk{sum} on a list
+produces a thunk that is linear in the size of of the input list.
+%
+We can fix this space leak for \lk{sum} using a strict fold, \lk{foldl'},
+which forces the accumulator.
+
+However, even this is not enough in general. Consider a
+single-traversal computation of an average where we keep both a
+running sum and a running length in an accumulator.
+%
+\begin{inlinecode}
+avg = uncurry (/) . foldl (\(s,n) x -> (s + x, n + 1)) (0,0)
+\end{inlinecode}
+%
+In this case, both \lk{foldl} and \lk{foldl'} construct thunks linear in the
+size of the input list, since forcing a pair into weak head normal form does not
+force either component\leo{of the pair?}.
+
+How, then, can we test and catch these laziness bugs before they
+manifest themselves in production? Is it possible to precisely specify
+laziness behavior just as we would functional correctness?
+%
+\leo{Rewrite the next.}
+In this paper, we demonstrate how to observe laziness from within
+Haskell itself, and how to write down a specification for laziness
+also using Haskell.
 

### 2018-01-11 paper/main.tex
rewriting intro

--- 2017-12-29 paper/main.tex
@@
 
-\begin{itemize}
-\item Laziness is a fundamental feature in Haskell \leo{WhyFPmatters?}
-\item However, reasoning about programs with laziness (performance, behavior) is hard.
-\item It is very easy to use a too-lazy or too-strict implementation that yields different results.
-\item FRP quote here?
-\end{itemize}
+\hzh{Rewrite counter: 1}
+\leo{I want to keep a counter on how many times we'll rewrite this intro :)}
 
-Consider the following implementation of the \lk{sum} function:
+John Hughes has argued adequately that lazy evaluation brings tremendous power
+to modularization of functional programming in his seminal work \textit{Why
+Functional Programming Matters} ~\cite{WhyFPMatters}. Haskell is one of the most
+popular programming languages that gives its programmers the power of lazy
+evaluation.
+%
+Unfortunately, reasoning about the performance of lazy programs in Haskell is
+difficult. Programmers often accidentally make a program too lazy or too strict,
+and as a result, their programs behave differently than expected.
+
+\leo{Didn't we have an FRP quote by someone that Laziness was the third source of
+bugs or something? Work that in?}
+\hzh{Ryan Trinkle (author of Reflex) said laziness was the 3rd most difficult thing to
+get right, but let's check with him on whether we can quote him on that.}
+
+For example, consider the following simple Haskell implementation of the \lk{sum} function:
 \leo{I created an environment for this so that we can tweak them all as needed}
@@
 %
-\begin{itemize}
-\item As many readers/Haskellers have probably encountered, such an implementation
-   can be inefficient due to space leaks because of laziness.
-\item When summing over a large list, instead of calculating the running sum as we traverse
-   the list, we build an increasingly larger thunk that will be forced in the end.
-\item An easy solution is to use a stricter fold, like \lk{foldl'}, which forces the
-   fold accumulator in weak head normal form.
-\item However, even that is not enough for all cases.
-\item Consider a single-traversal computation of an average where we keep both a running sum and a running length in an accumulator.
-\end{itemize}
+Although this seemingly innocent implementation of sum will produce correct
+results on small lists of numbers, it will quickly run out of memory on larger
+lists. The culprit here is the laziness behavior of \lk{foldl}---it does not
+force the accumulated result during the fold. Instead, calling this \lk{sum} on
+a list produces a thunk that is linear in the size of of the input list, and
+this thunk will not be evaluated until the result of sum is used in another
+computation.
+%
+We can fix this space leak for \lk{sum} using a strict fold, \lk{foldl'}, which
+eagerly forces the accumulator, and avoids creating a space-leaking thunk.
+
+However, using \lk{foldl'} is not sufficient to avoid space leaks in
+general. Consider a single-traversal computation of an average where we keep
+both a running sum and a running counter in an accumulator.
 %
@@
 %
-\begin{itemize}
-\item In this case, both \lk{foldl} and \lk{foldl'}, construct a thunk, since forcing a pair into weak head normal form does not force either component.
-\item We need an implementation that is strict in those components as well.
-\item Which leads to the natural question, how can we test the laziness behavior of functions?
-\item Even worse, how can we even {\em specify} the laziness behavior?
-\end{itemize}
+In this case, both \lk{foldl} and \lk{foldl'} construct thunks linear in the
+size of the input list, since forcing a lazy pair only evaluates it to its weak
+head normal form, and it does not evaluate either component of the pair.
 
-\leo{We probably need a paragraph that explains what we want to do here, not directly a jump into the contributions}
+How, then, can we test and catch these laziness bugs before they
+manifest themselves in production? Is it possible to precisely specify
+laziness behavior just as we would for functional correctness?
+%
+In this paper, we demonstrate techniques for observing laziness in Haskell, and
+propose a system of creating specifications of lazy Haskell programs. We
+implement these ideas as a Haskell library.
 
@@
 \item We introduce a novel approach to specifying the laziness behavior of functions that is
-{\em precise} \leo{define}. \leo{Also add drawbacks/partiality here? If not here, where in the intro?}
-\item We propose a random testing approach to test such specifications.
-\item We implement StrictCheck. \leo{In order for these last two points to be different, we need
-something about the cool/awesome hacks in the implementation to point out here}
-\item As a side contribution? We fix the function generaton of QuickCheck to produce lazy/partial functions instead
-of fully strict ones all the time.
+{\em precise}. Such specifications are expressive enough to capture exactly how
+much evaluation occurs within a data structure \leo{Also add
+drawbacks/partiality here? If not here, where in the intro?} \hzh{I'm not sure
+if we should bring in partiality (in the sense of divergence/nontermination) yet
+since that seems like a rather technical detail of the spec}
+\item We propose a framework for testing against such specifications.
+\item We improve upon the random function generators from QuickCheck to generate functions with arbitrary laziness behaviors.
+\item We demonstrate non-trivial generic programming techniques to apply StrictCheck to almost all data types.
 \end{itemize}

### 2018-01-18 paper/main.tex
fixing Kenny's name
push comments from Tuesday's comments

@@
 %% Author with single affiliation.
-\author{Kenneth Fonner}
+\author{Kenneth Foner}
 %\authornote{with author1 note}          %% \authornote is optional;
@@
 }
-\email{kwf@very.science}          %% \email is recommended
+\email{kfoner@seas.upenn.edu}          %% \email is recommended
 
@@
 \hzh{Rewrite counter: 1}
+\hzh{Rewrite the intro using different running examples, foldl vs foldr on stream manipulations?}
+\hzh{We can write a hand-crafted spec for take, can we show that we can create this spec by composing the spec for foldr (but not using the one with foldl)}
+\hzh{implement take in 3 ways: foldl, foldr and explicitly as the opening example, and then ask if they're equivalent? (they're not!)}
+\hzh{Okasaki has 2 functions that appends a list to the reversal of another list, one of which is too strict and the other one is just the right amount of lazy (thus incremental)}
+\hzh{The amortized one is hard to specify exactly, but we can introduce approximate specification and specify it approximately}
+\hzh{we need to carefully avoid saying this is a tool for specifying the time complexity of functions}
+\hzh{Implementing functional programs that make clever use of laziness is very easy to get wrong, and we provide a solution to make sure the programmer got it right.
+We almost have no way of testing it right now before StrictCheck.}
+\hzh{Email Ryan Trinkle and Dan Winograd-Cort about FRP and strictness, and try to use their examples as another problem we can solve in the intro}
+\hzh{Lazy maps will mess up memoization if you're overly strict, use knapsack as an example}
 \leo{I want to keep a counter on how many times we'll rewrite this intro :)}
@@
 %
+\hzh{TODO: rewrite this, let bind the foldl}
 \begin{inlinecode}
@@
 \item We introduce a novel approach to specifying the laziness behavior of functions that is
-{\em precise}. Such specifications are expressive enough to capture exactly how
-much evaluation occurs within a data structure \leo{Also add
-drawbacks/partiality here? If not here, where in the intro?} \hzh{I'm not sure
-if we should bring in partiality (in the sense of divergence/nontermination) yet
-since that seems like a rather technical detail of the spec}
+{\em precise} \hzh{and compositional}. Such specifications are
+expressive enough to capture exactly how much evaluation occurs within
+a data structure \leo{Also add drawbacks/partiality here? If not here,
+where in the intro?} \hzh{I'm not sure if we should bring in
+partiality (in the sense of divergence/nontermination) yet since that
+seems like a rather technical detail of the spec}
 \item We propose a framework for testing against such specifications.

### 2018-01-23 paper/main.tex
rewrote intro again

@@
 
-\hzh{Rewrite counter: 1}
+\hzh{Rewrite counter: 2}
 \hzh{Rewrite the intro using different running examples, foldl vs foldr on stream manipulations?}
@@
 \hzh{Email Ryan Trinkle and Dan Winograd-Cort about FRP and strictness, and try to use their examples as another problem we can solve in the intro}
-\hzh{Lazy maps will mess up memoization if you're overly strict, use knapsack as an example}
+\hzh{Maps will mess up memoization if you're overly strict, use knapsack as an example}
 \leo{I want to keep a counter on how many times we'll rewrite this intro :)}
 
-John Hughes has argued adequately that lazy evaluation brings tremendous power
-to modularization of functional programming in his seminal work \textit{Why
-Functional Programming Matters} ~\cite{WhyFPMatters}. Haskell is one of the most
-popular programming languages that gives its programmers the power of lazy
-evaluation.
+Lazy evaluation brings tremendous power to modularization of functional
+programming ~\cite{WhyFPMatters}. Haskell is one of the most popular programming
+languages that gives its programmers the power of lazy evaluation.
+
+To demonstrate the benefits of lazy evaluaation, let us consider a dynamic
+programming solution of the knapsack problem in Haskell.
 %
-Unfortunately, reasoning about the performance of lazy programs in Haskell is
-difficult. Programmers often accidentally make a program too lazy or too strict,
-and as a result, their programs behave differently than expected.
+The inputs of the knapsack problem are two sequences of numbers $w_i$ and $v_i$
+where $i = 1, \dots, n$, and each of $w_i$ and $v_i$ specify the weight and the
+value of the $i$th item. The solution presents the maximum value achievable
+using a subset of the items such that their combined weight is under a limit
+$W$.
+
+The typical dynamic programming approach for knapsack maintains a table $OPT[][]$,
+where the entry $OPT[j][k]$ keeps the best value achievable using only items
+$1\dots j$ and a weight limit of $k$. The table satisfies the following
+recurrence relation: $$OPT[j][k] = \max(OPT[j-1][w], w_j + OPT[j-1][k-w_j])$$
+
+With typical imperative languages, we look for shared sub-problems of the
+original knapsack problem, and then we design a path along the solution table
+that maximizes sharing while filling out the answers.
+%
+However, lazy evaluation automatically finds this path if we
+use \lk{Data.Map.Lazy} and setup the solution table with the recurrence
+relation \hzh{We can probably make this code prettier}
+\begin{inlinecode}
+ws, vs :: [Int]
+n,  w  :: Int
+
+solution :: Map (Int, Int) Int
+solution = fromList [ ((j, k),
+                       if (j-1) < 0 || k - ws !! j < 0
+                       then 0
+                       else max (solution ! (j-1, k))
+                                (solution ! (j-1, k - ws !! j) + vs !! j)
+                      )
+                    | j <- [0..2], k <- [0..50] ]
+\end{inlinecode}
+
+Using the lazy variant of \lk{Map} here is critical for the correctness
+of \lk{solution}. If we used \lk{Data.Map.Strict} instead, then \lk{fromList}
+would have eagerly evaluated the \lk{if} expression. The \lk{if} expression
+contains value-level recursion to itself, and this would manifest as an infinite
+loop at runtime. This is an incredibly subtle bug related to laziness.
+
+Unfortunately, reasoning about laziness in Haskell is difficult. Programmers
+often accidentally make a program too lazy or too strict, and as a result, their
+programs behave differently than expected.
 
@@
 
-For example, consider the following simple Haskell implementation of the \lk{sum} function:
-\leo{I created an environment for this so that we can tweak them all as needed}
-\begin{inlinecode}
-sum = foldl (+) 0
-\end{inlinecode}
-%
-Although this seemingly innocent implementation of sum will produce correct
-results on small lists of numbers, it will quickly run out of memory on larger
-lists. The culprit here is the laziness behavior of \lk{foldl}---it does not
-force the accumulated result during the fold. Instead, calling this \lk{sum} on
-a list produces a thunk that is linear in the size of of the input list, and
-this thunk will not be evaluated until the result of sum is used in another
-computation.
-%
-We can fix this space leak for \lk{sum} using a strict fold, \lk{foldl'}, which
-eagerly forces the accumulator, and avoids creating a space-leaking thunk.
-
-However, using \lk{foldl'} is not sufficient to avoid space leaks in
-general. Consider a single-traversal computation of an average where we keep
-both a running sum and a running counter in an accumulator.
-%
-\hzh{TODO: rewrite this, let bind the foldl}
-\begin{inlinecode}
-avg = uncurry (/) . foldl (\(s,n) x -> (s + x, n + 1)) (0,0)
-\end{inlinecode}
-%
-In this case, both \lk{foldl} and \lk{foldl'} construct thunks linear in the
-size of the input list, since forcing a lazy pair only evaluates it to its weak
-head normal form, and it does not evaluate either component of the pair.
-
-How, then, can we test and catch these laziness bugs before they
-manifest themselves in production? Is it possible to precisely specify
-laziness behavior just as we would for functional correctness?
+How, then, can we test and catch these laziness bugs during development? Is it
+possible to precisely specify laziness behavior just as we would for functional
+correctness?
 %

### 2018-01-23 paper/knapsack.hs

@@ -1,0 +1,21 @@
+import Data.Map.Lazy
+
+main = return ()
+
+ws :: [Int]
+ws = [10, 20, 30]
+
+vs :: [Int]
+vs = [60, 100, 120]
+
+w :: Int
+w = 50
+
+solution :: Map (Int, Int) Int
+solution = fromList [ ((j, k),
+                       if (j-1) < 0 || k - ws !! j < 0
+                       then 0
+                       else max (solution ! (j-1, k))
+                                (solution ! (j-1, k - ws !! j) + vs !! j)
+                      )
+                    | j <- [0..2], k <- [0..50] ]

### 2018-01-31 paper/main.tex
Some work on the intro

@@
 
-Lazy evaluation brings tremendous power to modularization of functional
-programming ~\cite{WhyFPMatters}. Haskell is one of the most popular programming
-languages that gives its programmers the power of lazy evaluation.
+\leo{Point of first paragraph: lazy evaluation is powerful}
+
+Lazy evaluation brings tremendous power to modularization of
+functional programming ~\cite{WhyFPMatters} and is actively used in
+various practical applications. For instance, the big-data framework
+Spark uses laziness to drive computations on demand\cn, Apple's
+programming language Swift leverages it to initialize expensive
+resources as needed\cn, and many functional reactive programming web
+frameworks rely on it\cn \leo{This point might need more exposition
+about frp}. Haskell is the most popular programming language that
+allows programmers to enjoy the power of lazy evaluation by default.
+
+\leo{Point of second paragraph: lazy evaluation is hard}
+
+\leo{I'm sorry in advance for the following sentence. It was just too tempting.}
+\begin{itemize}
+\item Unfortunately, such power comes with great responsibility.\leo{-Ben
+Parker :P}
+\item Reasoning about lazy programs is notoriously hard, because
+it is difficult to predict if and when thunks are evaluated.
+\item Moreover, there is \leo{currently?} no way to specify or check the expected
+laziness behavior of a function.
+\item That would be good.
+\end{itemize}
+
+\leo{Example 1: Okasaki queues}
+As a first example of Laziness being both useful and error-prone, we can turn to
+purely functional data structures\cn, and in particular queues.
+\leo{Summarize queue example}
+
+\leo{Example 2: Knapsack}
 

### 2018-02-04 paper/main.tex

--- 2018-01-23 paper/main.tex
@@
 
-\hzh{Rewrite counter: 2}
+\hzh{Rewrite counter: 4}
 \hzh{Rewrite the intro using different running examples, foldl vs foldr on stream manipulations?}
@@
 
-Lazy evaluation brings tremendous power to modularization of functional
-programming ~\cite{WhyFPMatters}. Haskell is one of the most popular programming
-languages that gives its programmers the power of lazy evaluation.
+\leo{Point of first paragraph: lazy evaluation is powerful}
 
-To demonstrate the benefits of lazy evaluaation, let us consider a dynamic
-programming solution of the knapsack problem in Haskell.
-%
-The inputs of the knapsack problem are two sequences of numbers $w_i$ and $v_i$
-where $i = 1, \dots, n$, and each of $w_i$ and $v_i$ specify the weight and the
-value of the $i$th item. The solution presents the maximum value achievable
-using a subset of the items such that their combined weight is under a limit
-$W$.
+Lazy evaluation brings tremendous power to functional programming
+~\cite{WhyFPMatters} and is actively used in various practical
+applications. For instance, the big-data framework Spark uses laziness
+to drive computations on demand\cn, Apple's programming language Swift
+leverages it to initialize expensive resources as needed\cn, and many
+functional reactive programming web frameworks rely on it\cn \leo{This
+point might need more exposition about frp}. Haskell is the most
+popular programming language that allows programmers to enjoy the
+power of lazy evaluation by default. Let us consider a few examples
+that demonstrate the power of laziness, and remind us of the
+intricacies of programming with laziness in Haskell.
 
-The typical dynamic programming approach for knapsack maintains a table $OPT[][]$,
-where the entry $OPT[j][k]$ keeps the best value achievable using only items
-$1\dots j$ and a weight limit of $k$. The table satisfies the following
-recurrence relation: $$OPT[j][k] = \max(OPT[j-1][w], w_j + OPT[j-1][k-w_j])$$
+\leo{Example 1: Okasaki queues}
 
-With typical imperative languages, we look for shared sub-problems of the
-original knapsack problem, and then we design a path along the solution table
-that maximizes sharing while filling out the answers.
+Purely functional queue is a very popular data structure in
+Haskell. Programmers typically implement such queues with two
+lists---a front list holding the queue elements in FIFO order, and a
+back list holding the other elments in reversed order. When the front
+list is empty, a dequeue operation requires a reversal of the back
+list to refill the front list. These unlucky dequeue operations take
+$O(n)$ time to finish, and are much slower than their imperative
+counterparts.
+
+However, Okasaki designed a very clever purely functional queue using
+lazy lists that both incrementally performs the reversal of the back
+list, and incrementally appends the reversal of the back list onto the
+front list ~\cite{Okasaki95simpleand}. Okasaki's queue relies on the
+following \lk{rotate} function to incrementally reverse and append the
+back list.
+\begin{inlinecode}
+rotate []    back acc = (head back)  : acc
+rotate front back acc = (head front) : rest
+  where rest = rotate (tail front) (tail back) ((head back) : acc)
+\end{inlinecode}
+If we computed the expression \lk{rotate front back []}, its value
+will be equivalent to \lk{front ++ (reverse back)}. However, the
+function \lk{rotate} is much lazier than the naive reverse and append
+operation. The function \lk{reverse} has to evaluate the entire
+structure of the back list, which is not incremental. On the other
+hand, the function \lk{rotate} produces a list immediately, and packs
+the rest of the computation into the lazy list \lk{rest}.
+
+Since \lk{rotate} is functionally equivalent to
+\lk{front ++ (reverse back)}, a programmer could have implemented it naively
+and never discover the mistake through traditional testing! The two
+only differ in their laziness behavior, and there is no way of
+specifying such subtle differences in any testing frameworks today.
+
+\leo{Example 2: Knapsack}
+
+Laziness also shines in solutions to dynamic programming problems. Let
+us consider a solution to the famous knapsack problem that levareges
+laziness.
+
+The inputs of the knapsack problem are two sequences of numbers $w_i$
+and $v_i$ where $i = 1, \dots, n$, and each of $w_i$ and $v_i$ specify
+the weight and the value of the $i$th item. The solution presents the
+maximum value achievable using a subset of the items such that their
+combined weight is under a limit $W$.
+
+The usual dynamic programming approach for knapsack maintains a
+table $OPT[][]$, where the entry $OPT[j][k]$ keeps the best value
+achievable using only items $1\dots j$ and a weight limit of $k$. The
+table satisfies the following recurrence relation: $$OPT[j][k]
+= \max(OPT[j-1][w], w_j + OPT[j-1][k-w_j])$$
+
+To efficiently compute this table using imperative langauges, we look
+for shared sub-problems of the original problem, and design a path
+along the solution table that maximizes sharing while filling out the
+answers.
 %
 However, lazy evaluation automatically finds this path if we
-use \lk{Data.Map.Lazy} and setup the solution table with the recurrence
-relation \hzh{We can probably make this code prettier}
+use \lk{Data.Map.Lazy} and setup the solution table with the
+recurrence relation \hzh{We can probably make this code prettier}
 \begin{inlinecode}
@@
                       )
-                    | j <- [0..2], k <- [0..50] ]
+                    | j <- [0..n-1], k <- [0..w] ]
 \end{inlinecode}
@@
 
-Unfortunately, reasoning about laziness in Haskell is difficult. Programmers
-often accidentally make a program too lazy or too strict, and as a result, their
-programs behave differently than expected.
+As these examples show, the power of laziness puts great
+responsibility on programmers' shoulders. An incorrect implementation
+can either behave functionally the same but with inferior performance,
+or it can completely break down! Reasoning about lazy programs is also
+notoriously hard because predicting when a lazy value gets evaluated
+is difficult.
 

### 2018-02-06 paper/main.tex

@@
 
-\hzh{Rewrite counter: 4}
-\hzh{Rewrite the intro using different running examples, foldl vs foldr on stream manipulations?}
-\hzh{We can write a hand-crafted spec for take, can we show that we can create this spec by composing the spec for foldr (but not using the one with foldl)}
-\hzh{implement take in 3 ways: foldl, foldr and explicitly as the opening example, and then ask if they're equivalent? (they're not!)}
+%\hzh{Rewrite the intro using different running examples, foldl vs foldr on stream manipulations?}
+%\hzh{We can write a hand-crafted spec for take, can we show that we can create this spec by composing the spec for foldr (but not using the one with foldl)}
+%\hzh{implement take in 3 ways: foldl, foldr and explicitly as the opening example, and then ask if they're equivalent? (they're not!)}
 \hzh{Okasaki has 2 functions that appends a list to the reversal of another list, one of which is too strict and the other one is just the right amount of lazy (thus incremental)}
@@
 \hzh{Maps will mess up memoization if you're overly strict, use knapsack as an example}
-\leo{I want to keep a counter on how many times we'll rewrite this intro :)}
 
@@
 
-Lazy evaluation brings tremendous power to functional programming
+Lazy evaluation gives tremendous power to functional programmers
 ~\cite{WhyFPMatters} and is actively used in various practical
@@
 leverages it to initialize expensive resources as needed\cn, and many
-functional reactive programming web frameworks rely on it\cn \leo{This
-point might need more exposition about frp}. Haskell is the most
-popular programming language that allows programmers to enjoy the
-power of lazy evaluation by default. Let us consider a few examples
-that demonstrate the power of laziness, and remind us of the
-intricacies of programming with laziness in Haskell.
+functional reactive programming web frameworks crucially rely on
+it\cn \leo{This point might need more exposition about frp}. Haskell
+is the most popular programming language that allows programmers to
+enjoy the power of lazy evaluation by default. However, that power
+does not come without a price: reasoning about lazy programs is
+notoriously hard. Let us consider two examples that demonstrate the
+power of laziness, as well as the intricacies of programming with it.
 
@@
 
+\leo{The singular feels weird here. Is it just me?}
 Purely functional queue is a very popular data structure in
@@
 list is empty, a dequeue operation requires a reversal of the back
-list to refill the front list. These unlucky dequeue operations take
+list to refill the front list. These specific dequeue operations take
 $O(n)$ time to finish, and are much slower than their imperative
-counterparts.
+counterparts. This occasional performance degradation is compounded by
+persistence (where the $O(n)$ operation could be repeated many times),
+and can be unacceptable in real-time systems.
 
-However, Okasaki designed a very clever purely functional queue using
-lazy lists that both incrementally performs the reversal of the back
-list, and incrementally appends the reversal of the back list onto the
-front list ~\cite{Okasaki95simpleand}. Okasaki's queue relies on the
+However, Okasaki cleverly improved upon this design using lazy lists
+that both incrementally performs the reversal of the back list, and
+incrementally appends the reversal of the back list onto the front
+list ~\cite{Okasaki95simpleand}. Okasaki's queue relies on the
 following \lk{rotate} function to incrementally reverse and append the
@@
 
-Since \lk{rotate} is functionally equivalent to
-\lk{front ++ (reverse back)}, a programmer could have implemented it naively
-and never discover the mistake through traditional testing! The two
-only differ in their laziness behavior, and there is no way of
-specifying such subtle differences in any testing frameworks today.
+Since \lk{rotate} is functionally equivalent to \lk{front ++ (reverse
+back)}, a programmer could have implemented it naively and never
+discover the mistake through traditional testing! The two only differ
+in their laziness behavior, and there is no way of specifying such
+subtle differences in any testing frameworks today.
 

### 2018-02-11 paper/main.tex
backup

@@
 \begin{inlinecode}
-rotate []    back acc = (head back)  : acc
-rotate front back acc = (head front) : rest
-  where rest = rotate (tail front) (tail back) ((head back) : acc)
+rotate :: [a] -> [a] -> [a] -> [a]
+rotate      []       []  as =                       as
+rotate      []  (b : bs) as =     rotate [] bs (b : as)
+rotate (f : fs)      []  as = f : rotate fs []      as
+rotate (f : fs) (b : bs) as = f : rotate fs bs (b : as)
 \end{inlinecode}
@@
 %
+José Manuel Calderón Trilla,
 Matt Parsons,
+Stephanie Weirich,
+Jennifer Paykin,
+Robert Rand,
+Antal Spector-Zabusky,
 Matthew Weaver,
-Jennifer Paykin,
-Antal Spector-Zabusky,
 and the Penn PLClub

### 2018-02-12 paper/main.tex
[minor] Add acknowledgement for Robert and José
[paper] Lazy/Strict Maps

@@
 \leo{The singular feels weird here. Is it just me?}
-Purely functional queue is a very popular data structure in
-Haskell. Programmers typically implement such queues with two
-lists---a front list holding the queue elements in FIFO order, and a
-back list holding the other elments in reversed order. When the front
-list is empty, a dequeue operation requires a reversal of the back
-list to refill the front list. These specific dequeue operations take
-$O(n)$ time to finish, and are much slower than their imperative
-counterparts. This occasional performance degradation is compounded by
-persistence (where the $O(n)$ operation could be repeated many times),
-and can be unacceptable in real-time systems.
+As a first example, consider purely functional queues, a very
+popular data structure in Haskell. Programmers typically implement
+such queues with two lists---a front list holding the queue elements
+in FIFO order, and a back list holding the other elments in reversed
+order. When the front list is empty, a dequeue operation requires a
+reversal of the back list to refill the front list. These specific
+dequeue operations take $O(n)$ time to finish, and are much slower
+than their imperative counterparts. This occasional performance
+degradation is compounded by persistence (where the $O(n)$ operation
+could be repeated many times), and can be unacceptable in real-time
+systems.
 
@@
 
-\leo{Example 2: Knapsack}
+\leo{Example 2: Maps / Knapsack}
 
-Laziness also shines in solutions to dynamic programming problems. Let
-us consider a solution to the famous knapsack problem that levareges
-laziness.
+For the second example, we turn to a different data functional data
+structure where both a strict and a lazy variant can be situationally
+useful: maps. There are two reasonable laziness behaviors that
+implement a map interface; we could implement a completely strict map,
+or we could lazily evaluate the values stored as needed.
+\footnote{A third option could be to have lazy keys, but they would
+necessarily be forced at every insertion or lookup, with no gain.}
+Haskell's standard library provides these two variants in the form of
+\lk{Data.Map.Strict} and \lk{Data.Map.Lazy}. However, there is no way
+to {\em specify} the difference between the two versions! At best, one
+could hope for a hand-crafted suite of unit tests containing bottoms
+to observe a difference in termination.
+\leo{One thing we could do is ignore the knapsack and expand on the above.}
 
-The inputs of the knapsack problem are two sequences of numbers $w_i$
-and $v_i$ where $i = 1, \dots, n$, and each of $w_i$ and $v_i$ specify
-the weight and the value of the $i$th item. The solution presents the
+Moreover, it is important to be able to distinguish whether such a
+datatype follows to a desired laziness behavior. For instance, the
+standard functional solution for dynamic programming problems
+crucially uses laziness. In particular, consider the famous knapsack
+problem.
+%
+The inputs of the problem are two sequences of numbers $w_i$ and $v_i$
+where $i = 1, \dots, n$, and each of $w_i$ and $v_i$ specify the
+weight and the value of the $i$th item. The solution presents the
 maximum value achievable using a subset of the items such that their
@@
 
+\leo{Is this true? Where would the infinite loop be? there is a bottom to the recursion.
+I thought the problem was that it computes everything when it doesn't need to}
 Using the lazy variant of \lk{Map} here is critical for the correctness

### 2018-02-13 paper/main.tex

@@
 combined weight is under a limit $W$.
-
+%
 The usual dynamic programming approach for knapsack maintains a
@@
 %
-However, lazy evaluation automatically finds this path if we
-use \lk{Data.Map.Lazy} and setup the solution table with the
-recurrence relation \hzh{We can probably make this code prettier}
+However, lazy evaluation automatically finds this path if we use lazy
+maps to represent the solution table according to the recurrence
+relation \hzh{We can probably make this code prettier}
 \begin{inlinecode}
@@
 
+To query the optimal achievable values from the solution table, we can
+just index into the map. When an entry from the map is looked up at
+the first time, a sequence of evaluations is triggered, and their
+results memoized in the lazy map automatically.
+
 \leo{Is this true? Where would the infinite loop be? there is a bottom to the recursion.
@@
 Using the lazy variant of \lk{Map} here is critical for the correctness
-of \lk{solution}. If we used \lk{Data.Map.Strict} instead, then \lk{fromList}
-would have eagerly evaluated the \lk{if} expression. The \lk{if} expression
-contains value-level recursion to itself, and this would manifest as an infinite
-loop at runtime. This is an incredibly subtle bug related to laziness.
+of \lk{solution}.
+However, using the lazy variant of \lk{Map} here is critical for the
+correctness of \lk{solution}. If we used a map strict in its values
+instead, then \lk{fromList} would have eagerly evaluated the \lk{if}
+expression. The \lk{if} expression contains value-level recursion to
+the solution map itself, and this would manifest as an infinite loop
+at runtime. This is an incredibly subtle bug related to laziness.
 
@@
 or it can completely break down! Reasoning about lazy programs is also
-notoriously hard because predicting when a lazy value gets evaluated
-is difficult.
+notoriously hard --- the programmer needs to have a mental model of the
+order of evaluation of the code, and this order is not always obvious
+from the syntax of the code.
 
@@
 \label{sec:quickchick}
+
+In this section, we will discuss how to specify the laziness behaviors
+of Haskell functions through running examples of purely functional
+maps from integers to integers. The map operations of interest
+are \lk{insert} and \lk{lookup} with the following types:
+\begin{inlinecode}
+insert :: IntMap -> Int -> Int -> IntMap Int
+lookup :: IntMap -> Int        -> Maybe  Int
+\end{inlinecode}
+%
+The \lk{IntMap} can be implemented with a few different laziness
+flavors. We can choose a binary search tree representation for
+efficient lookups, or we can choose an association list representation
+for easy inserts. The binary search tree representation necessarily
+evaluates the keys of the map to order the keys, but it can be lazy in
+the values. On the other hand, the association list representation can
+be lazy in the keys as well. To summarise, we have maps that are
+strict in both keys and values, maps that are strict in keys but lazy
+in values, and maps that are fully lazy in both keys and values.
+
+In Haskell, triggering the evaluation of a lazy function requires
+exerting some demand on its result. Such demands can be pattern
+matches, or further computations that depend on the result. In typical
+Haskell programming, such demands are exerted by the contextual code
+surrounding the result of a lazy function. We will use the
+phrase \textit{demand context} to refer to functions that causes
+evaluation of lazy values.
+
+An evaluation of a lazy data structure can partially explore the lazy
+structure's shape. For example, \lk{lookup} on an association list
+based map evaluates the association list until it finds the first
+matching key. If such a key is halfway inside the association list,
+then the values past this key will not be evaluated in this operation.
+%
+To describe how much of a data structure is evaluated, we will
+use \textit{demand types}. A \textit{demand type} is almost isomorphic
+to the corresponding data type, except we will wrap all the
+constructor arguments in a \lk{Thunk} to represent the possible lack
+of evaluation. For example, consider the \lk{Tree} type and
+its \textit{demand type} \lk{TreeD}.
 

### 2018-02-14 paper/main.tex
[paper] Restructure after today's discussion

@@
-% !!! PLEASE DON'T CHANGE THESE !!! INSTEAD DEFINE YOUR OWN texdirectives.tex !!!
+% !!! Please Don't CHANGE THESE !!! INSTEAD DEFINE YOUR OWN texdirectives.tex !!!
 % These flags are for the final camera ready version for POPL
@@
 enjoy the power of lazy evaluation by default. However, that power
-does not come without a price: reasoning about lazy programs is
-notoriously hard. Let us consider two examples that demonstrate the
-power of laziness, as well as the intricacies of programming with it.
+does not come without a price.
+\leo{Didn't we have an FRP quote by someone that Laziness was the third source of
+bugs or something? Work that in?}
+\hzh{Ryan Trinkle (author of Reflex) said laziness was the 3rd most difficult thing to
+get right, but let's check with him on whether we can quote him on that.}
+
+The power of laziness puts great responsibility on programmers'
+shoulders. An incorrect implementation can either behave functionally
+the same but with inferior performance, or it can completely break
+down! Reasoning about lazy programs is also notoriously hard --- the
+programmer needs to have a mental model of the order of evaluation of
+the code, and this order is not always obvious from the syntax of the
+code.
+
+How, then, can we test and catch these laziness bugs during
+development? Is it possible to precisely specify laziness behavior
+just as we would for functional correctness?
+%
+In this paper, we demonstrate techniques for observing laziness in
+Haskell, and propose a system of creating specifications of lazy
+Haskell programs. We implement these ideas as a Haskell library. For
+concreteness, consider the following small example.
+
+\leo{Insert take example here. There is enough space to have both code
+and contributions in this page. yay :) }
+
+In the next section we work through two more examples, functional
+queues and functional maps, demonstrating both the difficulties of
+programming with laziness and how our approach can be used to
+alleviate them.
+%
+In the rest of the paper, we present our technical contributions:
+\begin{itemize}
+\item We introduce a novel approach to specifying the laziness behavior
+of functions that is {\em precise} and {\em compositional}. Such
+specifications are expressive enough to capture exactly how much
+evaluation occurs within a data structure
+%\leo{Also add
+%drawbacks/partiality here? If not here, where in the intro?} \hzh{I'm
+%not sure if we should bring in partiality (in the sense of
+%divergence/nontermination) yet since that seems like a rather
+%technical detail of the spec}
+\item We propose a framework for testing against such specifications.
+\item We improve upon the random function generators from QuickCheck to generate functions with arbitrary laziness behaviors.
+\item We demonstrate non-trivial generic programming techniques to apply StrictCheck to almost all data types.
+\end{itemize}
+%
+Section~\ref{sec:related} discusses related work. We conclude and draw
+directions for future work in Section~\ref{sec:concl}.
+
+\leo{These examples should go in the next section. Putting them here to avoid conflicts with Hengchu}
 
@@
 at runtime. This is an incredibly subtle bug related to laziness.
-
-As these examples show, the power of laziness puts great
-responsibility on programmers' shoulders. An incorrect implementation
-can either behave functionally the same but with inferior performance,
-or it can completely break down! Reasoning about lazy programs is also
-notoriously hard --- the programmer needs to have a mental model of the
-order of evaluation of the code, and this order is not always obvious
-from the syntax of the code.
-
-\leo{Didn't we have an FRP quote by someone that Laziness was the third source of
-bugs or something? Work that in?}
-\hzh{Ryan Trinkle (author of Reflex) said laziness was the 3rd most difficult thing to
-get right, but let's check with him on whether we can quote him on that.}
-
-How, then, can we test and catch these laziness bugs during development? Is it
-possible to precisely specify laziness behavior just as we would for functional
-correctness?
-%
-In this paper, we demonstrate techniques for observing laziness in Haskell, and
-propose a system of creating specifications of lazy Haskell programs. We
-implement these ideas as a Haskell library.
-
-Our technical contributions are as follows:
-\begin{itemize}
-\item We introduce a novel approach to specifying the laziness behavior of functions that is
-{\em precise} \hzh{and compositional}. Such specifications are
-expressive enough to capture exactly how much evaluation occurs within
-a data structure \leo{Also add drawbacks/partiality here? If not here,
-where in the intro?} \hzh{I'm not sure if we should bring in
-partiality (in the sense of divergence/nontermination) yet since that
-seems like a rather technical detail of the spec}
-\item We propose a framework for testing against such specifications.
-\item We improve upon the random function generators from QuickCheck to generate functions with arbitrary laziness behaviors.
-\item We demonstrate non-trivial generic programming techniques to apply StrictCheck to almost all data types.
-\end{itemize}
-%
-Section~\ref{sec:related} discusses related work. We conclude and draw
-directions for future work in Section~\ref{sec:concl}.
 

### 2018-02-19 paper/main.tex
[paper] work in progress for section 2

@@
 \usepackage{relsize}
+\usepackage{float}
+\usepackage{cleveref}
 %\usepackage{mathpartir}
@@
 of Haskell functions through running examples of purely functional
-maps from integers to integers. The map operations of interest
-are \lk{insert} and \lk{lookup} with the following types:
+maps. The map operations of interest are \lk{insert} and \lk{lookup}
+with the following types:
 \begin{inlinecode}
@@
 for easy inserts. The binary search tree representation necessarily
-evaluates the keys of the map to order the keys, but it can be lazy in
-the values. On the other hand, the association list representation can
-be lazy in the keys as well. To summarise, we have maps that are
-strict in both keys and values, maps that are strict in keys but lazy
-in values, and maps that are fully lazy in both keys and values.
+evaluates the keys of the map to order the keys, but it has flexible
+laziness behavior on the values of the map. On the other hand, the
+association list representation is even more flexible since it can be
+lazy in the keys as well. To summarise, we have maps that are strict
+in both keys and values, maps that are strict in keys but lazy or
+strict in values, and maps that are fully lazy in both keys and
+values.
 
@@
 matching key. If such a key is halfway inside the association list,
-then the values past this key will not be evaluated in this operation.
+then the map entries past this key will not be evaluated.
 %
@@
 constructor arguments in a \lk{Thunk} to represent the possible lack
-of evaluation. For example, consider the \lk{Tree} type and
-its \textit{demand type} \lk{TreeD}.
+of evaluation. For example, let the binary tree representation of maps
+be the \lk{Tree} type and let us consider its \textit{demand
+type} \lk{TreeD}.
+\begin{inlinecode}
+data Tree =              data TreeD =
+  | Leaf                   | LeafD
+  | Node Tree Int Tree     | NodeD (Thunk TreeD) (Thunk IntD) (Thunk TreeD)
+\end{inlinecode}
+%
+Values of \lk{TreeD} type describe a shape of values in the \lk{Tree}
+type, and values of \lk{Thunk} describe any prefix shape of a given
+tree since it has the following definition:
+\begin{inlinecode}
+data Thunk a = E !a | T
+\end{inlinecode}
+%
+The \lk{E} constructor stands for \textit{evaluated} and the \lk{T}
+constructor stand for \textit{thunk}. In \Cref{fig:tree-treeD}, the
+value \lk{treeShape} matches an evaluation that only demands the root
+of \lk{tree} and the value in its left branch.
+
+\begin{figure}[H]
+\begin{minipage}{0.4\textwidth}
+\begin{inlinecode}
+tree = Node
+         Node Leaf 1 Leaf
+         2
+         Node Leaf 3 Leaf
+\end{inlinecode}
+\end{minipage}
+\begin{minipage}{0.4\textwidth}
+\begin{inlinecode}
+treeShape = NodeD
+              E (NodeD T (E 1) T)
+              E 2
+              T
+\end{inlinecode}
+\end{minipage}
+\caption{Example of tree shape}
+\label{fig:tree-treeD}
+\end{figure}
 

### 2018-02-20 paper/main.tex
[paper] backup

@@
 \label{sec:quickchick}
-
+\iffalse
 In this section, we will discuss how to specify the laziness behaviors
@@
 constructor arguments in a \lk{Thunk} to represent the possible lack
-of evaluation. For example, let the binary tree representation of maps
-be the \lk{Tree} type and let us consider its \textit{demand
-type} \lk{TreeD}.
+of evaluation.
+%
+Concretely, here are the 3 types of maps of interest
+in \Cref{fig:map-types}.
+%
+\begin{figure}[H]
+\centering
 \begin{inlinecode}
-data Tree =              data TreeD =
-  | Leaf                   | LeafD
-  | Node Tree Int Tree     | NodeD (Thunk TreeD) (Thunk IntD) (Thunk TreeD)
+data LazyMap   = Leaf
+               | Node LazyMap Key Value LazyMap
+
+data StrictMap = SLeaf
+               | SNode StrictMap Key !Value StrictMap
+
+data ListMap   = Nil
+               | Cons Key Value ListMap
 \end{inlinecode}
+\caption{3 types of maps with different laziness}
+\label{fig:map-types}
+\end{figure}
 %
-Values of \lk{TreeD} type describe a shape of values in the \lk{Tree}
-type, and values of \lk{Thunk} describe any prefix shape of a given
-tree since it has the following definition:
+The \textit{demand types} of the 3 kinds of maps and the \lk{Thunk}
+type are defined in \Cref{fig:map-demand-types}. We will use the
+Haskell list-of-pairs syntax when referring to maps of
+type \lk{ListMap}.
+%
+\begin{figure}[H]
+\centering
 \begin{inlinecode}
-data Thunk a = E !a | T
+data Thunk a    = E !a
+                | T
+
+data LazyMapD   = LeafD
+                | NodeD (Thunk LazyMapD)
+                        (Thunk KeyD)
+                        (Thunk ValueD)
+                        (Thunk LazyMapD)
+
+data StrictMapD = SLeafD
+                | SNodeD (Thunk StrictMapD)
+                         (Thunk KeyD)
+                         ValueD
+                         (Thunk StrictMapD)
+
+data ListMapD   = NilD
+                | ConsD (Thunk KeyD)
+                        (Thunk ValueD)
+                        (Thunk ListMapD)
 \end{inlinecode}
+\caption{Demand types}
+\label{fig:map-demand-types}
+\end{figure}
 %
-The \lk{E} constructor stands for \textit{evaluated} and the \lk{T}
-constructor stand for \textit{thunk}. In \Cref{fig:tree-treeD}, the
-value \lk{treeShape} matches an evaluation that only demands the root
-of \lk{tree} and the value in its left branch.
-
-\begin{figure}[H]
-\begin{minipage}{0.4\textwidth}
-\begin{inlinecode}
-tree = Node
-         Node Leaf 1 Leaf
-         2
-         Node Leaf 3 Leaf
-\end{inlinecode}
-\end{minipage}
-\begin{minipage}{0.4\textwidth}
-\begin{inlinecode}
-treeShape = NodeD
-              E (NodeD T (E 1) T)
-              E 2
-              T
-\end{inlinecode}
-\end{minipage}
-\caption{Example of tree shape}
-\label{fig:tree-treeD}
-\end{figure}
+The \lk{E} constructor stands for \textit{evaluated}, and the \lk{T}
+constructor stands for \textit{thunk}. Using the 3 \textit{demand
+types}, we can describe the shape of a demand exerted onto the
+maps. Take the example list map \lk{[(1, 1), (2, 2), (3, 3)]}.
+\fi
 

### 2018-02-20 paper/main.tex
[paper] backup, finished spec for rotate but the text around it still needs much work

--- 2018-02-14 paper/main.tex
@@
 
-\leo{The singular feels weird here. Is it just me?}
-As a first example, consider purely functional queues, a very
-popular data structure in Haskell. Programmers typically implement
-such queues with two lists---a front list holding the queue elements
-in FIFO order, and a back list holding the other elments in reversed
-order. When the front list is empty, a dequeue operation requires a
-reversal of the back list to refill the front list. These specific
-dequeue operations take $O(n)$ time to finish, and are much slower
-than their imperative counterparts. This occasional performance
-degradation is compounded by persistence (where the $O(n)$ operation
-could be repeated many times), and can be unacceptable in real-time
-systems.
-
-However, Okasaki cleverly improved upon this design using lazy lists
-that both incrementally performs the reversal of the back list, and
-incrementally appends the reversal of the back list onto the front
-list ~\cite{Okasaki95simpleand}. Okasaki's queue relies on the
-following \lk{rotate} function to incrementally reverse and append the
-back list.
-\begin{inlinecode}
-rotate :: [a] -> [a] -> [a] -> [a]
-rotate      []       []  as =                       as
-rotate      []  (b : bs) as =     rotate [] bs (b : as)
-rotate (f : fs)      []  as = f : rotate fs []      as
-rotate (f : fs) (b : bs) as = f : rotate fs bs (b : as)
-\end{inlinecode}
-If we computed the expression \lk{rotate front back []}, its value
-will be equivalent to \lk{front ++ (reverse back)}. However, the
-function \lk{rotate} is much lazier than the naive reverse and append
-operation. The function \lk{reverse} has to evaluate the entire
-structure of the back list, which is not incremental. On the other
-hand, the function \lk{rotate} produces a list immediately, and packs
-the rest of the computation into the lazy list \lk{rest}.
-
-Since \lk{rotate} is functionally equivalent to \lk{front ++ (reverse
-back)}, a programmer could have implemented it naively and never
-discover the mistake through traditional testing! The two only differ
-in their laziness behavior, and there is no way of specifying such
-subtle differences in any testing frameworks today.
-
 \leo{Example 2: Maps / Knapsack}
@@
 
-In this section, we will discuss how to specify the laziness behaviors
-of Haskell functions through running examples of purely functional
-maps from integers to integers. The map operations of interest
-are \lk{insert} and \lk{lookup} with the following types:
+As a first example, consider purely functional queues, a very
+popular data structure in Haskell. Programmers typically implement
+such queues with two lists---a front list holding the queue elements
+in FIFO order, and a back list holding the other elments in reversed
+order. When the front list is empty, a dequeue operation requires a
+reversal of the back list to refill the front list. These specific
+dequeue operations take $O(n)$ time to finish, and are much slower
+than their imperative counterparts. This occasional performance
+degradation is compounded by persistence where the $O(n)$ operation
+could be repeated many times, and can be unacceptable in real-time
+systems.
+
+However, Okasaki cleverly improved upon this design using lazy lists
+that both incrementally performs the reversal of the back list, and
+incrementally appends the reversal of the back list onto the front
+list ~\cite{Okasaki95simpleand}. Okasaki's queue relies on the
+following \lk{rotate} function to incrementally reverse and append the
+back list.
 \begin{inlinecode}
-insert :: IntMap -> Int -> Int -> IntMap Int
-lookup :: IntMap -> Int        -> Maybe  Int
+rotate :: [a] -> [a] -> [a] -> [a]
+rotate      []       []  as =                       as
+rotate      []  (b : bs) as =     rotate [] bs (b : as)
+rotate (f : fs)      []  as = f : rotate fs []      as
+rotate (f : fs) (b : bs) as = f : rotate fs bs (b : as)
 \end{inlinecode}
+If we computed the expression \lk{rotate front back []}, its value
+will be equivalent to \lk{front ++ (reverse back)}. However, the
+function \lk{rotate} is much lazier than the naive reverse and append
+operation. The function \lk{reverse} has to evaluate the entire
+structure of the back list, which is not incremental. On the other
+hand, the function \lk{rotate} produces a list immediately, and packs
+the rest of the computation into the lazy list \lk{rest}.
+
+Since \lk{rotate} is functionally equivalent to \lk{front ++ (reverse
+back)}, a programmer could have implemented it naively and never
+discover the mistake through traditional testing! The two only differ
+in their laziness behavior. Fortunately, we can use StrictCheck to
+write a specification that makes this property of \lk{rotate}
+testable.
+
+To describe the laziness behavior, we need to first introduce a few
+types that describe how lists are used. We use the
+term \textit{demand} to refer to the shape of such usage. Concretely,
+a \textit{demand} is a shape that describes the used part of a data
+structure. For lists, \textit{demands} will be prefixes of lists. We
+use the type \lk{ListD} to represent such demands (\Cref{fig:ListD}).
 %
-The \lk{IntMap} can be implemented with a few different laziness
-flavors. We can choose a binary search tree representation for
-efficient lookups, or we can choose an association list representation
-for easy inserts. The binary search tree representation necessarily
-evaluates the keys of the map to order the keys, but it can be lazy in
-the values. On the other hand, the association list representation can
-be lazy in the keys as well. To summarise, we have maps that are
-strict in both keys and values, maps that are strict in keys but lazy
-in values, and maps that are fully lazy in both keys and values.
+\begin{figure}[H]
+\centering
+\begin{inlinecode}
+data ListD a = NilD
+             | ConsD (Thunk a) (Thunk (ListD a))
+\end{inlinecode}
+\caption{\lk{ListD} type}
+\label{fig:ListD}
+\end{figure}
+%
+The \lk{Thunk} type used in \lk{ListD} has 2 constructors: \lk{E}
+for \textit{evaluated} and a nullary constructor \lk{T}
+for \textit{thunk}. In \lk{ListD}, \lk{Thunk} wraps around all the
+list constructor fields, and this allows us to represent lack of
+further evaluation by using the \lk{T} constructor
+from \lk{Thunk}. For example, given the list \lk{[1, 2, 3]}, the
+value \lk{ConsD T (ConsD T T)} represents the \textit{demand} that
+reveals the first 2 \lk{Cons} constructors of the list, but does not
+evaluate any other parts of the list. To improve readability of demand
+values, we will also use the syntax \lk{[\_, \_, ...]} where
+appropriate.
 
-In Haskell, triggering the evaluation of a lazy function requires
-exerting some demand on its result. Such demands can be pattern
-matches, or further computations that depend on the result. In typical
-Haskell programming, such demands are exerted by the contextual code
-surrounding the result of a lazy function. We will use the
-phrase \textit{demand context} to refer to functions that causes
-evaluation of lazy values.
+In Haskell, to trigger the evaluation of a lazy function, we need to
+exert some \textit{demand} on the result of that function. This
+suggests a natural flow for the specification of a lazy function ---
+given the demand on the output, the specification should calculate the
+demands on the input. However, this is not enough in general because
+the demand on the input could depend on the value of the inputs as
+well. The \lk{take} function introduced in \Cref{sec:intro} is such a
+case since the length of the output from \lk{take} depends on its
+integer argument.
 
-An evaluation of a lazy data structure can partially explore the lazy
-structure's shape. For example, \lk{lookup} on an association list
-based map evaluates the association list until it finds the first
-matching key. If such a key is halfway inside the association list,
-then the values past this key will not be evaluated in this operation.
+Putting these pieces together, we arrive at an appropriate type for
+the specification of \lk{rotate}: \lk{ListD a -> [a] -> [a] -> [a] ->
+(ListD a, ListD a, ListD a)}.
 %
-To describe how much of a data structure is evaluated, we will
-use \textit{demand types}. A \textit{demand type} is almost isomorphic
-to the corresponding data type, except we will wrap all the
-constructor arguments in a \lk{Thunk} to represent the possible lack
-of evaluation. For example, consider the \lk{Tree} type and
-its \textit{demand type} \lk{TreeD}.
+\hzh{The description here needs improvement}
+To specify the laziness behavior of \lk{rotate}, we need to inspect
+the 4 different cases of \lk{rotate}. When both \lk{fs} and \lk{bs}
+are empty, the accumulator is returned directly; when \lk{fs} is an
+empty list, \lk{rotate} immediately recurses on the tail of \lk{bs};
+when \lk{bs} is empty, \lk{rotate} produces a new list with the
+original head of \lk{fs} as the new head, and the tail of that list is
+the result of a recursion; finally, when both \lk{fs} and \lk{bs} are
+not empty, \lk{rotate} also immediately produces a new list with the
+appropriate head, and conses \hzh{is this a word} the original head
+of \lk{bs} onto the accumulator.
+
+Turning our intuition above into a Haskell function, we complete the
+specification of \lk{rotate} in \Cref{fig:rotate-spec}
+\begin{figure}[H]
+\centering
+\begin{inlinecode}
+rotate_spec :: ListD a
+            -> [a] -> [a] -> [a]
+            -> (Thunk (ListD a), Thunk (ListD a), Thunk (ListD a))
+rotate_spec NilD                     fs bs as = (E NilD, E NilD, E NilD)
+rotate_spec dOut@(ConsD dElem dTail) fs bs as =
+  case (fs, bs) of
+    ([],           []) ->
+      (E NilD, E NilD, E dOut)
+    ([],      b : bs') ->
+      let (_, dBs', dBAs) = rotate_spec dTail [] bs' (b:as)
+      in case dBAs of
+           T                -> (E NilD, T, T)
+           E (ConsD dB dAs) -> (E NilD, ConsD dB dBs', dAs)
+           E NilD           -> error ``Impossible''
+    (_ : fs',      []) ->
+      let (dFs', _, dAs) = rotate_spec dTail fs' [] as
+      in  (E (ConsD dElem dFs'), E NilD, dAs)
+    (_ : fs', b : bs') ->
+      let (dFs', dBs', dBAs) = rotate_spec dTail as' bs' (b:as)
+          dFs                = E (ConsD dElem dFs')
+      in case dBAs of
+           T                -> (dFs, T, T)
+           E (ConsD dB dAs) -> (dFs, E (ConsD dB dBs'), dAs)
+           E NilD           -> error ``Impossible''
+\end{inlinecode}
+\caption{Specification for \lk{rotate}}
+\label{fig:rotate-spec}
+\end{figure}
 

### 2018-02-25 paper/main.tex
[paper] more work on section 2

@@
 
-\leo{These examples should go in the next section. Putting them here to avoid conflicts with Hengchu}
-
-\leo{Example 1: Okasaki queues}
-
-\leo{Example 2: Maps / Knapsack}
-
-For the second example, we turn to a different data functional data
-structure where both a strict and a lazy variant can be situationally
-useful: maps. There are two reasonable laziness behaviors that
-implement a map interface; we could implement a completely strict map,
-or we could lazily evaluate the values stored as needed.
-\footnote{A third option could be to have lazy keys, but they would
-necessarily be forced at every insertion or lookup, with no gain.}
-Haskell's standard library provides these two variants in the form of
-\lk{Data.Map.Strict} and \lk{Data.Map.Lazy}. However, there is no way
-to {\em specify} the difference between the two versions! At best, one
-could hope for a hand-crafted suite of unit tests containing bottoms
-to observe a difference in termination.
-\leo{One thing we could do is ignore the knapsack and expand on the above.}
-
-Moreover, it is important to be able to distinguish whether such a
-datatype follows to a desired laziness behavior. For instance, the
-standard functional solution for dynamic programming problems
-crucially uses laziness. In particular, consider the famous knapsack
-problem.
-%
-The inputs of the problem are two sequences of numbers $w_i$ and $v_i$
-where $i = 1, \dots, n$, and each of $w_i$ and $v_i$ specify the
-weight and the value of the $i$th item. The solution presents the
-maximum value achievable using a subset of the items such that their
-combined weight is under a limit $W$.
-%
-The usual dynamic programming approach for knapsack maintains a
-table $OPT[][]$, where the entry $OPT[j][k]$ keeps the best value
-achievable using only items $1\dots j$ and a weight limit of $k$. The
-table satisfies the following recurrence relation: $$OPT[j][k]
-= \max(OPT[j-1][w], w_j + OPT[j-1][k-w_j])$$
-
-To efficiently compute this table using imperative langauges, we look
-for shared sub-problems of the original problem, and design a path
-along the solution table that maximizes sharing while filling out the
-answers.
-%
-However, lazy evaluation automatically finds this path if we use lazy
-maps to represent the solution table according to the recurrence
-relation \hzh{We can probably make this code prettier}
-\begin{inlinecode}
-ws, vs :: [Int]
-n,  w  :: Int
-
-solution :: Map (Int, Int) Int
-solution = fromList [ ((j, k),
-                       if (j-1) < 0 || k - ws !! j < 0
-                       then 0
-                       else max (solution ! (j-1, k))
-                                (solution ! (j-1, k - ws !! j) + vs !! j)
-                      )
-                    | j <- [0..n-1], k <- [0..w] ]
-\end{inlinecode}
-
-To query the optimal achievable values from the solution table, we can
-just index into the map. When an entry from the map is looked up at
-the first time, a sequence of evaluations is triggered, and their
-results memoized in the lazy map automatically.
-
-\leo{Is this true? Where would the infinite loop be? there is a bottom to the recursion.
-I thought the problem was that it computes everything when it doesn't need to}
-Using the lazy variant of \lk{Map} here is critical for the correctness
-of \lk{solution}.
-However, using the lazy variant of \lk{Map} here is critical for the
-correctness of \lk{solution}. If we used a map strict in its values
-instead, then \lk{fromList} would have eagerly evaluated the \lk{if}
-expression. The \lk{if} expression contains value-level recursion to
-the solution map itself, and this would manifest as an infinite loop
-at runtime. This is an incredibly subtle bug related to laziness.
-
 \section{Strictness Specifications, by Example}
@@
 evaluate any other parts of the list. To improve readability of demand
-values, we will also use the syntax \lk{[\_, \_, ...]} where
+values, we will also use the syntax \lk{\_ : \_ : \_} where
 appropriate.
@@
 \hzh{The description here needs improvement}
-To specify the laziness behavior of \lk{rotate}, we need to inspect
-the 4 different cases of \lk{rotate}. When both \lk{fs} and \lk{bs}
-are empty, the accumulator is returned directly; when \lk{fs} is an
-empty list, \lk{rotate} immediately recurses on the tail of \lk{bs};
-when \lk{bs} is empty, \lk{rotate} produces a new list with the
-original head of \lk{fs} as the new head, and the tail of that list is
-the result of a recursion; finally, when both \lk{fs} and \lk{bs} are
-not empty, \lk{rotate} also immediately produces a new list with the
-appropriate head, and conses \hzh{is this a word} the original head
-of \lk{bs} onto the accumulator.
+To specify the laziness behavior of \lk{rotate}, we return to Okasaki's insight:
+as long as the front list is longer than the back list, demanding the first
+element from the result of \lk{rotate} only causes evaluation of the first
+element of the front list ~\cite{Okasaki95simpleand}. Using this precondition
+for \lk{rotate}, we arrive at the following executable
+specification \lk{rotate\_spec}.
 
@@
 \centering
+\hzh{TODO}
 \begin{inlinecode}
-rotate_spec :: ListD a
-            -> [a] -> [a] -> [a]
-            -> (Thunk (ListD a), Thunk (ListD a), Thunk (ListD a))
-rotate_spec NilD                     fs bs as = (E NilD, E NilD, E NilD)
-rotate_spec dOut@(ConsD dElem dTail) fs bs as =
-  case (fs, bs) of
-    ([],           []) ->
-      (E NilD, E NilD, E dOut)
-    ([],      b : bs') ->
-      let (_, dBs', dBAs) = rotate_spec dTail [] bs' (b:as)
-      in case dBAs of
-           T                -> (E NilD, T, T)
-           E (ConsD dB dAs) -> (E NilD, ConsD dB dBs', dAs)
-           E NilD           -> error ``Impossible''
-    (_ : fs',      []) ->
-      let (dFs', _, dAs) = rotate_spec dTail fs' [] as
-      in  (E (ConsD dElem dFs'), E NilD, dAs)
-    (_ : fs', b : bs') ->
-      let (dFs', dBs', dBAs) = rotate_spec dTail as' bs' (b:as)
-          dFs                = E (ConsD dElem dFs')
-      in case dBAs of
-           T                -> (dFs, T, T)
-           E (ConsD dB dAs) -> (dFs, E (ConsD dB dBs'), dAs)
-           E NilD           -> error ``Impossible''
+TODO: rewrite this spec so it looks like the invariant in Okasaki's paper.
 \end{inlinecode}
@@
 \label{fig:rotate-spec}
+\end{figure}
+
+\lk{rotate\_spec} exactly captures the core idea from Okasaki's \lk{rotate}
+operation, and testing the reverse and append implementation against this
+specification immediately produces a counter example. \hzh{should we show what a
+counter example looks like?}
+
+\leo{Example 2: Maps / Knapsack}
+
+For the second example, we turn to a different functional data structure where
+both a strict and a lazy variant can be situationally useful: maps. There are
+two reasonable laziness behaviors that implement a map interface; we could
+implement a completely strict map, or we could lazily evaluate the values stored
+as needed.
+\footnote{A third option could be to have lazy keys, but they would
+necessarily be forced at every insertion or lookup, with no gain.}
+Haskell's standard library provides these two variants in the form of
+\lk{Data.Map.Strict} and \lk{Data.Map.Lazy}. However, there is no way
+to {\em specify} the difference between the two versions! At best, one
+could hope for a hand-crafted suite of unit tests containing bottoms
+to observe a difference in termination.
+\leo{One thing we could do is ignore the knapsack and expand on the above.}
+
+Moreover, it is important to be able to distinguish whether such a datatype
+follows a desired laziness behavior. For instance, the standard functional
+solution for dynamic programming problems crucially uses laziness \hzh{is this
+solution really standard? Is there some work we can cite to establish this
+approach as the canonical approach for dynamic programming in functional
+programming?}. In particular, consider the famous knapsack problem.
+%
+The inputs of the problem are two sequences of numbers $w_i$ and $v_i$
+where $i = 1, \dots, n$, and each of $w_i$ and $v_i$ specify the
+weight and the value of the $i$th item. The solution presents the
+maximum value achievable using a subset of the items such that their
+combined weight is under a limit $W$.
+%
+The usual dynamic programming approach for knapsack maintains a
+table $OPT[][]$, where the entry $OPT[j][k]$ keeps the best value
+achievable using only items $1\dots j$ and a weight limit of $k$. The
+table satisfies the following recurrence relation: $$OPT[j][k]
+= \max(OPT[j-1][w], w_j + OPT[j-1][k-w_j])$$
+
+To efficiently compute this table using imperative langauges, we look
+for shared sub-problems of the original problem, and design a path
+along the solution table that maximizes sharing while filling out the
+answers.
+%
+However, lazy evaluation automatically finds this path if we use lazy
+maps to represent the solution table according to the recurrence
+relation \hzh{We can probably make this code prettier}
+\begin{inlinecode}
+ws, vs :: [Int]
+n,  w  :: Int
+
+solutionPre :: Map (Int, Int) Int -> Map (Int, Int) Int
+solutionPre soln = fromList [ ((j, k),
+                               if (j-1) < 0 || k - ws !! j < 0
+                               then if ws !! j <= k then vs !! j else 0
+                               else max (soln ! (j-1, k))
+                                        (soln ! (j-1, k - ws !! j) + vs !! j)
+                              )
+                            | j <- [0..2], k <- [0..50] ]
+
+solution :: Map (Int, Int) Int
+solution = fix solutionPre
+\end{inlinecode}
+
+To query the optimal achievable values from the solution table, we can
+just index into the map. When an entry from the map is looked up at
+the first time, a sequence of evaluations is triggered, and their
+results memoized in the lazy map automatically.
+
+However, using the lazy variant of \lk{Map} here is critical for the
+correctness of \lk{solution}. If we used a map strict in its values
+instead, then \lk{fromList} would have eagerly evaluated the \lk{if}
+expression. The \lk{if} expression contains value-level recursion to
+the solution map itself, and this would manifest as an infinite loop
+at runtime. This is an incredibly subtle bug related to laziness.
+
+To catch this behavior through testing, we can write another specification that
+captures this correctness condition. The well-foundedness \hzh{is this a word?}
+of \lk{solution} depends on what is evaluated when the value at key \lk{(j, k)}
+of the map is demanded. Let $\lk{solutionPre}_n$ be the $n$-times iterated
+pre-fixpoint \lk{solutionPre . solutionPre . ... . solutionPre}, then if we
+demand the value at \lk{(j, k)} from the result of $\lk{solutionPre}_n$, its
+input map should not be evaluated at the \lk{(j, k)} entry since otherwise
+evaluating the value at \lk{(j, k)} becomes an infinite loop at the fixpoint.
+
+We can write another executable specification for this property
+of $\lk{solutionPre}_n$ (\Cref{fig:spec-map}). Testing the implementation based on
+strict maps also immediately produces a counter example.
+
+\begin{figure}[H]
+\centering
+\hzh{TODO: specification for map}
+\begin{inlinecode}
+\end{inlinecode}
+\caption{Specification for \lk{solutionPre}}
+\label{fig:spec-map}
 \end{figure}
@@
 \item Maps are particularly well-suited for a discussion of laziness because of the interesting different
-possible behaviors: strict-maps, value-lazy maps \leo{Point out Data.Map.Strict vs Data.Map}, and fully lazy maps
+possible behaviors: strict-maps, value-lazy maps \leo{Point out Data.Map.Strict
+vs Data.Map}, and fully lazy maps
 \item For simplicity, we will use binary search trees as maps, assuming the keys are integers and forego balancing

### 2018-02-25 paper/knapsack.hs

@@ -1,4 +1,5 @@
 import Data.Map.Lazy
+import Data.Function
 
 main = return ()
 
@@ -11,11 +12,14 @@ vs = [60, 100, 120]
 w :: Int
 w = 50
 
+solutionPre :: Map (Int, Int) Int -> Map (Int, Int) Int
+solutionPre soln = fromList [ ((j, k),
+                               if (j-1) < 0 || k - ws !! j < 0
+                               then if ws !! j <= k then vs !! j else 0
+                               else max (soln ! (j-1, k))
+                                        (soln ! (j-1, k - ws !! j) + vs !! j)
+                              )
+                            | j <- [0..2], k <- [0..50] ]
+
 solution :: Map (Int, Int) Int
-solution = fromList [ ((j, k),
-                       if (j-1) < 0 || k - ws !! j < 0
-                       then 0
-                       else max (solution ! (j-1, k))
-                                (solution ! (j-1, k - ws !! j) + vs !! j)
-                      )
-                    | j <- [0..2], k <- [0..50] ]
+solution = fix solutionPre

### 2018-02-27 paper/main.tex
[paper] Editing

@@
 
-As a first example, consider purely functional queues, a very
-popular data structure in Haskell. Programmers typically implement
-such queues with two lists---a front list holding the queue elements
-in FIFO order, and a back list holding the other elments in reversed
-order. When the front list is empty, a dequeue operation requires a
-reversal of the back list to refill the front list. These specific
-dequeue operations take $O(n)$ time to finish, and are much slower
-than their imperative counterparts. This occasional performance
-degradation is compounded by persistence where the $O(n)$ operation
-could be repeated many times, and can be unacceptable in real-time
-systems.
+For our first example, we turn to a very popular data structure:
+purely functional queues.  Such queues are typically implemented with
+two lists---a front list holding the queue elements in FIFO order, and
+a back list holding the other elments in reversed order. When the
+front list is empty, a dequeue operation must reverse the back list to
+refill the front list. These specific dequeue operations take $O(n)$
+time to finish and are much slower than their imperative
+counterparts, a performance degradation only compounded by
+persistence---where the $O(n)$ operation could be repeated many times,
+and unacceptable for many real-time systems.
 
-However, Okasaki cleverly improved upon this design using lazy lists
-that both incrementally performs the reversal of the back list, and
+Amongst his work on purely functional data structures\cite{Okasaki},
+Okasaki cleverly improved upon this design using lazy lists that both
+incrementally performs the reversal of the back list, and
 incrementally appends the reversal of the back list onto the front
@@
 If we computed the expression \lk{rotate front back []}, its value
-will be equivalent to \lk{front ++ (reverse back)}. However, the
-function \lk{rotate} is much lazier than the naive reverse and append
-operation. The function \lk{reverse} has to evaluate the entire
-structure of the back list, which is not incremental. On the other
-hand, the function \lk{rotate} produces a list immediately, and packs
-the rest of the computation into the lazy list \lk{rest}.
+will be equivalent to \lk{front ++ (reverse back)}. However, unlike
+the naive reverse-and-append operation, \lk{rotate} is much
+lazier. A \lk{reverse} function has to evaluate the entire structure
+of the back list, which is not incremental. On the other
+hand, \lk{rotate} produces a list immediately, and packs the rest of
+the computation into the lazy list \lk{rest}.
 
@@
 use the type \lk{ListD} to represent such demands (\Cref{fig:ListD}).
+\leo{Why is this a figure and not just inline? Especially if we're inlining
+the figure!}
 %
@@
 The \lk{Thunk} type used in \lk{ListD} has 2 constructors: \lk{E}
-for \textit{evaluated} and a nullary constructor \lk{T}
+for \textit{evaluated} and \lk{T}, a nullary constructor,
 for \textit{thunk}. In \lk{ListD}, \lk{Thunk} wraps around all the
 list constructor fields, and this allows us to represent lack of
-further evaluation by using the \lk{T} constructor
-from \lk{Thunk}. For example, given the list \lk{[1, 2, 3]}, the
-value \lk{ConsD T (ConsD T T)} represents the \textit{demand} that
-reveals the first 2 \lk{Cons} constructors of the list, but does not
-evaluate any other parts of the list. To improve readability of demand
-values, we will also use the syntax \lk{\_ : \_ : \_} where
-appropriate.
+further evaluation by using the \lk{T} constructor. For example,
+given the list \lk{[1, 2, 3]}, the value \lk{ConsD T (ConsD T T)}
+represents the \textit{demand} that reveals the first 2 \lk{Cons}
+constructors of the list, but does not evaluate any other parts of the
+list. To improve readability of demand values, we will also use the
+syntax \lk{\_ : \_ : \_} where appropriate.
 
@@
 specification immediately produces a counter example. \hzh{should we show what a
-counter example looks like?}
+counter example looks like?}\leo{Definitely!}
 

### 2018-03-03 paper/main.tex
[paper] fixed accent issue

@@
 \leo{Point of first paragraph: lazy evaluation is powerful}
+\caleb{What does the word ``power'' mean? It's non-specific. Useful? Convenient?}
 
@@
 
+\caleb{Again, power and responsibility are not completely clear, but I do like the narrative}
 The power of laziness puts great responsibility on programmers'
 shoulders. An incorrect implementation can either behave functionally
-the same but with inferior performance, or it can completely break
-down! Reasoning about lazy programs is also notoriously hard --- the
+the same but with inferior \caleb{Inferior than what?} performance, or it can completely break
+down!\caleb{unclear. loops?} Reasoning about lazy programs is also notoriously hard --- the
 programmer needs to have a mental model of the order of evaluation of
@@
 \leo{Insert take example here. There is enough space to have both code
-and contributions in this page. yay :) }
+and contributions in this page. yay :) }\caleb{YES!}
 
@@
 %technical detail of the spec}
-\item We propose a framework for testing against such specifications.
-\item We improve upon the random function generators from QuickCheck to generate functions with arbitrary laziness behaviors.
-\item We demonstrate non-trivial generic programming techniques to apply StrictCheck to almost all data types.
+\item We propose a framework for testing against such specifications.\caleb{Automated testing? Can be more specific}
+\item We improve upon the random function generators from QuickCheck\cn to generate functions with arbitrary laziness behaviors.
+\item We demonstrate non-trivial generic programming techniques to apply StrictCheck to almost all data types.\caleb{I don't know
+the meaning of non-trivial and almost-all}
 \end{itemize}
@@
 %
-José Manuel Calderón Trilla,
+Jos\'e Manuel Calder\'on Trilla,
 Matt Parsons,

### 2018-03-03 paper/main.tex

@@
 %
-Jos\'e Manuel Calder\'on Trilla,
+José Manuel Calderón Trilla,
 Matt Parsons,

### 2018-03-03 paper/main.tex
[paper] more work on section 2

--- 2018-03-03 paper/main.tex
@@
 
-Lazy evaluation gives tremendous power to functional programmers
+Lazy evaluation gives tremendous  to functional programmers
 ~\cite{WhyFPMatters} and is actively used in various practical
@@
 Haskell programs. We implement these ideas as a Haskell library. For
-concreteness, consider the following small example.
-
-\leo{Insert take example here. There is enough space to have both code
-and contributions in this page. yay :) }\caleb{YES!}
+concreteness, consider the following small example on the
+function \lk{take}, which accepts an integer $n$ and a list, and returns
+a prefix of the list with length $n$. If $n$ is larger than the length
+of the original list, then the entire list is returned.
+%
+\begin{inlinecode}
+take_spec :: Shaped a => Spec '[Int, [a]] [a]
+take_spec =
+  Spec $ \predict d n xs ->
+    predict (nf n) (if n > length xs then E d else endThunk d)
+  where
+    endThunk :: PosDemand [a] -> Demand [a]
+    endThunk = \case
+      Nil'        -> T
+      dx :% T     -> E (dx :% T)
+      dx :% E dxs -> E (dx :% endThunk dxs)
+\end{inlinecode}
+%
+\hzh{TODO: describe what this spec means in English}
 
@@
 %technical detail of the spec}
-\item We propose a framework for testing against such specifications.\caleb{Automated testing? Can be more specific}
+\item We propose a framework for automated testing against such specifications.\caleb{Automated testing? Can be more specific}
 \item We improve upon the random function generators from QuickCheck\cn to generate functions with arbitrary laziness behaviors.
-\item We demonstrate non-trivial generic programming techniques to apply StrictCheck to almost all data types.\caleb{I don't know
+\item We discuss our generic programming-based impementation to apply StrictCheck to all algebraic data types.\caleb{I don't know
 the meaning of non-trivial and almost-all}
@@
 structure. For lists, \textit{demands} will be prefixes of lists. We
-use the type \lk{ListD} to represent such demands (\Cref{fig:ListD}).
-\leo{Why is this a figure and not just inline? Especially if we're inlining
-the figure!}
+use the type \lk{ListD} to represent such demands.
 %
-\begin{figure}[H]
-\centering
 \begin{inlinecode}
@@
 \end{inlinecode}
-\caption{\lk{ListD} type}
-\label{fig:ListD}
-\end{figure}
 %
@@
 list constructor fields, and this allows us to represent lack of
-further evaluation by using the \lk{T} constructor. For example,
-given the list \lk{[1, 2, 3]}, the value \lk{ConsD T (ConsD T T)}
-represents the \textit{demand} that reveals the first 2 \lk{Cons}
-constructors of the list, but does not evaluate any other parts of the
-list. To improve readability of demand values, we will also use the
-syntax \lk{\_ : \_ : \_} where appropriate.
+further evaluation by using the \lk{T} constructor. For example, given
+the list \lk{[1, 2, 3]}, the value \lk{ConsD T (ConsD T T)} represents
+the \textit{demand} that reveals the first 2 \lk{Cons} constructors of
+the list, but does not evaluate any other parts of the list. To
+improve readability of demand values, we will also use the
+syntax \lk{\_ : \_ : \_} where appropriate, the underscore $\lk{\_}$
+represents a thunk.
 
 In Haskell, to trigger the evaluation of a lazy function, we need to
-exert some \textit{demand} on the result of that function. This
-suggests a natural flow for the specification of a lazy function ---
-given the demand on the output, the specification should calculate the
-demands on the input. However, this is not enough in general because
-the demand on the input could depend on the value of the inputs as
-well. The \lk{take} function introduced in \Cref{sec:intro} is such a
-case since the length of the output from \lk{take} depends on its
+exert some demand on the result of that function. This suggests a
+natural flow for the specification of a lazy function --- given the
+demand on the output, the specification should calculate the demands
+on the input. However, this is not enough in general because the
+demand on the input could depend on the \textit{value} of the inputs
+as well. The \lk{take} function introduced in \Cref{sec:intro} is such
+a case since the length of the output from \lk{take} depends on its
 integer argument.
@@
 to {\em specify} the difference between the two versions! At best, one
-could hope for a hand-crafted suite of unit tests containing bottoms
-to observe a difference in termination.
+could hope for a hand-crafted suite of unit tests containing divergent
+terms to observe a difference in termination.
 \leo{One thing we could do is ignore the knapsack and expand on the above.}
@@
 approach as the canonical approach for dynamic programming in functional
-programming?}. In particular, consider the famous knapsack problem.
+programming?}. Let us consider the famous knapsack problem.
 %
@@
 %
-However, lazy evaluation automatically finds this path if we use lazy
-maps to represent the solution table according to the recurrence
-relation \hzh{We can probably make this code prettier}
+However, in Haskell, we can define a function that computes one step
+of the recurrence relation, and take the fixpoint of this function as
+the final solution table. Laziness allows us to define such a solution
+table without actually computing all of its entries. When an entry is
+demanded, laziness again helps us by only evaluating the required
+entries, freeing us from having to design an optimal path along the
+solution table.
+%
 \begin{inlinecode}
-ws, vs :: [Int]
-n,  w  :: Int
+weights :: [Int]
+values  :: [Int]
+limit   :: Int
 
-solutionPre :: Map (Int, Int) Int -> Map (Int, Int) Int
-solutionPre soln = fromList [ ((j, k),
-                               if (j-1) < 0 || k - ws !! j < 0
-                               then if ws !! j <= k then vs !! j else 0
+solutionStep :: Map (Int, Int) Int -> Map (Int, Int) Int
+solutionStep soln =
+  fromList [((j, k), knapsack j k) |
+            j <- [0 .. length weights-1], k <- [0 .. limit]]
+  where
+    knapsack j k = if j - 1 < 0 || k - weights !! j < 0
+                   then if weights !! j <= k then values !! j else 0
                    else max (soln ! (j-1, k))
-                                        (soln ! (j-1, k - ws !! j) + vs !! j)
-                              )
-                            | j <- [0..2], k <- [0..50] ]
+                            (soln ! (j-1, k - weights !! j) + values !! j)
 
 solution :: Map (Int, Int) Int
-solution = fix solutionPre
+solution = fix solutionStep
+
 \end{inlinecode}
-
+%
 To query the optimal achievable values from the solution table, we can
@@
 
-However, using the lazy variant of \lk{Map} here is critical for the
+Using the lazy variant of \lk{Map} here is critical for the
 correctness of \lk{solution}. If we used a map strict in its values
@@
 expression. The \lk{if} expression contains value-level recursion to
-the solution map itself, and this would manifest as an infinite loop
+\lk{solution} itself, and this would manifest as an infinite loop
 at runtime. This is an incredibly subtle bug related to laziness.
@@
 of \lk{solution} depends on what is evaluated when the value at key \lk{(j, k)}
-of the map is demanded. Let $\lk{solutionPre}_n$ be the $n$-times iterated
-pre-fixpoint \lk{solutionPre . solutionPre . ... . solutionPre}, then if we
-demand the value at \lk{(j, k)} from the result of $\lk{solutionPre}_n$, its
+of the map is demanded. Let $\lk{solutionStep}_n$ be the $n$-times iterated
+step function \lk{solutionStep . solutionStep . ... . solutionStep}, then if we
+demand the value at \lk{(j, k)} from the result of $\lk{solutionStep}_n$, its
 input map should not be evaluated at the \lk{(j, k)} entry since otherwise
@@
 We can write another executable specification for this property
-of $\lk{solutionPre}_n$ (\Cref{fig:spec-map}). Testing the implementation based on
+of $\lk{solutionStep}_n$. Testing the implementation based on
 strict maps also immediately produces a counter example.
-
-\begin{figure}[H]
-\centering
-\hzh{TODO: specification for map}
+%
 \begin{inlinecode}
+TODO: expectation-spec for solutionStep
 \end{inlinecode}
-\caption{Specification for \lk{solutionPre}}
-\label{fig:spec-map}
-\end{figure}
+%
 

### 2018-03-03 paper/knapsack.hs

@@ -1,25 +1,25 @@
-import Data.Map.Lazy
+import Data.Map.Lazy hiding (foldr)
 import Data.Function
 
 main = return ()
 
-ws :: [Int]
-ws = [10, 20, 30]
+weights :: [Int]
+weights = [10, 20, 30]
 
-vs :: [Int]
-vs = [60, 100, 120]
+values :: [Int]
+values = [60, 100, 120]
 
-w :: Int
-w = 50
+limit :: Int
+limit = 50
 
-solutionPre :: Map (Int, Int) Int -> Map (Int, Int) Int
-solutionPre soln = fromList [ ((j, k),
-                               if (j-1) < 0 || k - ws !! j < 0
-                               then if ws !! j <= k then vs !! j else 0
-                               else max (soln ! (j-1, k))
-                                        (soln ! (j-1, k - ws !! j) + vs !! j)
-                              )
-                            | j <- [0..2], k <- [0..50] ]
+solutionStep :: Map (Int, Int) Int -> Map (Int, Int) Int
+solutionStep soln =
+  fromList [((j, k), knapsack j k) | j <- [0 .. length weights-1], k <- [0 .. limit]]
+  where
+    knapsack j k = if j - 1 < 0 || k - weights !! j < 0
+                   then if weights !! j <= k then values !! j else 0
+                   else max (soln ! (j-1, k))
+                            (soln ! (j-1, k - weights !! j) + values !! j)
 
 solution :: Map (Int, Int) Int
-solution = fix solutionPre
+solution = fix solutionStep

### 2018-03-03 paper/main.tex
[paper] updates

@@
 
-We can write another executable specification for this property
-of $\lk{solutionStep}_n$. Testing the implementation based on
-strict maps also immediately produces a counter example.
+We can write another executable specification for this property of
+$\lk{solutionStep}_n$. To describe the concrete \textit{demand} on
+maps, we need to first give a data type definition for \lk{Map}. We
+choose a simple binary search tree representation here. We also define
+a \lk{lookupD} function \hzh{we should probably move this to the
+appendix} to lookup demand at a given key.
 %
 \begin{inlinecode}
-TODO: expectation-spec for solutionStep
+data Map  k v = Bin (Map k v) k v (Map k v)
+              | Empty
+
+data MapD k v = BinD (Thunk (MapD k v)) (Thunk k) (Thunk v) (Thunk (MapD k v))
+              | EmptyD
+
+lookupD :: k -> MapD k v -> Thunk v
+lookupD k EmptyD                          = T
+lookupD k (BinD _ T      _ _)             = T
+lookupD k (BinD l (E k') v r) | k == k'   = v
+                              | k <= k'   = recur l
+                              | otherwise = recur r
+  where recur (E demand) = lookupD k demand
+        recur T          = T
+\end{inlinecode}
+%
+Now, we can write an executable specification
+for \lk{solutionStep}. \hzh{TODO: write the spec with
+strictCheckWithResults, and show a counterexample}
+%
+\begin{inlinecode}
 \end{inlinecode}
@@
 
-\paragraph*{Binary Search Trees}
+\section{Demand Types and Specification}
+\label{sec:demand-types-and-spec}
+\hzh{In this section, we should explain how demand types and specification work in general.
+We don't actually have to define a ListD/TreeD for every type out there.
+There is also a generic pattern for the type of a specification given the type of a function.
+We should also talk about the non-exact versions of specifications, and how they work}
 
-\leo{data structures - code - short explanation}
-
-\paragraph*{Characterizing Laziness}
-
-\begin{itemize}
-\item Specification of laziness - TreeD data structure \leo{Mention how it will be generically derived}
-\item Strict / Spine-strict maps
-\item Fully lazy map as a between spec
-\end{itemize}
+\section{Generating lazy functions}
+\label{sec:gen-lazy-fun}
+\hzh{In this section, we should talk about how lazy functions are generated, and why QuickCheck's implementation doesn't generate lazy functions.}
 

### 2018-03-05 paper/main.tex
[paper] some work on section 3

@@
 
+There is a general pattern between a type \lk{a} and
+its \textit{demand type}, for which we will write \lk{Demand a} from
+now on. In both the example on lists and the example on maps,
+a \lk{Thunk} type constructor is placed around all of the constructor
+arguments to the original type. The \lk{Thunk} type has the following
+definition:
+\begin{inlinecode}
+data Thunk a = E !a | T
+\end{inlinecode}
+Its two constructors \lk{E} and \lk{T} represents an evaluated
+fragment of the original data, and unexplored regions of the original
+data, respectively.
+
+This design is motivated by the runtime representation of Haskell
+values. Boxed values in Haskell are indirectly represented by a
+pointer, and the pointer can lead to either a constructor with
+pointers to deeper parts of the value, or a piece of unevaluated code,
+which will produce a fragment of the value when evaluated. This piece
+of code is also know as a thunk.
+
+The \lk{Thunk} type reifies \hzh{there is a slight caveat here though,
+we don't really capture sharing with this representation} the runtime
+representation into a Haskell value. Programmers can then inspect
+these demand values within Haskell. Most importantly, the demand types
+form the vocabulary of specifications on laziness.
+
+A \textit{specification} in StrictCheck is a function that computes
+demand values. When we talk about Haskell function's laziness
+informally, we often refer to these functions as ``strict'' or
+``lazy'' in its arguments. There is a lot of information we can unpack
+from such phrases --- how strict the function is in its arguments, the
+dependency of the strictness on the argument values, and the
+relationship between demand on inputs with demand on the output.
+
+StrictCheck specifications allows programmers to convey these
+information accurately. A specification takes a demand on the output,
+all of the inputs to the specified function, and computes a demand on
+all the inputs. For example, the specification of the \lk{take}
+function will have type \lk{Demand [a] -> Int -> [a] -> (Demand Int,
+Demand [a])}.
+
+\hzh{I want to say something about why we chose this design, but I'm not
+actually sure why...}
+
+The translation between a data type to its demand type and between a
+function type to its specification type are mechanic, and although we
+have presented them in a way that involves creating new demand type
+definitions, our implementation handles all of this automatically
+through generic programming. We delay this discussion
+to \Cref{sec:generic-programming}.
+
 \section{Generating lazy functions}
@@
 \hzh{In this section, we should talk about how lazy functions are generated, and why QuickCheck's implementation doesn't generate lazy functions.}
+
+\section{Generic Programming}
+\label{sec:generic-programming}
 

### 2018-03-05 paper/main.tex
[paper] Intro/Sec 2 nits

@@
 
-Lazy evaluation gives tremendous  to functional programmers
+Lazy evaluation gives tremendous power to functional programmers
 ~\cite{WhyFPMatters} and is actively used in various practical
@@
 leverages it to initialize expensive resources as needed\cn, and many
-functional reactive programming web frameworks crucially rely on
-it\cn \leo{This point might need more exposition about frp}. Haskell
+functional reactive programming web frameworks rely on
+it crucially\cn \leo{This point might need more exposition about frp}. Haskell
 is the most popular programming language that allows programmers to
-enjoy the power of lazy evaluation by default. However, that power
+enjoy the power of lazy evaluation by default. However, it
 does not come without a price.
@@
 The power of laziness puts great responsibility on programmers'
-shoulders. An incorrect implementation can either behave functionally
-the same but with inferior \caleb{Inferior than what?} performance, or it can completely break
-down!\caleb{unclear. loops?} Reasoning about lazy programs is also notoriously hard --- the
-programmer needs to have a mental model of the order of evaluation of
-the code, and this order is not always obvious from the syntax of the
-code.
+shoulders. An implementation that uses laziness incorrectly can either
+behave functionally the same but with inferior \caleb{Inferior than
+what?} performance---compared to a strict counterpart, or it can
+completely break down!\caleb{unclear. loops?} Reasoning about lazy
+programs is also notoriously hard --- the programmer needs to have a
+mental model of the order of evaluation of the code, and this order is
+not always obvious from the syntax of the code.
 
-How, then, can we test and catch these laziness bugs during
+How, then, can we test and catch such laziness bugs during
 development? Is it possible to precisely specify laziness behavior
@@
 Haskell programs. We implement these ideas as a Haskell library. For
-concreteness, consider the following small example on the
-function \lk{take}, which accepts an integer $n$ and a list, and returns
+concreteness, consider the following example using the standard
+\lk{take} function on lists, which accepts an integer $n$ and a list, and returns
 a prefix of the list with length $n$. If $n$ is larger than the length
@@
 %technical detail of the spec}
-\item We propose a framework for automated testing against such specifications.\caleb{Automated testing? Can be more specific}
+\item We propose a framework for automated testing against such specifications.\caleb{Automated testing? Can be more specific}\leo{Should we mention QuickCheck here?}
 \item We improve upon the random function generators from QuickCheck\cn to generate functions with arbitrary laziness behaviors.
-\item We discuss our generic programming-based impementation to apply StrictCheck to all algebraic data types.\caleb{I don't know
-the meaning of non-trivial and almost-all}
+\item We discuss our generic programming-based impementation to apply StrictCheck to all algebraic data types.\leo{I guess ``packed'' maps do not qualified as algebraic?}
 \end{itemize}
@@
 \label{sec:quickchick}
+
+\subsection{Purely functional queues}
+\label{sec:queues}
 
@@
 Turning our intuition above into a Haskell function, we complete the
-specification of \lk{rotate} in \Cref{fig:rotate-spec}
+specification of \lk{rotate} in \Cref{fig:rotate-spec}.
+%
 \begin{figure}[H]
@@
 \end{figure}
-
-\lk{rotate\_spec} exactly captures the core idea from Okasaki's \lk{rotate}
+%
+The \lk{rotate\_spec} exactly captures the core idea from Okasaki's \lk{rotate}
 operation, and testing the reverse and append implementation against this
@@
 counter example looks like?}\leo{Definitely!}
+
+\subsection{Key- vs value-strict maps}
 
@@
 \leo{One thing we could do is ignore the knapsack and expand on the above.}
+\leo{Here we need a similar discussion as with lists}
 

### 2018-03-06 paper/main.tex
[paper] more on section 3

@@
   Spec $ \predict d n xs ->
-    predict (nf n) (if n > length xs then E d else endThunk d)
-  where
-    endThunk :: PosDemand [a] -> Demand [a]
-    endThunk = \case
-      Nil'        -> T
-      dx :% T     -> E (dx :% T)
-      dx :% E dxs -> E (dx :% endThunk dxs)
+    predict n (if n > length xs then d else d ++ thunk)
 \end{inlinecode}
@@
 \section{Strictness Specifications, by Example}
-\label{sec:quickchick}
+\label{sec:examples}
 
@@
 data, respectively.
-
+%
 This design is motivated by the runtime representation of Haskell
 values. Boxed values in Haskell are indirectly represented by a
-pointer, and the pointer can lead to either a constructor with
+pointer, and the pointer can lead to either a constructor bundled with
 pointers to deeper parts of the value, or a piece of unevaluated code,
-which will produce a fragment of the value when evaluated. This piece
+which will produce a fragment of the value upon execution. This piece
 of code is also know as a thunk.
@@
 these demand values within Haskell. Most importantly, the demand types
-form the vocabulary of specifications on laziness.
+form a critical part of specifications on laziness.
 
@@
 
-StrictCheck specifications allows programmers to convey these
-information accurately. A specification takes a demand on the output,
-all of the inputs to the specified function, and computes a demand on
-all the inputs. For example, the specification of the \lk{take}
-function will have type \lk{Demand [a] -> Int -> [a] -> (Demand Int,
-Demand [a])}.
+StrictCheck specifications allows programmers to convey this
+information accurately. In StrictCheck, specifications are represented
+by the \lk{Spec} type constructor. The \lk{Spec} type constructor
+takes two type arguments: the first is a type-level list holding the
+types of arguments to specified function, and the second is the type
+of the output of specified function. For example, consider the
+function \lk{take :: Int -> [a] -> [a]}, one possible type of its
+specification is \lk{Spec '[Int, [a]] [a]}.
 
-\hzh{I want to say something about why we chose this design, but I'm not
-actually sure why...}
+In general, specification types are not unique for the same
+function. Again consider \lk{take}, we can certainly choose the
+specification type \lk{Spec '[Int] ([a] -> [a])} as well. However, the
+meaning of this specification is very different from the other
+choice. A specification of this type predicts how the \lk{Int}
+argument is used given a demand on the partially-applied \lk{take}
+function, and this specification does not produce information on how
+the lists are used at all.
+
+StrictCheck specifications apply to higher-order functions, and are
+compositional. Take the example of \lk{map :: (a -> b) -> [a] -> [b]},
+the type of its specification is \lk{Spec '[a] b -> Spec '[[a]] [b]},
+a function from specifications to specifications. This allows
+programmers to apply the specification of \lk{map} to any
+specification of the function argument to \lk{map}, and conveniently
+synthesize a specification based on that of \lk{map}.
+
+The \lk{Spec} type also has a single constructor with the same
+name. This constructor takes a higher-order function in
+continuation-passing-style ~\cn, where the first argument to the
+function is a framework-supplied function that takes the predicted
+demand on the inputs. We will always name this argument \lk{predict}
+in this paper. The function is CPS'd to hide some internal plumbing
+that presents a variadic interface to the user, so that the number of
+arguments to \lk{predict} always matches the length of the type-level
+list in \lk{Spec}.
 
@@
 have presented them in a way that involves creating new demand type
-definitions, our implementation handles all of this automatically
-through generic programming. We delay this discussion
+definitions in \Cref{sec:examples}, our implementation handles this
+transition automatically through generic programming. We delay
+discussion of this implementation detail
 to \Cref{sec:generic-programming}.

### 2018-03-07 paper/main.tex
[paper] finished some TODOs in section 1 and 2

@@
 %
-\hzh{TODO: describe what this spec means in English}
+The \lk{Spec} constructor takes a function, whose first
+argument \lk{predict} is the predicted demand on the \lk{Int} argument
+of \lk{take}, and second argument the predicted demand on the \lk{[a]}
+argument. This spec simply says that the \lk{Int} argument will always
+be used, and the demand on the list argument is either the same as the
+demand on the output, or demand on the output plus some
+unevaluated \lk{thunk} at the end. The former case happens when $n$ is
+larger than the list's length, otherwise the latter case applies.
 
@@
 improve readability of demand values, we will also use the
-syntax \lk{\_ : \_ : \_} where appropriate, the underscore $\lk{\_}$
-represents a thunk.
+syntax \lk{\_ : \_ : \_} instead of explicit demand type values. In
+this expression, the first two underscores represent unevaluated
+elements in the list, while the last underscore represents unevaluated
+tail of the list.
 
@@
 as well. The \lk{take} function introduced in \Cref{sec:intro} is such
-a case since the length of the output from \lk{take} depends on its
-integer argument.
+a case.
 
-Putting these pieces together, we arrive at an appropriate type for
-the specification of \lk{rotate}: \lk{ListD a -> [a] -> [a] -> [a] ->
-(ListD a, ListD a, ListD a)}.
+Putting these pieces together, we can specify the laziness behavior
+of \lk{rotate} with Okasaki's insight: as long as the front list is
+longer than the back list, demanding the first element from the result
+of \lk{rotate} only causes evaluation of the first element of the
+front list ~\cite{Okasaki95simpleand}. Since \lk{rotate} is always
+used by supplying the empty list \lk{[]} as the accumulator argument,
+we define \lk{rot fs bs = rotate fs bs []}, and show a specification
+of \lk{rot}.
 %
-\hzh{The description here needs improvement}
-To specify the laziness behavior of \lk{rotate}, we return to Okasaki's insight:
-as long as the front list is longer than the back list, demanding the first
-element from the result of \lk{rotate} only causes evaluation of the first
-element of the front list ~\cite{Okasaki95simpleand}. Using this precondition
-for \lk{rotate}, we arrive at the following executable
-specification \lk{rotate\_spec}.
-
-Turning our intuition above into a Haskell function, we complete the
-specification of \lk{rotate} in \Cref{fig:rotate-spec}.
-%
-\begin{figure}[H]
-\centering
-\hzh{TODO}
 \begin{inlinecode}
-TODO: rewrite this spec so it looks like the invariant in Okasaki's paper.
+rot_spec :: Shaped a => Spec '[[a], [a]] [a]
+rot_spec =
+  Spec $ \predict d fs bs ->
+    let (fs', bs') = splitAt (length fs) (cap d)
+        spineLen  = length (cap (d ++ [undefined]))  -- # of spine thunks forced
+        overflow  = spineLen       > length fs  -- begun taking from bs?
+        overrot   = length (cap d) > length bs  -- forced all of bs?
+        padLength =
+          length bs `min`
+            if overflow
+            then length bs - length bs'
+            else length (cap d)
+        spinePad = replicate padLength thunk
+    in predict
+         (                    fs' ++ if overflow            then [] else thunk)
+         (spinePad ++ reverse bs' ++ if overflow || overrot then [] else thunk)
 \end{inlinecode}
-\caption{Specification for \lk{rotate}}
-\label{fig:rotate-spec}
-\end{figure}
 %
@@
 operation, and testing the reverse and append implementation against this
-specification immediately produces a counter example. \hzh{should we show what a
-counter example looks like?}\leo{Definitely!}
+specification immediately produces a counter example.
+\begin{figure}[H]
+\centering
+\includegraphics[width=0.8\textwidth]{revappend_counterexample}
+\caption{Counterexample for \lk{revAppend} under \lk{rot\_spec}}
+\end{figure}
 

### 2018-03-08 paper/main.tex
[paper] more on section 2 and 3

@@
 choose a simple binary search tree representation here. We also define
-a \lk{lookupD} function \hzh{we should probably move this to the
-appendix} to lookup demand at a given key.
+a function \lk{iterSolution} that performs $n$ iterations
+of \lk{solutionStep}, and indexes into the map at a given key. The
+specification is written for \lk{iterSolution}.
 %
@@
 
-data MapD k v = BinD (Thunk (MapD k v)) (Thunk k) (Thunk v) (Thunk (MapD k v))
-              | EmptyD
+iterSolution :: Key -> Int -> Map (Int, Int) Int -> Maybe Int
+iterSolution (Key k) n soln = lookup m k
+  where m | n <= 0    = soln
+          | otherwise = (iterate solutionStep soln) !! n
 
-lookupD :: k -> MapD k v -> Thunk v
-lookupD k EmptyD                          = T
-lookupD k (BinD _ T      _ _)             = T
-lookupD k (BinD l (E k') v r) | k == k'   = v
-                              | k <= k'   = recur l
-                              | otherwise = recur r
-  where recur (E demand) = lookupD k demand
-        recur T          = T
+iterSolution_spec :: Evaluation '[Key, Int, KMap] (Maybe Int)
+                  -> Maybe (Int, Int)
+iterSolution_spec (Evaluation args demands dOut) =
+  let I (Key evalK) = hd args
+      I nIter       = hd (tl args)
+      dInM          = hd (tl (tl demands))
+      inM           = replaceThunk Empty (fromDemand @KMap dInM)
+  in  if (inM == Empty)   ||
+         isBaseCase evalK ||
+         nIter <= 0       ||
+         (isNothing $ lookup inM evalK)
+      then Nothing
+      else Just evalK
+  where isBaseCase (j, k) = j - 1 < 0 || k - weights !! j < 0
 \end{inlinecode}
 %
-Now, we can write an executable specification
-for \lk{solutionStep}. \hzh{TODO: write the spec with
-strictCheckWithResults, and show a counterexample}
-%
-\begin{inlinecode}
-\end{inlinecode}
-%
+This specification is different from \lk{rot\_spec} --- it uses
+the \lk{Evaluation} type instead of \lk{Spec}. In StrictCheck, we
+reverse the \lk{Spec} type for specifications that computes the exact
+demand on inputs. However, there many cases where we might want to
+only check whether some property holds of the observed demands on the
+inputs. For example, here we only need to check that the value
+at \lk{(j, k)} is not evaluated, and we don't compute the exact demand
+on the input map. For these specifications, we write a function that
+takes an \lk{Evaluation} value, which contains the inputs, the
+observed demand on the inputs, and a randomly generated demand on the
+output by the framework.
 
-\leo{Punned names for sections? Or is that just an urns thing? :P }
-
-\begin{itemize}
-\item In this section we will showcase our proposed specification language \leo{It's not a language tho}
-\item while introducing the running example of our paper: maps.
-\item Maps are particularly well-suited for a discussion of laziness because of the interesting different
-possible behaviors: strict-maps, value-lazy maps \leo{Point out Data.Map.Strict
-vs Data.Map}, and fully lazy maps
-\item For simplicity, we will use binary search trees as maps, assuming the keys are integers and forego balancing
-as the added complexity is irrelevant to the laziness discussion. \leo{Can we claim it wouldn't matter at all? I think so}
-\end{itemize}
+The specification then checks whether the expected property holds for
+this evaluation, and if the evaluation fails the expectation, it
+outputs an evidence of the failure. The evidence is then returned to
+the programmer by the framework for further investigation. In this
+case, the evidence is simply the \lk{(j, k)} pair at which the
+solution table was accessed.
 
@@
 \label{sec:demand-types-and-spec}
-\hzh{In this section, we should explain how demand types and specification work in general.
-We don't actually have to define a ListD/TreeD for every type out there.
-There is also a generic pattern for the type of a specification given the type of a function.
-We should also talk about the non-exact versions of specifications, and how they work}
-
 There is a general pattern between a type \lk{a} and
@@
 pointers to deeper parts of the value, or a piece of unevaluated code,
-which will produce a fragment of the value upon execution. This piece
-of code is also know as a thunk.
+which will produce a fragment of the value upon execution
+~\cite{SPJSTG}. This piece of code is also know as a thunk.
 
@@
 
-A \textit{specification} in StrictCheck is a function that computes
-demand values. When we talk about Haskell function's laziness
+An \textit{exact specification} in StrictCheck is a function that
+computes demand values. When we talk about Haskell function's laziness
 informally, we often refer to these functions as ``strict'' or
@@
 function, and this specification does not produce information on how
-the lists are used at all.
+the lists are used at all. StrictCheck specifications apply to
+higher-order functions as well. Consider \lk{map :: (a -> b) -> [a] ->
+[b]}, the type of its specification is \lk{Spec '[a -> b, [a]]
+[b]}. In order to test higher-order functions, StrictCheck needs to
+generate functions with interesting laziness behavior. This is
+discussed in more details in \Cref{sec:gen-lazy-fun}.
 
-StrictCheck specifications apply to higher-order functions, and are
-compositional. Take the example of \lk{map :: (a -> b) -> [a] -> [b]},
-the type of its specification is \lk{Spec '[a] b -> Spec '[[a]] [b]},
-a function from specifications to specifications. This allows
-programmers to apply the specification of \lk{map} to any
-specification of the function argument to \lk{map}, and conveniently
-synthesize a specification based on that of \lk{map}.
-
-The \lk{Spec} type also has a single constructor with the same
-name. This constructor takes a higher-order function in
+The \lk{Spec} type has a single constructor with the same name. This
+constructor takes a higher-order function in
 continuation-passing-style ~\cn, where the first argument to the
@@
 to \Cref{sec:generic-programming}.
+
+The \lk{Evaluation} type constructor has the same type-level index
+as \lk{Spec}. Its first argument is the type-level list of all
+argument types of the specified function, and the second argument is
+the type of the specified function's result. Specifications
+using \lk{Evaluation} must output values of type \lk{Maybe e},
+where \lk{e} is the type of the evidence. StrictCheck is polymorphic
+in this evidence type, so programmers can choose to output any useful
+information for debugging. These specifications are spiritually
+similar to properties in QuickCheck
+~\cite{Claessen:2000:QLT:357766.351266}.
 

### 2018-03-08 paper/main.tex
[paper] wrote a subsection on connection between Demand a based specs and fromDemand-based specs

@@
 \label{sec:demand-types-and-spec}
+
+\subsection{Types}
 There is a general pattern between a type \lk{a} and
@@
 
+\subsection{Bridging ad-hoc laziness testing with StrictCheck}
+In Haskell development, programmers often inserts divergent terms such
+as \lk{error} or \lk{undefined} into the inputs of a function to test
+its laziness. If a function evaluates such a divergent term, execution
+gets aborted. We will call these divergent terms ``bottoms'' following
+the common terminology in literature.
+
+This is a very convenient approach to laziness testing because
+programmers can use functions on these partial data containing bottoms
+as if they are total values. As long as such functions don't require
+the bottom values at runtime, execution continues. However, this
+process typically involves repeated manual process of --- inserting
+bottoms, executing the function, and diagnosing.
+
+StrictCheck bridges the gap between this approach of laziness testing
+and property-based laziness testing by supplying the two
+functions \lk{fromDemand :: Demand a -> a} and \lk{toDemand :: a ->
+Demand a}. Intuitively, \lk{fromDemand} produces a partial data
+containing bottoms given a value of its demand type. This is performed
+by re-assembing a value of type \lk{a} using the shaped described by
+its demand value, and using a special bottom value at all locations
+where a \lk{Thunk} appears in its demand value.
+
+We implemented this process with a special exception, which is a kind
+of bottom value, that cannot be thrown nor caught by users of
+StrictCheck since its type is hidden. This allows \lk{toDemand} to
+determine the precise locations of such bottoms, and convert them back
+to the \lk{Thunk} value of type \lk{Demand a}.
+
+In fact, in all of the 3 examples presented, we have written
+specifications in this style instead of explicit manipulations of
+type \lk{Demand a}. This gives us the full arsenal of existing API on
+data types involved in the specification, and when runtime safety is
+required, we simply convert these partial values back to demand
+values.
+
+Since StrictCheck can generate random demands on values, programmers
+can eliminate the manual process inserting bottoms, and focus on the
+diagnosis phase of the usual debug loop.
+
 \section{Generating lazy functions}

### 2018-03-09 paper/main.tex
[paper] filled out some citations

@@
 applications. For instance, the big-data framework Spark uses laziness
-to drive computations on demand\cn, Apple's programming language Swift
-leverages it to initialize expensive resources as needed\cn, and many
-functional reactive programming web frameworks rely on
-it crucially\cn \leo{This point might need more exposition about frp}. Haskell
-is the most popular programming language that allows programmers to
-enjoy the power of lazy evaluation by default. However, it
-does not come without a price.
+to drive computations on demand
+~\cite{Zaharia:2012:RDD:2228298.2228301}, Apple's programming language
+Swift leverages it to initialize expensive resources as needed
+~\cite{Swift}, and many functional reactive programming web frameworks
+rely on it crucially\cn \leo{This point might need more exposition
+about frp}. Haskell is the most popular programming language that
+allows programmers to enjoy the power of lazy evaluation by
+default. However, it does not come without a price.
 \leo{Didn't we have an FRP quote by someone that Laziness was the third source of
@@
 \item We propose a framework for automated testing against such specifications.\caleb{Automated testing? Can be more specific}\leo{Should we mention QuickCheck here?}
-\item We improve upon the random function generators from QuickCheck\cn to generate functions with arbitrary laziness behaviors.
+\item We improve upon the random function generators from QuickCheck ~\cite{Claessen:2000:QLT:357766.351266} to generate functions with arbitrary laziness behaviors.
 \item We discuss our generic programming-based impementation to apply StrictCheck to all algebraic data types.\leo{I guess ``packed'' maps do not qualified as algebraic?}
@@
 constructor takes a higher-order function in
-continuation-passing-style ~\cn, where the first argument to the
-function is a framework-supplied function that takes the predicted
-demand on the inputs. We will always name this argument \lk{predict}
-in this paper. The function is CPS'd to hide some internal plumbing
-that presents a variadic interface to the user, so that the number of
-arguments to \lk{predict} always matches the length of the type-level
-list in \lk{Spec}.
+continuation-passing-style ~\cite{Reynolds:1993:DC:198112.198114},
+where the first argument to the function is a framework-supplied
+function that takes the predicted demand on the inputs. We will always
+name this argument \lk{predict} in this paper. The function is CPS'd
+to hide some internal plumbing that presents a variadic interface to
+the user, so that the number of arguments to \lk{predict} always
+matches the length of the type-level list in \lk{Spec}.
 

### 2018-03-09 paper/main.tex
[paper] minor edits

@@
 The \lk{Spec} constructor takes a function, whose first
-argument \lk{predict} is the predicted demand on the \lk{Int} argument
-of \lk{take}, and second argument the predicted demand on the \lk{[a]}
-argument. This spec simply says that the \lk{Int} argument will always
-be used, and the demand on the list argument is either the same as the
-demand on the output, or demand on the output plus some
-unevaluated \lk{thunk} at the end. The former case happens when $n$ is
-larger than the list's length, otherwise the latter case applies.
+argument \lk{predict} is is a continuation that takes the predicted
+demands on the inputs to \lk{take}, and second argument of the
+function is the observed demand on the output \lk{[a]}. This spec says
+that the \lk{Int} argument will always be used, and the demand on the
+list argument is either the same as the demand on the output, or
+demand on the output plus some unevaluated \lk{thunk} at the end. The
+former case happens when $n$ is larger than the list's length,
+otherwise the latter case applies.
 
@@
 hand, \lk{rotate} produces a list immediately, and packs the rest of
-the computation into the lazy list \lk{rest}.
+the computation into the tail of the lazy list.
 
@@
 term \textit{demand} to refer to the shape of such usage. Concretely,
-a \textit{demand} is a shape that describes the used part of a data
-structure. For lists, \textit{demands} will be prefixes of lists. We
-use the type \lk{ListD} to represent such demands.
+a \textit{demand} is a shape that describes the evaluated part of a
+data structure. For lists, \textit{demands} will be prefixes of
+lists. We use the type \lk{ListD} to represent such demands.
 %
@@
 longer than the back list, demanding the first element from the result
-of \lk{rotate} only causes evaluation of the first element of the
-front list ~\cite{Okasaki95simpleand}. Since \lk{rotate} is always
-used by supplying the empty list \lk{[]} as the accumulator argument,
-we define \lk{rot fs bs = rotate fs bs []}, and show a specification
+of \lk{rotate} only causes $O(1)$ amount of evaluation
+~\cite{Okasaki95simpleand}. Since \lk{rotate} is always used by
+supplying the empty list \lk{[]} as the accumulator argument, we
+define \lk{rot fs bs = rotate fs bs []}, and show a specification
 of \lk{rot}.
@@
 %
-The \lk{rotate\_spec} exactly captures the core idea from Okasaki's \lk{rotate}
+The \lk{rot\_spec} exactly captures the core idea from Okasaki's \lk{rotate}
 operation, and testing the reverse and append implementation against this
@@
 
-Moreover, it is important to be able to distinguish whether such a datatype
-follows a desired laziness behavior. For instance, the standard functional
-solution for dynamic programming problems crucially uses laziness \hzh{is this
-solution really standard? Is there some work we can cite to establish this
-approach as the canonical approach for dynamic programming in functional
+Moreover, it is important to be able to distinguish whether such a
+datatype follows a desired laziness behavior. \hzh{the previous
+sentence and the next one feels disconnected} For instance, the
+standard functional solution for dynamic programming problems
+crucially uses laziness \hzh{is this solution really standard? Is
+there some work we can cite to establish this approach as the
+canonical approach for dynamic programming in functional
 programming?}. Let us consider the famous knapsack problem.
@@
 
-To catch this behavior through testing, we can write another specification that
-captures this correctness condition. The well-foundedness \hzh{is this a word?}
-of \lk{solution} depends on what is evaluated when the value at key \lk{(j, k)}
-of the map is demanded. Let $\lk{solutionStep}_n$ be the $n$-times iterated
-step function \lk{solutionStep . solutionStep . ... . solutionStep}, then if we
-demand the value at \lk{(j, k)} from the result of $\lk{solutionStep}_n$, its
-input map should not be evaluated at the \lk{(j, k)} entry since otherwise
-evaluating the value at \lk{(j, k)} becomes an infinite loop at the fixpoint.
+To catch this behavior through testing, we can write another
+specification that captures this correctness condition. Termination
+of \lk{solution} depends on what is evaluated when the value at
+key \lk{(j, k)} of the map is demanded. Let $\lk{solutionStep}_n$ be
+the $n$-times iterated step function \lk{solutionStep . solutionStep
+. ... . solutionStep}, then if we demand the value at \lk{(j, k)} from
+the result of $\lk{solutionStep}_n$, its input map should not be
+evaluated at the \lk{(j, k)} entry since otherwise evaluating the
+value at \lk{(j, k)} becomes an infinite loop at the fixpoint.
 
@@
 the \lk{Evaluation} type instead of \lk{Spec}. In StrictCheck, we
-reverse the \lk{Spec} type for specifications that computes the exact
+reserve the \lk{Spec} type for specifications that computes the exact
 demand on inputs. However, there many cases where we might want to
@@
 on the input map. For these specifications, we write a function that
-takes an \lk{Evaluation} value, which contains the inputs, the
-observed demand on the inputs, and a randomly generated demand on the
-output by the framework.
+takes an \lk{Evaluation} value, which contains all participants
+necessary for analyzing laziness properties of \lk{iterSolution}.
 

### 2018-03-09 paper/main.tex
Refactoring a bit in the rot example

--- 2018-03-08 paper/main.tex
@@
 
-To describe the laziness behavior, we need to first introduce a few
-types that describe how lists are used. We use the
-term \textit{demand} to refer to the shape of such usage. Concretely,
-a \textit{demand} is a shape that describes the used part of a data
-structure. For lists, \textit{demands} will be prefixes of lists. We
-use the type \lk{ListD} to represent such demands.
+To describe the laziness behavior, we need to first introduce a few of
+StrictCheck's types. First of all, we use a type \lk{Thunk} to capture
+whether a particular field is forced or not.
+%
+\begin{inlinecode}
+  data Thunk = E !a | T
+\end{inlinecode}
+%
+Similarly to a \lk{Maybe}, the \lk{Thunk} datatype has two
+constructors: a nullary one \lk{T} corresponding to thunks and a
+strict single-argument one \lk{E} corresponding to evaluated values.
+
+Secondly, StrictCheck introduces the notion of a \textit{demand}: a
+shape that describes the used part of a data structure. For
+lists, \textit{demands} will be prefixes of lists. We use the
+type \lk{ListD} to represent such list demands.
 %
@@
 %
-The \lk{Thunk} type used in \lk{ListD} has 2 constructors: \lk{E}
-for \textit{evaluated} and \lk{T}, a nullary constructor,
-for \textit{thunk}. In \lk{ListD}, \lk{Thunk} wraps around all the
-list constructor fields, and this allows us to represent lack of
-further evaluation by using the \lk{T} constructor. For example, given
-the list \lk{[1, 2, 3]}, the value \lk{ConsD T (ConsD T T)} represents
-the \textit{demand} that reveals the first 2 \lk{Cons} constructors of
-the list, but does not evaluate any other parts of the list. To
-improve readability of demand values, we will also use the
-syntax \lk{\_ : \_ : \_} instead of explicit demand type values. In
-this expression, the first two underscores represent unevaluated
-elements in the list, while the last underscore represents unevaluated
-tail of the list.
+In \lk{ListD}, \lk{Thunk} wraps around all the list constructor
+fields, and this allows us to represent lack of further evaluation by
+using the \lk{T} constructor.
+
+\leo{The following is slightly too dense. How can we take out some
+part of this as code?}
+For example, given the list \lk{[1, 2, 3]}, the value \lk{ConsD T
+(ConsD T T)} represents the \textit{demand} that reveals the first
+2 \lk{Cons} constructors of the list, but does not evaluate any other
+parts of the list. To improve readability of demand values, we will
+also use the syntax \lk{\_ : \_ : \_} instead of explicit demand type
+values. In this expression, the first two underscores represent
+unevaluated elements in the list, while the last underscore represents
+unevaluated tail of the list.
+%
+We also use metavariables \lk{\X{i}} to denote either a thunk or an
+evaluated field. For instance, \lk{\X1 : \X2 : \_} can denote the
+above 2-\lk{Cons} demand \lk{\_ : \_ : \_}, a similar demand where
+both cons cells are evaluated \lk{ 1 : 2 : \_}, or a mix of both.
 
@@
 Putting these pieces together, we can specify the laziness behavior
-of \lk{rotate} with Okasaki's insight: as long as the front list is
-longer than the back list, demanding the first element from the result
-of \lk{rotate} only causes evaluation of the first element of the
-front list ~\cite{Okasaki95simpleand}. Since \lk{rotate} is always
-used by supplying the empty list \lk{[]} as the accumulator argument,
-we define \lk{rot fs bs = rotate fs bs []}, and show a specification
-of \lk{rot}.
+of \lk{rotate}. Since \lk{rotate} is always used by supplying the
+empty list \lk{[]} as the accumulator argument, we define \lk{rot fs
+bs = rotate fs bs []}, and show a specification of \lk{rot}.  Before
+giving the actual specification however, let's try to build up some
+intuition a series of example demands for the simpler, reverse and
+append version:
+%
+\begin{inlinecode}
+  rot_simple fs bs = fs ++ reverse bs
+\end{inlinecode}
+
+All of our example demands will be of the following form.
+%
+\rotex{[1,2,3]}{[6,5,4]}{[1,2,3,4,5,6]}
+      {\X1 : \X2 : \_}{\X1 : \X2 : \_}{\_}
+%
+The fist column shows the input lists \lk{fs} and \lk{bs}; here they
+are \lk{[1,2,3]} and \lk{[4,5,6]} respectively. The second column
+shows the result of \lk{rot\_simple fs bs} when fully evaluated;
+here \lk{[1,2,3,4,5,6]}. The third column shows the actual output
+demand exerted; just like we described above, the demand \lk{\X1 : \X2
+: \_} means that we only force the first two cons cells and, depending
+on whether the \lk{\X{i}} are thunks or not, we might also force the
+first two elements. The first and third column together are the actual
+inputs to our specification. The second column is there for the
+readers convenience.
+
+The last column is the demand on the inputs predicted by our
+specification.  In this first example, where the output demand is
+smaller than the input list, the demand on the first argument \lk{fs}
+is equal to the demand, while the demand on the second list \lk{bs} is
+the empty demand. Indeed, if concretely the output demand is \lk{ \_ :
+2 : \_} (which could for example be obtained by printing \lk{hd \$
+tl \$ rot\_simple [1,2,3] [6,5,4]}), then only the first
+two \lk{Cons}-cells of \lk{fs} will be forced, and of those only the
+second one will have its integer field evaluated.
+
+On the other hand, if the output demand is larger than the first list,
+then \textit{all} of the spine of second argument has to be evaluated.
+%
+\rotex{[1,2,3]}{[6,5,4]}{[1,2,3,4,5,6]}
+      {\X1 : \X2 : \X3 : \X4 : \_}{ [ \X1 , \X2 , \X3 ] } { [ \_ , \_ , \X4 ] }
+%
+The predicted demand on the first list will be the top part of the
+demand of length equal to \lk{length fs}. The predicted demand on the
+second list will be the reverse of the remaining demand, padded with
+thunks if necessary.
+\leo{One more tricky case, when the demand on the output forces
+the null constructor of the bs list. How do we talk about that?}
+
+Now we can give the actual StrictCheck specification
+of \lk{rot\_simple}:
+
+\leo{Debug the stupid spec and add it and describe it}
+
+We can now move on to the more complex \lk{rot} function from Okasaki:
+as long as the front list is longer than the back list, demanding the
+first element from the result of \lk{rotate} only causes evaluation of
+the first element of the front list ~\cite{Okasaki95simpleand}.
 %

### 2018-03-10 paper/main.tex
More writing on the rot example

@@
 thunks if necessary.
-\leo{One more tricky case, when the demand on the output forces
-the null constructor of the bs list. How do we talk about that?}
 
-Now we can give the actual StrictCheck specification
-of \lk{rot\_simple}:
+The two cases above give rise to the following StrictCheck specification:
+%
+\begin{inlinecode}
+rot_simple_spec :: (Show a, Shaped a) => Spec '[[a], [a]] [a]
+rot_simple_spec =
+  Spec $ \predict d fs bs ->
+    let demandOnFs
+          | length (cap d) > length fs =
+              take (length fs) d
+          | otherwise = d
+        demandOnBs
+          | length (cap d) > length fs =
+              reverse $ take (length bs)
+                      $ (drop (length fs) (cap d)) ++ repeat thunk
+          | otherwise = thunk
+    in predict demandOnFs demandOnBs
+\end{inlinecode}
+%
+Just like the \lk{take\_spec} of the introduction, this specification
+is the constructor \lk{Spec} applied to a function that takes a
+function \lk{predict}, the output demand \lk{d}, as well as the actual
+input lists \lk{fs} and \lk{bs} and \lk{predict}s the demand
+on \lk{fs} and \lk{bs}. In particular, if we don't demand more from
+the output of \lk{rot\_simple} than the length of \lk{fs} then the
+demand on \lk{fs} is just \lk{d} itself, while \lk{bs} is left
+completely unevaluated. On the other hand, if we do, only the first
+half of the demand is actually relevant to \lk{fs}, while the rest
+constitutes the demand on \lk{bs} after being appropriately padded
+with thunks if necessary and reversed.
 
-\leo{Debug the stupid spec and add it and describe it}
+Let's test this specification!
+%
+\leo{Screengrab bug? How did you do the other one?}
+%
+StrictCheck quickly reveals a bug: when the output demand is
+spinestrict and the back list \lk{bs} is empty, then the second list
+will also be forced to ensure it is of the correct shape
+(\lk{[]}). \leo{TODO: corrected spec. Thinking side-by-side figure
+with the next one?}
 
